
# Visual Congruence

@fig-prop shows a pie chart of the proportions of total youth 
offenders between 2011 and 2021 in 
different ethnic groups and a bar plot of the same data.
We learned in @sec-mapping-visual-features that mapping 
**quantitative** proportions to **lengths**, like in the
bar plot, is more appropriate
and more accurate than mapping the proportions to **angles**
or **areas**, like in the pie chart.

However, which data visualisation is superior at conveying the
fact that Māori offenders make up approximately 50% of the total?
Which visualisation is better at conveying 
the proportion of offenders that are European/Other?

Although it is easy to perceive that the bar for Māori is just below 0.5
in length from the bar plot, it is even easier to see that the light blue
wedge in the pie chart is just under half a circle.
It is also easy to see that the red wedge is close to a third of a circle.

```{r}
#| echo: false
#| label: fig-prop
#| fig.width: 4
#| fig-cap: Data visualisations of the proportion of offenders from different ethnic groups.  
#| fig-subcap:
#|   - A pie chart.
#|   - A bar plot.
#| layout-ncol: 2
ggplot(crimeGroupTotal) + 
    geom_col(aes(x=total/sum(total), y="", fill=group), colour="black") +
    coord_polar() +
    labs(title="Proportions of Youth Offenders 2011 to 2021") +
    theme(aspect.ratio=1,
          axis.title=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.y=element_blank())

gg <- ggplot(crimeGroupTotal) + 
    geom_col(aes(x=total/sum(total), y=group), colour="black") +
    scale_x_continuous(expand=expansion(c(0, .05))) +
    theme(aspect.ratio=1,
          axis.title=element_blank(),
          axis.ticks.y=element_blank())
grid.newpage()
pushViewport(viewport(height=.8, width=.8))
print(gg, newpage=FALSE)
popViewport()
```

The purpose of this section is to show that the effectiveness
of a data visualisation does not just 
depend on the accuracy with which we can
map back from basic visual features 
to raw data values.

In terms of our simple model of visual processing (@sec-visual-processing),
this section deals with higher levels of processing where connections
can be made between existing knowledge and the visual features that
arise from the perception of basic visual features
(see @fig-visual-model).

## Part-to-whole

The reason why the pie chart in @fig-prop is more effective
than a bar plot for conveying proportions is because 
the wedges in a pie chart are **congruent** with fractions of a whole.
A semi-circle has a strong visual association with *half* of a circle
and thirds and quarters are similarly evocative.

> One reason why a pie chart can be effective is because 
  segments of a pie vividly convey fractions of a whole.

@fig-bar-part shows that if we place a bar representing a proportion bar 
within a reference rectangle that corresponds to a proportion of 1,
then we can also create a part-to-whole congruence with bars, though the effect
is not as strong as it is for pie wedges.

```{r}
#| echo: false
#| label: fig-bar-part
#| fig-cap: A bar plot of the proportion of offenders from different ethnic groups with reference rectangles to enhance the part-to-whole relationship.  
gg <- ggplot(crimeGroupTotal) + 
    geom_col(aes(x=1, y=group), colour="black", fill="grey80") +
    geom_col(aes(x=total/sum(total), y=group), colour="black") +
    scale_x_continuous(expand=expansion(0)) +
    theme(aspect.ratio=1,
          axis.title=element_blank(),
          axis.ticks.y=element_blank())
pushViewport(viewport(height=.8, width=.8))
print(gg, newpage=FALSE)
popViewport()
```

## Dark is more

## More is more

Representing negative values.

## Same is same

## Visual dissonance

Pie chart of crime level - larger bars have lighter colour so the messages
are a bit mixed.  To do with interaction of visual features ?
Can you make a plot that does not create the visual dissonance
(e.g., darker colours with different positions?)

## TODO

Also related to familiarity of visual objects ...

More is more for length and area esp.
=> bars MUST start at zero

Dark is more for colour.

The above need citations

Angle is part-of-whole.

Lines with peaks and valleys have a natural correspondence with 
mountain ranges.

At least some of the ideas in "congruence" are similar to "expressiveness"
cite tamarara munzner and jock mackinlay 
(and Enrico Bertini? https://filwd.substack.com/p/beyond-precision-expressiveness-in)
although tamara's expressiveness is more like the quant/qual distinction
(or possibly capacity?).

Looks like Barbara Tversky uses congruence more in the sense we are using
it here.

Need to name check "cognitive fit theory" ?
