
# Visual Congruence

@fig-prop shows a pie chart of the proportions of total youth 
offenders between 2011 and 2021 in 
different ethnic groups and a bar plot of the same data.
We learned in @sec-mapping-visual-features that mapping 
**quantitative** proportions to **lengths**, like in the
bar plot, is more appropriate
and more accurate than mapping the proportions to **angles**
or **areas**, like in the pie chart.

However, which data visualisation is superior at conveying the
fact that Māori offenders make up approximately 50% of the total?
Which visualisation is better at conveying 
the proportion of offenders that are European/Other?

Although it is easy to perceive that the bar for Māori is just below 0.5
in length from the bar plot, it is even easier to see that the light blue
wedge in the pie chart is just under half a circle.
It is also easy to see that the red wedge is close to a third of a circle.

```{r}
#| echo: false
#| label: fig-prop
#| fig.width: 4
#| fig-cap: Data visualisations of the proportion of offenders from different ethnic groups.  
#| fig-subcap:
#|   - A pie chart.
#|   - A bar plot.
#| layout-ncol: 2
ggplot(crimeGroupTotal) + 
    geom_col(aes(x=total/sum(total), y="", fill=group), colour="black") +
    coord_polar() +
    labs(title="Proportions of Youth Offenders 2011 to 2021") +
    theme(aspect.ratio=1,
          axis.title=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.y=element_blank())

gg <- ggplot(crimeGroupTotal) + 
    geom_col(aes(x=total/sum(total), y=group), colour="black") +
    scale_x_continuous(expand=expansion(c(0, .05))) +
    theme(aspect.ratio=1,
          axis.title=element_blank(),
          axis.ticks.y=element_blank())
grid.newpage()
pushViewport(viewport(height=.8, width=.8))
print(gg, newpage=FALSE)
popViewport()
```

The purpose of this section is to show that we need to consider more
than just the accuracy or capacity of our mappings from 
data values to basic visual feature in order to understand the effectiveness
of a data visualisation.

## Part-to-whole

The reason why the pie chart in @fig-prop is more effective
than a bar plot for conveying proportions is because 
the wedges in a pie chart are **congruent** with fractions of a whole.
A semi-circle has a strong visual association with *half* of a circle
and thirds and quarters are similarly evocative.

> One reason why a pie chart can be effective is because 
  segments of a pie vividly convey fractions of a whole.

@fig-bar-part shows that if we place a bar representing a proportion bar 
within a reference rectangle that corresponds to a proportion of 1,
then we can also create a part-to-whole congruence with bars, though the effect
is not as strong as it is for pie wedges.

```{r}
#| echo: false
#| label: fig-bar-part
#| fig-cap: A bar plot of the proportion of offenders from different ethnic groups with reference rectangles to enhance the part-to-whole relationship.  
gg <- ggplot(crimeGroupTotal) + 
    geom_col(aes(x=1, y=group), colour="black", fill="grey80") +
    geom_col(aes(x=total/sum(total), y=group), colour="black") +
    scale_x_continuous(expand=expansion(0)) +
    theme(aspect.ratio=1,
          axis.title=element_blank(),
          axis.ticks.y=element_blank())
pushViewport(viewport(height=.8, width=.8))
print(gg, newpage=FALSE)
popViewport()
```

## More is more

One reason why a bar plot like @fig-prop-2 conveys relative proportions
well is because a longer bar is congruent with *more* of something.
This is part of the reason why visual features like **length** and **area** are 
appropriate for conveying quantitative values (@sec-quant-features).

Representing negative values.

horizontal bars (distances) versus vertical bars (amounts) 
look up Tversky and kosslyn

## Dark is more

## Same is same

In @sec-differences we saw that the visual system is very
sensitive to **differences**, 
particularly when everything else remains the **same**.
This means that data symbols that change if the data values change
will be effective because the visual system will notice the difference
in the data symbols.

For example, 
in a bar plot like @fig-prop-2, the lengths of the bars are different,
so we map *back* to changes in the data values.
This is helped by the fact that the bars are all the same "width";
they do not change in the vertical dimension.

* Things that do not change are not salient (e.g., the widths of bars in
  bar plot), BUT things that DO change are salient (so only do this to
  represent changes in the data).  This should be cited to something.
  Principle of unambiguity ?

  Example is mosaic plot which changes *shape* as well as area so
  two identical areas could end up looking different => BAD!
  This has to be mentioned in the mosaic plot case study
  because that comes LATER.

  Another example is word cloud which changes orientation NOT based on data.

  AND have "different is different"?
  And just cite the relevant literature for these points in footnotes?
  


## Visual dissonance

Pie chart of crime level - larger bars have lighter colour so the messages
are a bit mixed.  To do with interaction of visual features ?
Can you make a plot that does not create the visual dissonance
(e.g., darker colours with different positions?)

## TODO

Also related to familiarity of visual objects ...

More is more for length and area esp.
=> bars MUST start at zero

Lines with peaks and valleys have a natural correspondence with 
mountain ranges.

Lines imply continuity (at least) if not linear interpolation
(density plot this is absolutely ok)

At least some of the ideas in "congruence" are similar to "expressiveness"
cite tamarara munzner and jock mackinlay 
(and Enrico Bertini? https://filwd.substack.com/p/beyond-precision-expressiveness-in)
although tamara's expressiveness is more like the quant/qual distinction
(or possibly capacity?).

Looks like Barbara Tversky uses congruence more in the sense we are using
it here.

Need to name check "cognitive fit theory" ?

