
# Visual Congruence

@fig-prop shows a pie chart of the proportions of total youth 
offenders between 2011 and 2021 in 
different ethnic groups and a bar plot of the same data.
We learned in @sec-mapping-visual-features that mapping 
**quantitative** proportions to **lengths**, like in the
bar plot, is more appropriate
and more accurate than mapping the proportions to **angles**
or **areas**, like in the pie chart.

However, which data visualisation is superior at conveying the
fact that Māori offenders make up approximately 50% of the total?
Which visualisation is better at conveying 
the proportion of offenders that are European/Other?

Although it is easy to perceive that the bar for Māori is just below 0.5
in length from the bar plot, it is even easier to see that the light blue
wedge in the pie chart is just under half a circle.
It is also easy to see that the red wedge is close to a third of a circle.

```{r}
#| echo: false
#| label: fig-prop
#| fig.width: 4
#| fig-cap: Data visualisations of the proportion of offenders from different ethnic groups.  
#| fig-subcap:
#|   - A pie chart.
#|   - A bar plot.
#| layout-ncol: 2
ggplot(crimeGroupTotal) + 
    geom_col(aes(x=total/sum(total), y="", fill=group), colour="black") +
    coord_polar() +
    labs(title="Proportions of Youth Offenders 2011 to 2021") +
    theme(aspect.ratio=1,
          axis.title=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.y=element_blank())

gg <- ggplot(crimeGroupTotal) + 
    geom_col(aes(x=total/sum(total), y=group), colour="black") +
    scale_x_continuous(expand=expansion(c(0, .05))) +
    theme(aspect.ratio=1,
          axis.title=element_blank(),
          axis.ticks.y=element_blank())
grid.newpage()
pushViewport(viewport(height=.8, width=.8))
print(gg, newpage=FALSE)
popViewport()
```

The purpose of this section is to show that we need to consider more
than just the accuracy or capacity of our mappings from 
data values to basic visual feature in order to understand the effectiveness
of a data visualisation.

## Part-to-whole

The reason why the pie chart in @fig-prop is more effective
than a bar plot for conveying proportions is because 
the wedges in a pie chart are **congruent** with fractions of a whole.
A semi-circle has a strong visual association with *half* of a circle
and thirds and quarters are similarly evocative.

> One reason why a pie chart can be effective is because 
  segments of a pie vividly convey fractions of a whole.

@fig-bar-part shows that if we place a bar representing a proportion bar 
within a reference rectangle that corresponds to a proportion of 1,
then we can also create a part-to-whole congruence with bars, though the effect
is not as strong as it is for pie wedges.

```{r}
#| echo: false
#| label: fig-bar-part
#| fig-cap: A bar plot of the proportion of offenders from different ethnic groups with reference rectangles to enhance the part-to-whole relationship.  
gg <- ggplot(crimeGroupTotal) + 
    geom_col(aes(x=1, y=group), colour="black", fill="grey80") +
    geom_col(aes(x=total/sum(total), y=group), colour="black") +
    scale_x_continuous(expand=expansion(0)) +
    theme(aspect.ratio=1,
          axis.title=element_blank(),
          axis.ticks.y=element_blank())
pushViewport(viewport(height=.8, width=.8))
print(gg, newpage=FALSE)
popViewport()
```

## More is more

We saw in @sec-quant-features that 
visual features like **length** and **area** are 
appropriate for conveying quantitative values. 
For example, in @fig-more-is-more, the lengths of the bars
are appropriate for conveying the total number of offenders in 
different ethnic groups.

Part of the reason why **length** and **area** work so well is because
a longer bar naturally conveys the idea of *more*.
By contrast, the **position** of data points in a dot plot of the same data,
as shown in @fig-more-is-more, do not convey the same sense of *more*.
In a bar plot, the data symbols that correspond to larger data values
are actually larger data symbols;
the presentation of larger data values is more abstract in a dot plot.

```{r}
#| echo: false
#| label: fig-more-is-more
#| fig.width: 4
#| fig-cap: A bar plot and a dot plot of the total number of offenders for different ethnic groups.
#| fig-subcap:
#|   - Bars.
#|   - Dots.
#| layout-ncol: 2

gg <- ggplot(crimeGroupTotal) + 
    geom_col(aes(x=total, y=group)) +
    scale_x_continuous(expand=expansion(c(0, .05))) +
    theme(aspect.ratio=1)
grid.newpage()
pushViewport(viewport(width=.9, height=.9))
print(gg, newpage=FALSE)
popViewport()

gg <- ggplot(crimeGroupTotal) + 
    geom_segment(aes(x=-Inf, xend=Inf, y=group, yend=group), 
                 linetype="dotted") +
    geom_point(aes(x=total, y=group)) +
    scale_x_continuous(name="total") +
    theme(aspect.ratio=1)
grid.newpage()
pushViewport(viewport(width=.9, height=.9))
print(gg, newpage=FALSE)
popViewport()
```

Another way that **length** can convey information more naturally than
**position** is visualising negative values.
For example @fig-neg-bar shows the total points differential
(points scored minus points conceded) for
Tier One nations at Rugby World Cups.
The *direction* of the bars very clearly indicates the change from
positive points differential to negative points differential.

```{r}
#| echo: false
#| label: fig-neg-bar
#| fig-cap: A bar plot of the total points differential for Tier One nations at Rugby World Cups.
rwcAlltime <- aggregate(rwcAll[c("scored", "conceded")], 
                        list(team=rwcAll$team), sum)
rwcAlltime$diff <- rwcAlltime$scored - rwcAlltime$conceded
rwcAlltime$team <- reorder(rwcAlltime$team, rwcAlltime$diff)
ggplot(rwcAlltime) +
    geom_col(aes(x=diff, y=team)) +
    scale_x_continuous(name="points differential") +
    theme(aspect.ratio=2/3)
```

By comparison, the change from positive to negative points differential
is much more abstract in @fig-neg-dot, which shows a dot plot of the
same data.

```{r}
#| echo: false
#| label: fig-neg-dot
#| fig-cap: A dot plot of the total points differential for Tier One nations at Rugby World Cups.
ggplot(rwcAlltime) +
    geom_point(aes(x=diff, y=team)) +
    geom_segment(aes(x=-Inf, xend=Inf, y=team, yend=team), linetype="dotted") +
    annotate("segment", x=0, xend=0, y=-Inf, yend=Inf, linetype="solid") +
    scale_x_continuous(name="points differential") +
    theme(aspect.ratio=2/3)
```

A more subtle effect can be seen in @fig-bar-orientation,
which shows two different bar plots of the total number
of offenders in different ethnic groups.
The vertical bars are more natural for conveying count data like this
because of the natural correspondence with stacks of items.[^vertical-bars]

[^vertical-bars]:
  horizontal bars (distances) versus vertical bars (amounts) 
  look up Tversky and kosslyn

```{r}
#| echo: false
#| label: fig-bar-orientation
#| fig.width: 4
#| fig-cap: Data visualisations of the number of offenders from different ethnic groups.  
#| fig-subcap:
#|   - Horizontal bars.
#|   - Vertical bars.
#| layout-ncol: 2
ggplot(crimeGroupTotal) + 
    geom_col(aes(x=total, y=group)) +
    scale_x_continuous(expand=expansion(c(0, .05))) +
    scale_y_discrete(name=NULL) +
    force_panelsizes(rows = unit(2.5, "in"),
                     cols = unit(2.5, "in"))

temp <- crimeGroupTotal
temp$group <- gsub("/", "/\n", temp$group)
ggplot(temp) + 
    geom_col(aes(y=total, x=group)) +
    scale_x_discrete(name=NULL) +
    scale_y_continuous(expand=expansion(c(0, .05))) +
    force_panelsizes(rows = unit(2.5, "in"),
                     cols = unit(2.5, "in"))
```

## Case study: Divergent colour palettes

@fig-neg-tile shows a heatmap of the points differential for Tier One
nations at each Rugby World Cup.  The data symbols are rectangles, or "tiles".
The teams are mapped to the vertical **positions** of the tiles and
the years are mapped to the horizontal **positions** of the tiles.
These are both effective mappings and we can easily map *back* to
teams or years.
The points differentials are mapped to the **colour** of the tiles,
but in a slightly complicated way.  Positive data values map to a shade
of purple, while negative values map to a shade of brown.
In both cases, we have a constant **hue**, with increasing **chroma** and
decreasing **luminance** as the data values get further away from zero.

The use of two **hues** creates two different groups (positive and negative
values), while the changes in **chroma** and **luminance** convey
changes in absolute magnitude (if only for ordinal comparisons).

@fig-neg-tile also demonstrates another example of visual congruence because
darker colours are naturally associated with larger data values.[^dark-is-more]

[^dark-is-more]: 
  citation please!

```{r}
#| echo: false
#| label: fig-neg-tile
#| fig-cap: A heatmap of the total points differentials for Tier One nations at individual Rugby World Cups.  South Africa was excluded from the 1987 and 1991 tournament due to its apartheid policies.[^gggrid-bite]
rwcAllyear <- aggregate(rwcAll[c("scored", "conceded")], 
                        list(team=rwcAll$team, year=rwcAll$year), sum)
rwcAllyear$diff <- rwcAllyear$scored - rwcAllyear$conceded
rwcAllyear$team <- factor(rwcAllyear$team, levels=levels(rwcAlltime$team))
rwcAllyear <- merge(rwcAllyear,
                    melt(with(rwcAllyear, table(team, year)) == 0, 
                         value.name="absent"),
                    all=TRUE)
edge <- function(data, coords) {
    polygonGrob(c(0, 0,
                  coords$x[data$absent][2] + .05, 
                  coords$x[data$absent][2] + .05, 
                  0, 0, 1, 1),
                c(1, 
                  coords$y[data$absent][1] + .05, 
                  coords$y[data$absent][1] + .05, 
                  coords$y[data$absent][1] - .05, 
                  coords$y[data$absent][1] - .05,
                  0, 0, 1),
                gp=gpar(fill=NA))
}
ggplot(rwcAllyear) +
    geom_tile(aes(x=year, y=team, fill=diff)) +
    scale_fill_continuous_diverging(palette="Purple-Brown", rev=TRUE,
                                    na.value="transparent") +
    scale_x_continuous(breaks=unique(rwcAllyear$year), expand=expansion(0)) +
    scale_y_discrete(expand=expansion(0)) +
    grid_panel(edge, aes(x=year, y=team, absent=absent)) +
    theme(panel.border=element_blank(),
          aspect.ratio=1)
```

[^gggrid-bite]:
  The {gggrid} package [@pkg-gggrid]
  was used to take the "bite" out of the panel
  border in @fig-neg-tile where South Africa was excluded.

> A diverging colour palette is effective because it creates two distinct
  ordinal visual features.

## Same is same {#sec-same-same}

In @sec-differences we saw that the visual system is very
sensitive to **differences**, 
particularly when everything else remains the **same**.
This means that data symbols that change if the data values change
will be effective because the visual system will notice the difference
in the data symbols.
More specifically, if a **visual feature** of a data symbol
changes then the visual system will notice that change.
Furthermore, if a visual feature of a data symbol
changes while all other visual features of the data symbol remain the **same**,
the visual system will notice the change more easily.

For example, 
in a bar plot like @fig-bar, the data symbols are bars
and only the **lengths** of the bars and the vertical **positions** of
the bars are different.
The "widths" of the bars and the **colours** of the bars remain
the same.

> Another reason why bar plots are effective is because the bars
  differ in length, but are otherwise the same.

## Visual dissonance {#sec-dissonance}

The fact that **length** and **area** naturally convey amounts means
that care must be taken *not* to confuse that association.
For example, @fig-bar-not-zero shows the number of times each team
entered the opposition's final third during a match at the 
2023 Women's World Cup.
The final third of the pitch is shown as darker green regions, broken
into five different zones, with a number of entries for each zone
and for each team represented by a line with an arrow.

In @fig-bar-not-zero, longer lines correspond to a larger number of 
entries.  However, the bars start at the halfway line on the pitch
rather than at the final third of the pitch.
This means that the zero entries by South Africa into the second-from-right
zone is represented by a non-zero-length line.
Apart from making comparisons of the line lengths very inaccurate,
this creates a confusing visual effect because the viewer
is expected to associate a non-zero length with a value of zero.

```{r}
#| echo: false
#| label: fig-bar-not-zero
#| fig-cap: The number of times each team entered the opposition's final third in the Women's World Cup match between the Netherlands and South Africa.  This data visualisation is an adaptation of a graphic shown during television coverage of the match.
entries <- data.frame(team=rep(c("Netherlands", "South Africa"), each=5),
                      region=rep(1:5, 2),
                      count=c(10, 4, 1, 2, 3, 
                              4, 2, 2, 0, 2))
hue <- 120
grid.newpage()
pushViewport(viewport(width=.5, height=.9,
                      gp=gpar(lwd=1, fill=NA)),
             viewport(y=0, height=.75, just="bottom"))
grid.rect(gp=gpar(fill=hcl(hue, 70, 80)))
pushViewport(viewport(clip="on"))
grid.circle(y=0, r=10/80)
popViewport()
grid.rect(y=1, height=2/3, just="top", 
          gp=gpar(col=NA, fill=hcl(hue, 70, 60)))
grid.rect(y=1, height=2/3, just="top", 
          width=44/80, gp=gpar(col=NA, fill=hcl(hue, 70, 50)))
grid.rect(y=1, height=2/3, just="top", 
          width=20/80, gp=gpar(col=NA, fill=hcl(hue, 70, 40)))
## grid.rect(y=1, height=2/3, just="top")
grid.rect(y=1, height=6/50, width=20/80, just="top")
grid.rect(y=1, height=18/50, width=44/80, just="top")
pushViewport(viewport(clip=rectGrob(y=1 - 18/50, just="top")))
grid.circle(y=1 - 12/50, r=.2)
popViewport()
grid.circle(y=1 - 12/50, r=unit(1, "mm"), gp=gpar(fill="black"))
x <- unit(rep(c(9/80, 24/80, .5, 56/80, 71/80), 2), "npc") +
     unit(rep(c(-3, 3), each=5), "mm")
y <- 1/3 + (entries$count)/max(entries$count)*2/3
col <- rep(c("pink", "orange"), each=5)
grid.segments(x, 0, x, y, 
              arrow=arrow(type="closed"),
              gp=gpar(col=col, fill=col, lwd=1, 
                      linejoin="mitre", lineend="butt"))
grid.segments(x, 0, x, y - .05, 
              gp=gpar(col=col, fill=col, lwd=15, lineend="butt"))
grid.text(entries$count, x, unit(10, "mm"), gp=gpar(fontface="bold"))
grid.rect()
pushViewport(viewport(y=unit(1, "npc") + unit(10, "pt"), height=unit(20, "pt"),
                      just="bottom"))
grid.text("Final Third Entries",
          gp=gpar(fontsize=20))
popViewport()
pushViewport(viewport(y=unit(1, "npc") + unit(40, "pt"), height=unit(20, "pt"),
                      just="bottom"))
grid.rect(gp=gpar(fill=hcl(hue, 70, 40)))
grid.text(c("Netherlands", "South Africa"), 
          x=unit(0:1, "npc") + unit(c(2, -2), "mm"),
          hjust=c(0, 1),
          gp=gpar(col=c("pink", "orange"), fontface="bold"))
popViewport()
```

For any example of **congruence**, 
where a visual feature naturally corresponds to
a property of the data,
there is an opportunity for **dissonance** if the visual feature
is used inappropriately.  We can improve the effectiveness of 
a data visualisation by taking advantage of visual congruence
and we can reduce the effectiveness of a data visualisation 
if we create visual dissonance.

@fig-stroop shows that we can also create dissonance with our mappings
of data values to colours.

```{r}
#| echo: false
#| label: fig-stroop
#| fig-cap:  The Stroop effect. Try to name the *colour* of each word.
#| fig-height: 1
grid.newpage()
pushViewport(viewport(width=.5))
grid.text(c("red", "green", "blue"), 1:3/4, 
          gp=gpar(col=c("green", "blue", "red"), fontsize=20, fontface="bold"))
```

We saw in @sec-same-same that we want to map differences in data values
to differences in visual features *and* keep everything else the
same.  Failing to follow that approach will make it harder to 
perceive changes in the data values.
However, a worse error, and 
another source of visual dissonance arises if we change
visual features *without* any underlying change in the data values.

**TODO** refer back to using quant feature for qual data as a form
of this problem - then foreshadow (or even link to) examples to come
later on.

Pie chart of crime level - larger bars have lighter colour so the messages
are a bit mixed.  To do with interaction of visual features ?
Can you make a plot that does not create the visual dissonance
(e.g., darker colours with different positions?)

> A data visualisation is *not* effective if it contains mappings
  that are dissonant because that will cause slower and/or
  incorrect mappings *back* to the data.

## TODO

Lines contain angles which represent differences in x/y (kosslyn 1989)

Lines with peaks and valleys have a natural correspondence with 
mountain ranges.

Lines imply continuity (at least) if not linear interpolation
(density plot this is absolutely ok)

Lines (angles) imply trends (whereas bars are discrete) (Zacks and Tversky 1999)
People use "trend words" (increasing/decreasing) versus "discrete words"
(higher/lower than) AND are faster to judge absolute values from bars
but slope from line graphs.

At least some of the ideas in "congruence" are similar to "expressiveness"
cite tamarara munzner and jock mackinlay 
(and Enrico Bertini? https://filwd.substack.com/p/beyond-precision-expressiveness-in)
although tamara's expressiveness is more like the quant/qual distinction
(or possibly capacity?).

Looks like Barbara Tversky uses congruence more in the sense we are using
it here.

Need to name check "cognitive fit theory" ?

## Summary
