
# Visual Congruence {#sec-congruence}

@fig-prop shows a pie chart of the proportions of total youth 
offenders between 2011 and 2021 in 
different ethnic groups and a bar plot of the same data.
We learned in @sec-accuracy that mapping 
**quantitative** proportions to **lengths**, like in the
bar plot, is more accurate than mapping the proportions to **angles**
or **areas**, like in the pie chart.

However, the superior represenation of the
fact that Māori offenders make up approximately 50% of the total is
the pie chart rather than the bar plot.

Although it is easy to determine from the length of the bar and the scale 
on the x-axis in the bar plot that the bar for Māori is just below 0.5, 
it is even easier, without any guides at all, to see that the light blue
wedge in the pie chart is just under half a circle.
It is also easy to see that the red wedge is close to a third of a circle.

```{r}
#| echo: false
#| label: fig-prop
#| fig.width: 4
#| fig-cap: Data visualisations of the proportion of offenders from different ethnic groups.  
#| fig-subcap:
#|   - A pie chart.
#|   - A bar plot.
#| layout-ncol: 2
ggplot(crimeGroupTotal) + 
    geom_col(aes(x=total/sum(total), y="", fill=group), colour="black") +
    coord_polar() +
    labs(title="Proportions of Youth Offenders 2011 to 2021") +
    theme(aspect.ratio=1,
          axis.title=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.y=element_blank())

gg <- ggplot(crimeGroupTotal) + 
    geom_col(aes(x=total/sum(total), y=group), colour="black") +
    scale_x_continuous(expand=expansion(c(0, .05))) +
    theme(aspect.ratio=1,
          axis.title=element_blank(),
          axis.ticks.y=element_blank())
grid.newpage()
pushViewport(viewport(height=.8, width=.8))
print(gg, newpage=FALSE)
popViewport()
```

The purpose of this section is to show that we need to consider more
than just the accuracy or capacity of our mappings from 
data values to basic visual features in order to understand the effectiveness
of a data visualisation.[^congruence-lit]

[^congruence-lit]:
  @TVERSKY2002247 discusses congruence at least in the sense of 
  matching the spatial properties of data values to the spatial
  arrangement of visual features.  The concept is also discussed in
  @DBLP:journals/corr/abs-2008-11310.

## Part-to-whole

The reason why the pie chart in @fig-prop is more effective
than a bar plot for conveying proportions is because 
the wedges in a pie chart are **congruent** with fractions of a whole.
A semi-circle has a strong visual association with *half* of a circle
and thirds and quarters are similarly evocative.

> One reason why a pie chart can be effective is because 
  segments of a pie vividly convey fractions of a whole.

@fig-bar-part shows that if we place a bar representing a proportion bar 
within a reference rectangle that corresponds to a proportion of 1,
then we can also create a part-to-whole congruence with bars, though the effect
is not as strong as it is for pie wedges.

```{r}
#| echo: false
#| label: fig-bar-part
#| fig-cap: A bar plot of the proportion of offenders from different ethnic groups with reference rectangles to enhance the part-to-whole relationship.  
gg <- ggplot(crimeGroupTotal) + 
    geom_col(aes(x=1, y=group), colour="black", fill="grey80") +
    geom_col(aes(x=total/sum(total), y=group), colour="black") +
    scale_x_continuous(expand=expansion(0)) +
    theme(aspect.ratio=1,
          axis.title=element_blank(),
          axis.ticks.y=element_blank())
pushViewport(viewport(height=.8, width=.8))
print(gg, newpage=FALSE)
popViewport()
```

## More is more

We saw in @sec-quant-features that 
visual features like **length** and **area** are 
appropriate for conveying quantitative values. 
For example, in @fig-more-is-more, the lengths of the bars
are appropriate for conveying the total number of offenders in 
different ethnic groups.

Part of the reason why **length** and **area** work so well is because
a longer bar naturally conveys the idea of *more*.
By contrast, the **position** of data points in a dot plot of the same data,
as shown in @fig-more-is-more, do not inherently 
convey the same sense of *more*.
In a bar plot, the data symbols that correspond to larger data values
are actually larger data symbols;
the presentation of larger data values is more abstract in a dot plot.

```{r}
#| echo: false
#| label: fig-more-is-more
#| fig.width: 4
#| fig-cap: A bar plot and a dot plot of the total number of offenders for different ethnic groups.
#| fig-subcap:
#|   - Bars.
#|   - Dots.
#| layout-ncol: 2

gg <- ggplot(crimeGroupTotal) + 
    geom_col(aes(x=total, y=group)) +
    scale_x_continuous(expand=expansion(c(0, .05))) +
    theme(aspect.ratio=1)
grid.newpage()
pushViewport(viewport(width=.9, height=.9))
print(gg, newpage=FALSE)
popViewport()

gg <- ggplot(crimeGroupTotal) + 
    geom_segment(aes(x=-Inf, xend=Inf, y=group, yend=group), 
                 linetype="dotted") +
    geom_point(aes(x=total, y=group)) +
    scale_x_continuous(name="total") +
    theme(aspect.ratio=1)
grid.newpage()
pushViewport(viewport(width=.9, height=.9))
print(gg, newpage=FALSE)
popViewport()
```

Another way that **length** can convey information more naturally than
**position** is visualising negative values.
For example @fig-neg-bar shows the total points differential
(points scored minus points conceded) for
Tier One nations at Rugby World Cups.
The *direction* of the bars very clearly indicates the change from
positive points differential to negative points differential.

```{r}
#| echo: false
#| label: fig-neg-bar
#| fig-cap: A bar plot of the total points differential for Tier One nations at Rugby World Cups.
rwcAlltime$team <- reorder(rwcAlltime$team, rwcAlltime$diff)
ggplot(rwcAlltime) +
    geom_col(aes(x=diff, y=team)) +
    scale_x_continuous(name="points differential") +
    theme(aspect.ratio=2/3)
```

By comparison, the change from positive to negative points differential
is much more abstract in a dot plot of the
same data (@fig-neg-dot).

```{r}
#| echo: false
#| label: fig-neg-dot
#| fig-cap: A dot plot of the total points differential for Tier One nations at Rugby World Cups.
ggplot(rwcAlltime) +
    geom_point(aes(x=diff, y=team)) +
    geom_segment(aes(x=-Inf, xend=Inf, y=team, yend=team), linetype="dotted") +
    annotate("segment", x=0, xend=0, y=-Inf, yend=Inf, linetype="solid") +
    scale_x_continuous(name="points differential") +
    theme(aspect.ratio=2/3)
```

> One reason why bar plots are effective is because larger data values are
  represented by larger data symbols.

A more subtle effect can be seen in @fig-bar-orientation,
which shows two different bar plots of the total number
of offenders in different ethnic groups.
In this case,
the vertical bars are more natural for conveying count data like this
because of the natural correspondence with stacks of items.[^vertical-bars]

[^vertical-bars]:
  horizontal bars (distances) versus vertical bars (amounts) 
  look up Tversky and kosslyn
  OTOH, horizontal bars are more accurate than vertical bars (citation please!)

```{r}
#| echo: false
#| label: fig-bar-orientation
#| fig.width: 4
#| fig-cap: Data visualisations of the number of offenders from different ethnic groups.  
#| fig-subcap:
#|   - Horizontal bars.
#|   - Vertical bars.
#| layout-ncol: 2
ggplot(crimeGroupTotal) + 
    geom_col(aes(x=total, y=group)) +
    scale_x_continuous(expand=expansion(c(0, .05))) +
    scale_y_discrete(name=NULL) +
    force_panelsizes(rows = unit(2.5, "in"),
                     cols = unit(2.5, "in"))

temp <- crimeGroupTotal
temp$group <- gsub("/", "/\n", temp$group)
ggplot(temp) + 
    geom_col(aes(y=total, x=group)) +
    scale_x_discrete(name=NULL) +
    scale_y_continuous(expand=expansion(c(0, .05))) +
    force_panelsizes(rows = unit(2.5, "in"),
                     cols = unit(2.5, "in"))
```

## Same is same {#sec-same-same}

In @sec-differences we saw that the visual system is very
sensitive to **differences**, 
particularly when everything else remains the **same**.
This means that mapping *changes* in data values to *changes* in 
data symbols 
will be effective because the visual system will notice the differences
in the data symbols.
More specifically, if a **visual feature** of a data symbol
changes then the visual system will notice that change.
Furthermore, if a visual feature of a data symbol
changes while all other visual features of the data symbol remain the **same**,
the visual system will notice the change more easily.

For example, 
in a bar plot like @fig-bar, the data symbols are bars
and only the **lengths** of the bars and the vertical **positions** of
the bars are different.
The "widths" of the bars and the **colours** of the bars remain
the same.

> Another reason why bar plots are effective is because the bars
  differ in length, but are otherwise the same.

## Visual dissonance {#sec-dissonance}

The fact that **length** and **area** naturally convey amounts means
that care must be taken *not* to confuse that association.
For example, @fig-bar-not-zero shows the number of times each team
entered the opposition's final third during a match at the 
2023 Women's World Cup.
The final third of the pitch is shown as darker green regions, broken
into five different zones, with a number of entries for each zone
and for each team represented by a line with an arrow.

In @fig-bar-not-zero, longer lines correspond to a larger number of 
entries.  However, the bars start at the halfway line on the pitch
rather than at the final third of the pitch.
This means that the zero entries by South Africa into the second-from-right
zone is represented by a non-zero-length line.
Apart from making comparisons of the line lengths very inaccurate,
this creates a confusing visual effect because the viewer
is expected to associate a non-zero length with a value of zero.

```{r}
#| echo: false
#| label: fig-bar-not-zero
#| fig-cap: The number of times each team entered the opposition's final third in the Women's World Cup match between the Netherlands and South Africa.[^final-third]
entries <- data.frame(team=rep(c("Netherlands", "South Africa"), each=5),
                      region=rep(1:5, 2),
                      count=c(10, 4, 1, 2, 3, 
                              4, 2, 2, 0, 2))
hue <- 120
grid.newpage()
pushViewport(viewport(width=.5, height=.9,
                      gp=gpar(lwd=1, fill=NA)),
             viewport(y=0, height=.75, just="bottom"))
grid.rect(gp=gpar(fill=hcl(hue, 70, 80)))
pushViewport(viewport(clip="on"))
grid.circle(y=0, r=10/80)
popViewport()
grid.rect(y=1, height=2/3, just="top", 
          gp=gpar(col=NA, fill=hcl(hue, 70, 60)))
grid.rect(y=1, height=2/3, just="top", 
          width=44/80, gp=gpar(col=NA, fill=hcl(hue, 70, 50)))
grid.rect(y=1, height=2/3, just="top", 
          width=20/80, gp=gpar(col=NA, fill=hcl(hue, 70, 40)))
## grid.rect(y=1, height=2/3, just="top")
grid.rect(y=1, height=6/50, width=20/80, just="top")
grid.rect(y=1, height=18/50, width=44/80, just="top")
pushViewport(viewport(clip=rectGrob(y=1 - 18/50, just="top")))
grid.circle(y=1 - 12/50, r=.2)
popViewport()
grid.circle(y=1 - 12/50, r=unit(1, "mm"), gp=gpar(fill="black"))
x <- unit(rep(c(9/80, 24/80, .5, 56/80, 71/80), 2), "npc") +
     unit(rep(c(-3, 3), each=5), "mm")
y <- 1/3 + (entries$count)/max(entries$count)*2/3
col <- rep(c("pink", "orange"), each=5)
grid.segments(x, 0, x, y, 
              arrow=arrow(type="closed"),
              gp=gpar(col=col, fill=col, lwd=1, 
                      linejoin="mitre", lineend="butt"))
grid.segments(x, 0, x, y - .05, 
              gp=gpar(col=col, fill=col, lwd=15, lineend="butt"))
grid.text(entries$count, x, unit(10, "mm"), gp=gpar(fontface="bold"))
grid.rect()
pushViewport(viewport(y=unit(1, "npc") + unit(10, "pt"), height=unit(20, "pt"),
                      just="bottom"))
grid.text("Final Third Entries",
          gp=gpar(fontsize=20))
popViewport()
pushViewport(viewport(y=unit(1, "npc") + unit(40, "pt"), height=unit(20, "pt"),
                      just="bottom"))
grid.rect(gp=gpar(fill=hcl(hue, 70, 40)))
grid.text(c("Netherlands", "South Africa"), 
          x=unit(0:1, "npc") + unit(c(2, -2), "mm"),
          hjust=c(0, 1),
          gp=gpar(col=c("pink", "orange"), fontface="bold"))
popViewport()
```

[^final-third]:
  This data visualisation is an adaptation of a graphic shown during 
  television coverage of the match.

For any example of **congruence**, 
where a visual feature naturally corresponds to
a property of the data,
there is an opportunity for **dissonance** if the visual feature
is used inappropriately.  We can improve the effectiveness of 
a data visualisation by taking advantage of visual congruence
and we can reduce the effectiveness of a data visualisation 
if we create visual dissonance.

**TODO** need expansion of colour dissonance (possibly preceded by
section on colour associations?) 

@fig-stroop shows that we can also create dissonance with our mappings
of data values to colours.

```{r}
#| echo: false
#| label: fig-stroop
#| fig-cap:  The Stroop effect. Try to name the *colour* of each word.
#| fig-height: 1
grid.newpage()
pushViewport(viewport(width=.5))
grid.text(c("red", "green", "blue"), 1:3/4, 
          gp=gpar(col=c("green", "blue", "red"), fontsize=20, fontface="bold"))
```

We saw in @sec-same-same that we want to map differences in data values
to differences in visual features *and* keep everything else the
same.  Failing to follow that approach will make it harder to 
perceive changes in the data values.
However, a worse error, and 
another source of visual dissonance arises if we change
visual features *without* any underlying change in the data values.

@fig-listener-beans shows a data visualisation of the relative number of 
people aged over 100 for females versus males.
There are numerous problems with this data visualisation, but here
we will just focus on the horizontal **position** of the cake images.
The cakes for females are positioned to the left of the cakes for males,
which is acceptable because that difference in position reflects
a difference in the data---males versus females.
However, within each gender, the cakes are also staggered a small amount
to the left then to the right.
This visual difference in the cake positions does not correspond to
any difference in the data values.  We are tempted to seek meaning from the
the left-to-right staggering of the cakes when no meaning exists.

We will see other examples of visual changes that do not relate to
changes in the data in @sec-mosaic and @sec-word-cloud.

```{r}
#| echo: false
#| label: fig-listener-beans
#| fig-cap:  A data visualisation of the number of females aged over 100 relative to the number of makes aged over 100.[^listener-cakes]
#| warning: false
cake <- readPNG("Images/birthday-cake-cropped.png")
old <- data.frame(gender=c("Male", "Female"),
                  count=c(2, 5))
cakes <- function(data, coords) {
    grobTree(rasterGrob(cake, 
                        rep(coords$x, data$y + 1) + c(.05, -.05), 
                        unlist(mapply(function(x, y) {
                                          seq(0, x, length.out=y)
                                       },
                                       coords$y, data$y + 1)), 
                        just="bottom", 
                        height=.2))
}
ggplot(old) +
    geom_col(aes(x=gender, y=count), fill="transparent") +
    grid_panel(cakes, aes(x=gender, y=count - 1)) +
    scale_x_discrete(name=NULL, labels=c("Males 100+", "Females 100+")) +
    scale_y_continuous(expand=expansion(0, .05)) +
    theme(axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank(),
          panel.grid.major.y=element_line(linewidth=.5),
          aspect.ratio=1) +
    labs(title="For every five females age 100+,\nthere are two males aged 100+")
```

[^listener-cakes]:
  This plot is based on a data visualisation from the New Zealand Listener 
  October 14-20 2023.

> A data visualisation is *not* effective if it contains mappings
  that are dissonant because that can cause distraction and slower and/or
  incorrect mappings *back* to the data.

## Case study: Jittering

The Rugby World Cup is a competition that is held every four years, with the
first event taking place in 1987.
We have data on the number of points that were scored in World Cup matches
for Tier One nations (the top 10 rugby-playing nations).
@tbl-rwc-all shows a sample of these data.[^rwc]

[^rwc]:
  Attribution to kaggle data set.

```{r}
#| echo: false
#| message: false
#| label: tbl-rwc-all
#| tbl-cap: A table of the number of points scored and conceded by Tier One nations in Rugby World Cup matches, along with the team name, the global hemisphere of origin, the year, and a match identifier.  Each row represents the points scored by one team in one match.  There are 294 rows in total, with just the first 6 rows shown here.
kable(head(rwcAll[,c("team", "opposition", "year", "scored", "conceded", "match")]), digits=0)
```

@fig-jitter shows a scatterplot of the number of points scored 
by teams from different hemispheres at Rugby World Cup matches.
In this data visualisation, the data symbols are dots,
the number of points scored is mapped to the horizontal **position** of
the dots, and the hemisphere is mapped to the vertical **position**
of the dots.

To ameliorate the overlapping of data points, a random amount of "jitter"
has been
added the vertical position
of the dots.  The filled interior of each dot is also semitransparent.

```{r}
#| echo: false
#| label: fig-jitter
#| fig-cap: The number of points scored in Rugby World Cup games by teams from the northern and southern hemispheres.
ggplot(rwcAll) +
    geom_jitter(aes(x=scored, y=hemisphere), height=.2, width=0, alpha=.5) +
    scale_x_continuous(name="points scored")
```

The scatter plot in @fig-jitter is effective for mapping *back* to 
the number of points scored because each data value maps to the
horizontal position of a single data symbol.
Similarly, we can easily map *back* to the separate hemispheres.

On the other hand, the jittering means that there are visual differences
between the vertical positions of the dots that do not correspond to
any differences in data values.  For example, there are two dots representing
matches where a team from the Northern hemisphere scored zero points.
The data values for these two dots are identical (zero points scored and
northern hemisphere), but they are visually different.
The randomness of the jittering may help to dilute any temptation
to seek meaning from the artificial visual difference, but the
addition of jittering may cause confusion.

> Jittering data symbols is effective because it reduces the amount of overlap
  amongst data
  symbols, but jittering may also 
  reduce the effectiveness of a data visualisation because it introduces
  visual dissonance.

## Summary

The effectiveness of a data visualisation may depend on more than
just the accuracy and capacity of visual features.

Visual **congruence** refers to whether
a data symbol or visual feature has 
a natural correspondence with properties of the data.

Some data visualisations are effective because they are visually congruent.

Data symbols that are larger for larger data values are more effective.

Data symbols that only change if the data values change are more effective.

Visual **dissonance** refers to whether a data symbol or visual feature
contradicts properties of the data.

Some data visualisations are *not* effective because they are visually
dissonant.
