
# Mapping Structure

@fig-rep-guide shows a scatter plot of the number of clean breaks and the
number of tries for countries
at the 2023 Rugby World Cup (@tbl-rwc-2023).
This data visualisation is similar to @fig-scatter, but with an
additional encoding of the hemisphere for each country as the colour
of the data points.

The legend (guide) in @fig-rep-guide demonstrates the scale for the colour
encoding:  which colour is used to encode each hemisphere.
In @sec-guides, we described how the proximity of the labels "South"
and "North" to the coloured points in the legend allow us to
associate the colours with the different hemispheres.
The different values of hemisphere are encoded as the vertical
positions of the points and the text labels;
both the orange point and the label "South" have the same
vertical position.
 
```{r}
#| echo: false
#| label: fig-rep-guide
#| fig-cap: A scatter plot with a legend.
repcols <- pal_npg()(4)
gg <- ggplot(RWCperGame) +
    geom_point(aes(breaks, tries, colour=sphere)) +
    scale_colour_manual(values=repcols[c(1, 3)], name="hemisphere") +
    scale_x_continuous(name="clean breaks") +
    scale_y_continuous(name="tries scored") +
    theme(aspect.ratio=1)
pushViewport(viewport(height=.8))
print(gg, newpage=FALSE)
popViewport()
```

What about the horizontal position of the points and the text in
the legend?  These do not encode hemisphere values because they
are the same for both North and South.
What about the position of the legend title, "hemisphere"?

All of the elements of the legend are positioned together
and to the right of the main plot.
Their proximity to each other (and separation from
the main plot) mean that we see the legend as a separate
group (@sec-visual-model-groups).
This creates a visual structure for the plot, which helps
the viewer to navigate and comprehend the different elements
of the plot (could do with a CITE here).
This is analogous to reading a block of text, which is split into
paragraphs and sections with headings, to help the reader
navigate the text.

One way to express this is that the **position** of the elements of the legend 
**encode** the **structure** of the overall data visualisation.
Our visual system **decodes** the structure based on the proximity
of the legend elements.

In this section we look at the overall arrangement of the elements
of a plot.
An effective data visualisation has to convey information about
how to read the plot as well as information about the data itself.

**TODO** 
surely must be a call-back to facetting somewhere here?

## Mapping importance

Contrast

@sec-differences

@sec-attention

## Mapping ???

Repetition

```{r}
#| echo: false
#| label: fig-rep-none
#| fig-cap: A scatter plot with a discordant legend.
#| fig-keep: last
gg <- ggplot(RWCperGame) +
    geom_point(aes(breaks, tries, colour=sphere)) +
    scale_colour_manual(values=repcols[c(1, 3)], name="hemisphere") +
    scale_x_continuous(name="clean breaks") +
    scale_y_continuous(name="tries scored") +
    theme(aspect.ratio=1)
pushViewport(viewport(height=.8))
print(gg, newpage=FALSE)
popViewport()
grid.force()
keyPoint <- grid.grep("key::points", grep=TRUE, global=TRUE)
grid.edit(keyPoint[[1]], gp=gpar(col=repcols[2]))
grid.edit(keyPoint[[2]], gp=gpar(col=repcols[4]))
```

```{r}
#| echo: false
#| label: fig-rep-double
#| fig-cap: A scatter plot with a legend.
#| fig-keep: last
gg <- ggplot(RWCperGame) +
    geom_point(aes(breaks, tries, colour=sphere)) +
    scale_colour_manual(values=repcols[c(1, 3)], name="hemisphere") +
    scale_x_continuous(name="clean breaks") +
    scale_y_continuous(name="tries scored") +
    theme(aspect.ratio=1)
pushViewport(viewport(height=.8))
print(gg, newpage=FALSE)
popViewport()
grid.force()
keyText <- grid.grep("label::text", grep=TRUE, global=TRUE)
grid.edit(keyText[[1]], gp=gpar(col=repcols[1], fontface="bold"))
grid.edit(keyText[[2]], gp=gpar(col=repcols[3], fontface="bold"))
```

```{r}
#| echo: false
#| label: fig-rep-single
#| fig-cap: A scatter plot with a legend.
#| fig-keep: last
gg <- ggplot(RWCperGame) +
    geom_point(aes(breaks, tries, colour=sphere)) +
    scale_colour_manual(values=repcols[c(1, 3)], name="hemisphere") +
    scale_x_continuous(name="clean breaks") +
    scale_y_continuous(name="tries scored") +
    theme(aspect.ratio=1)
pushViewport(viewport(height=.8))
print(gg, newpage=FALSE)
popViewport()
grid.force()
keyText <- grid.grep("label::text", grep=TRUE, global=TRUE)
grid.edit(keyText[[1]], gp=gpar(col=repcols[1], fontface="bold"))
grid.edit(keyText[[2]], gp=gpar(col=repcols[3], fontface="bold"))
keyPoints <- grid.grep("key::points", grep=TRUE, global=TRUE)
grid.edit(keyPoints[[1]], gp=gpar(col=NA))
grid.edit(keyPoints[[2]], gp=gpar(col=NA))
```

```{r}
#| echo: false
#| label: fig-rep-title
#| fig-cap: A scatter plot with a title.
#| fig-keep: last
gg <- ggplot(RWCperGame) +
    geom_point(aes(breaks, tries, colour=sphere)) +
    scale_colour_manual(values=repcols[c(1, 3)], name="hemisphere") +
    scale_x_continuous(name="clean breaks") +
    scale_y_continuous(name="tries scored") +
    ggtitle(paste('<span style="color: ', repcols[3], '">**North**</span>',
                  "versus",
                  '<span style="color: ', repcols[1], '">**South**</span>')) +
    theme(aspect.ratio=1,
          plot.title=element_markdown(),
          legend.title=element_text(colour=NA))
pushViewport(viewport(height=.8))
print(gg, newpage=FALSE)
popViewport()
grid.force()
keyText <- grid.grep("label::text", grep=TRUE, global=TRUE)
grid.edit(keyText[[1]], gp=gpar(col=NA))
grid.edit(keyText[[2]], gp=gpar(col=NA))
keyPoints <- grid.grep("key::points", grep=TRUE, global=TRUE)
grid.edit(keyPoints[[1]], gp=gpar(col=NA))
grid.edit(keyPoints[[2]], gp=gpar(col=NA))
```

## Mapping navigation

Alignment

good gestalt

## Mapping connections

@sec-visual-model-groups

@fig-yji shows plots from Youth Justice Indicator Report
that encode different categories (from different variables)
as the same colour!

Use this as chapter intro ?
We need to consider larger questions about how elements
of a plot fit together.
Could also deliberately make the alignment bad so that we can talk about
that too?

```{r}
#| echo: false
#| label: fig-yji
youthCols <- c("#bf0000", "#2b4689", "#e1b728", "#8fbb22",
               "#0087c0", "#bdaa7c", "#f15a22")
gg1 <- ggplot(subset(crimeGroup, group != "Unknown")) +
    geom_line(aes(year, rate, colour=group)) +
    scale_colour_manual(values=youthCols[1:3]) +
    crimeAgeLineTitle +
    theme(axis.title=element_blank(),
          legend.justification="left",
          legend.margin=margin(),
          legend.title=element_blank(),
          legend.position="top")
typeTotal <- subset(crimeTypeTotal, prop > .04)
typeTotal$type <- reorder(typeTotal$type, typeTotal$prop,
                          decreasing=TRUE)
levels(typeTotal$type) <- c(levels(typeTotal$type), "Other")
typeTotal <- rbind(typeTotal, 
                   data.frame(type="Other", total=NA,
                              prop=sum(subset(crimeTypeTotal, 
                                              prop <= .04)$prop)))
gg2 <- ggplot(typeTotal) +
    geom_col(aes(x=prop, y="", fill=type), colour=figbg) +
    scale_fill_manual(values=youthCols, 
                      guide=guide_legend(position="top", ncol=1)) +
    coord_polar(direction=-1) +
    theme_void() +
    theme(plot.margin=margin(4, 0, 0, 0, "pt"),
          legend.justification="centre",
          legend.margin=margin(),
          legend.title=element_blank(),
          legend.key.size=unit(2, "mm"))
grid.newpage()
pushViewport(viewport(layout=grid.layout(1, 2, widths=2:1)))
pushViewport(viewport(layout.pos.col=1))
print(gg1, newpage=FALSE)
popViewport()
pushViewport(viewport(layout.pos.col=2))
print(gg2, newpage=FALSE)
popViewport()
```

Much more subtle example from 
787resources/Examples/2024-11-premier-league-momentum-plot
where the purple label could be below (?)
Would need to extract the data values from the plot somehow
to reproduce it (initial attempts with AI have failed).

Refer CRAP design in footnote, but use gestalt principles to lead in to
contrast, repetition (similarity), alignment (continuity), 
and proximity (proximity!).

+ "Gestalt principles"
  Todorovic (2008)

  Summaries of the main Gestalt principles.

  Like a peer-reviewed Wikipedia ?

  Possible link to alignment ... ?

  "good Gestalt principle: elements tend to be grouped together if they are
   parts of a pattern which is a good Gestalt, meaning as simple, orderly,
   balanced, unified, coherent, regular, etc as possible"

