
# Mapping Structure

@fig-rep-guide shows a scatter plot of the number of clean breaks and the
number of tries for countries
at the 2023 Rugby World Cup (@tbl-rwc-2023).
This data visualisation is similar to @fig-scatter, but with an
additional encoding of the hemisphere for each country as the colour
of the data points.

The legend (guide) in @fig-rep-guide demonstrates the scale for the colour
encoding:  which colour is used to encode each hemisphere.
 
```{r}
#| echo: false
#| label: fig-rep-guide
#| fig-cap: A scatter plot with a legend.
repcols <- scales::hue_pal()(4)
gg <- ggplot(RWCperGame) +
    geom_point(aes(breaks, tries, colour=sphere)) +
    scale_colour_manual(values=repcols[c(1, 3)], name="hemisphere") +
    scale_x_continuous(name="clean breaks") +
    scale_y_continuous(name="tries scored") +
    theme(aspect.ratio=1)
pushViewport(viewport(height=.8))
print(gg, newpage=FALSE)
popViewport()
```

@fig-rep-none 

As is usual when there is an encoding of data values as colour, 
@fig-rep-none includes a legend (guide) to show the scale for
the colour encoding:  which colours correspond to which hemisphere values.
However, @fig-rep-none
There is a problem with 

```{r}
#| echo: false
#| label: fig-rep-none
#| fig-cap: A scatter plot with a discordant legend.
gg <- ggplot(RWCperGame) +
    geom_point(aes(breaks, tries, colour=sphere)) +
    scale_colour_manual(values=repcols[c(2, 4)], name="hemisphere") +
    geom_point(aes(breaks, tries), data=subset(RWCperGame, sphere == "North"),
               colour=repcols[1]) +
    geom_point(aes(breaks, tries), data=subset(RWCperGame, sphere == "South"),
               colour=repcols[3]) +
    scale_x_continuous(name="clean breaks") +
    scale_y_continuous(name="tries scored") +
    theme(aspect.ratio=1)
pushViewport(viewport(height=.8))
print(gg, newpage=FALSE)
popViewport()
```

```{r}
#| echo: false
#| label: fig-rep-double
#| fig-cap: A scatter plot with a legend.
#| fig-keep: last
gg <- ggplot(RWCperGame) +
    geom_point(aes(breaks, tries, colour=sphere)) +
    scale_colour_manual(values=repcols[c(1, 3)], name="hemisphere") +
    scale_x_continuous(name="clean breaks") +
    scale_y_continuous(name="tries scored") +
    theme(aspect.ratio=1)
pushViewport(viewport(height=.8))
print(gg, newpage=FALSE)
popViewport()
grid.force()
keyText <- grid.grep("label::text", grep=TRUE, global=TRUE)
grid.edit(keyText[[1]], gp=gpar(col=repcols[1], fontface="bold"))
grid.edit(keyText[[2]], gp=gpar(col=repcols[3], fontface="bold"))
```


Much more subtle example from 
787resources/Examples/2024-11-premier-league-momentum-plot
where the purple label could be below (?)
Would need to extract the data values from the plot somehow
to reproduce it (initial attempts with AI have failed).

Refer CRAP design in footnote, but use gestalt principles to lead in to
contrast, repetition (similarity), alignment (continuity), 
and proximity (proximity!).

+ "Gestalt principles"
  Todorovic (2008)

  Summaries of the main Gestalt principles.

  Like a peer-reviewed Wikipedia ?

  Possible link to alignment ... ?

  "good Gestalt principle: elements tend to be grouped together if they are
   parts of a pattern which is a good Gestalt, meaning as simple, orderly,
   balanced, unified, coherent, regular, etc as possible"

