
# Executive Summary {#sec-exec .unnumbered}

```{r}
#| echo: false
#| message: false
source("colours.R")
```

```{r}
#| echo: false
#| purl: false
enc <- function(x) {
    paste0('<span class="encode">', x, '</span>')
}
dec <- function(x) {
    paste0('<span class="decode">', x, '</span>')
}
data <- function(x) {
    paste0('<span style="color: ', cols[1], '">', x, '</span>')
}
feature <- function(x) {
    paste0('<span style="color: ', cols[2], '">', x, '</span>')
}
```

::: {.exec-summary}

* Creating a data visualisation involves an 
  `r enc("**encoding**")` of
  `r data("**data values**")` 
  as `r feature("**visual features**")`.

  For example, numeric `r data("**counts**")` 
  are `r enc("**encoded**")` 
  as the `r feature("**heights of bars**")`
  to create a bar plot.

* Reading a data visualisation involves a 
  `r dec("**decoding**")` of
  `r feature("**visual features**")` 
  to obtain `r data("**data values**")`.

  For example, the `r feature("**heights**")` of bars
  in a bar plot
  are `r dec("**decoded**")` 
  to obtain `r data("**counts**")`.

* The <span style="color: #8D324A">**effectiveness**</span> 
  of a data visualisation depends on the
  effectiveness of the `r dec("**decoding**")`.

  For example, a bar plot is effective for comparing 
  `r data("**counts**")`
  because the visual system is effective at comparing the 
  `r feature("**heights**")` of bars.

* More `r enc("**complex encodings**")` produce more 
  `r feature("**complex visual features**")`.

  For example, a scatter plot `r enc("**encodes**")`
  `r data("**two numeric variables**")`
  as the x- and y-locations of points, 
  which produces the 
  `r feature("**position in space**")` of points.

* More `r feature("**complex visual features**")` 
  lead to more `r dec("**complex decodings**")`.

  For example, the 
  `r feature("**position in space**")` of points
  in a scatter plot
  is `r dec("**decoded**")` 
  as the `r data("**correlation**")` 
  between two numeric variables.
  
* An <span style="color: #8D324A">**effective data visualisation**</span>
  `r enc("**encodes**")` data values to
  visual features that facilitate an effective 
  `r dec("**decoding**")` from visual features
  back to information about the data.

:::

