---
title: A Framework for Effective Data Visualisations
subtitle: With Examples in R
author: Paul Murrell
format: html
---

```{r}
#| echo: false

library(knitr)
library(reshape2)
library(ggplot2)
library(ggsci)
```

```{r, include=FALSE}
figbg <- "#F2F2F2"
knitr::opts_chunk$set(dev='svg', 
                      dev.args=list(bg=figbg),
                      fig.width=8, fig.height=4,
                      out.width="100%")
## Set default colour palettes
scale_colour_discrete <- function(...) {
    scale_colour_npg(...)
}
options(ggplot2.continuous.colour="npg")
## Set default line width
update_geom_defaults("line", list(linewidth=1))
## Set default theme
theme_set(theme_bw())
## Match the table odd row fill
theme_update(plot.background=element_rect(colour=NA, 
                                          fill=figbg),
             panel.background=element_blank(),
             panel.grid.minor=element_blank(),
             panel.grid.major=element_blank(),
             legend.background=element_blank())
```

```{r}
#| echo: false
crimeAge <- read.csv("Data/YouthCrime/crime-age.csv")
crimeAgeSimple <- subset(crimeAge, age %in% c("14", "15", "16"))
```

# Introduction

A data visualisation presents data values in a visual
format.  In an effective data visualisation, this allows
features of the data to be perceived very rapidly and without 
conscious effort.
For example, @tbl-crime shows data on the crime rate for
youth offenders in New Zealand, for the ages 14 to 16, for 
each year from 2011 to 2021.  With this purely text-based, tabular
presentation of the data, it is difficult to answer the following 
questions:

> Does the crime rate decrease monotonically for all ages?

> Is the crime rate for 15-year-olds
  always greater than the rate for 14-year-olds and less than the rate 
  for 16-year-olds?

```{r}
#| echo: false
#| message: false
#| label: tbl-crime
#| tbl-cap: A table of the number of youth offenders aged 14 to 16 from 2011 to 2021.

crimeTable <- dcast(crimeAgeSimple[c("age", "year", "rate")],
                    age ~ year)
kable(crimeTable, digits=0)
```

By contrast, with a simple line 
plot data visualisation like @fig-crime we can answer the questions
easily and immediately.

```{r}
#| echo: false
#| label: fig-crime
#| fig-cap: A line plot of the number of youth offenders aged 14 to 16 from 2011 to 2021.
ggplot(crimeAgeSimple) +
    geom_line(aes(year, rate, colour=age)) +
    theme(panel.grid.major.y=element_line(colour="black", linewidth=.1),
          aspect.ratio=1)
```

# Scope


# Sources of ideas

`fosFiles/Courses/Warwick2024/warwick-masterclass/`

`Files/Research/Leave/2024/Warwick/MasterClass/Session/README`

`Files/Research/DataVis/Mappings/README`


