
```{r}
#| echo: false
#| message: false
source("common.R")
```

# Guides {#sec-guides}

@fig-crime-guides is a version of @fig-crime with the guides highlighted
in purple.  There are tick marks and tick labels on the x-axis and y-axis,
there are horizontal grid lines, and there are legend elements to show the
mapping from the different ages to different colours.

```{r}
#| echo: false
#| label: fig-crime-guides
#| fig-cap: A line plot of the crime rate amongst youth offenders aged 14 to 16 from 2011 to 2021. The guides in this plot have been highlighted in purple.
#| fig-keep: last
ggCrimeGuides <- ggplot(crimeAgeSimple) +
    geom_line(aes(x=year, y=rate, colour=ageFactor), linewidth=.1) +
    crimeAgeLineTitle +
    crimeAgeLineXlab +
    crimeAgeLineYlab +
    geom_line(aes(x=year, y=rate, group=ageFactor), colour="grey90") +
    scale_x_continuous(breaks=seq(2012, 2020, 2)) +    
    scale_colour_manual(name="age",
                        values=pal_brewer(palette="Purples")(5)[-(1:2)]) +
    theme(panel.border=element_rect(colour="grey"),
          plot.title=element_text(colour="grey"),
          plot.subtitle=element_text(colour="grey"),
          panel.grid.major.y=element_line(colour=highlight),
          axis.line=element_line(colour="grey"),
          axis.ticks=element_line(colour=highlight),
          axis.text=element_text(colour=highlight, face="bold"),          
          axis.title=element_text(colour="grey"),
          legend.text=element_text(colour=highlight, face="bold"),
          legend.title=element_text(colour="grey"),
          aspect.ratio=1)
pushViewport(viewport(width=.9, height=.9))
print(ggCrimeGuides, newpage=FALSE)
popViewport()
grid.force()
grid.edit("key::segments", grep=TRUE, global=TRUE,
          gp=gpar(lwd=2))
```

The first thing to note about the guides in @fig-crime-guides is that they
are all composed of **data symbols**, with **data values** mapped to
**visual features** of the data symbols (@sec-mapping-visual-features).
This is perhaps easiest to see with
the grid lines.  Each grid line is a data symbol where a data value,
a rate of offending, 
has been encoded as the vertical **position** of the line.
Similarly, each tick mark on the y-axis
is a line data symbol where a rate of offending
has been encoded as the vertical **position** of the line
and each tick mark on the x-axis is a line data symbol where a 
year has been encoded as the horizontal **position** of the line.
Finally, each of the tick labels on the y-axis is a text data symbol
where a rate of offending has been encoded as the vertical **position**
of the text *and* as the **shape** of the text.  Similar mappings
apply to the
x-axis tick labels.
The legend is also made up of line data symbols and text data symbols,
this time with age data values encoded as the **colour** of lines
and the **shape** of text.

The **decoding** of the **shape** of the **text** data symbols 
within the guides shows the
critical role that text plays within guides.
If we look again at @fig-no-text, there are coloured lines on the plot
and there are coloured lines on the legend.
The fact that these lines have different colours indicates that
there are three different groups, but the colours alone
do not tell us what the groups are.  We can decode different values
from the colours of the lines, but we cannot decode exact data values.
In the original @fig-crime, the shapes of the text labels on the legend
allow us to decode the original data values---the three ages 14, 15, and 16.

**TODO** text is typically the **only** data symbol within a data visualisation
that can be decoded to an **individual**, **absolute** data value.  
Most other decoding is 
relative (e.g., @sec-rel-pos, @sec-perception-relative, @sec-colour-relative)
or to a data summary (e.g., @sec-line-plots)

The proximity of the text labels in the legend to the coloured lines
in the legend means that we associate each label with each
coloured line (@sec-visual-model-groups).
So the text labels in the legend are the essential data symbols
in @fig-crime because only they allow us to decode from the colours of the lines
to the three age groups.

The text labels on axes perform a similar service.
For example, the numbers on the axes in @fig-crime
are the only data symbols that provide exact decoding from
positions to data values (years or rates).
The lines within the plot, by themselves, only show that values 
are decreasing.
It is only the relative position of the lines within the plot
to the tick labels on the axes
(and the grid lines)
that allows us to decode from the lines within the plot
to (approximate) data values.

Text data symbols within guides are effective because 
guides contain only a few data values that need to be decoded 
precisely.
This is the situation where text excels.   

> Guides are effective because they encode known data values 
  as text data symbols as well as the data symbols used in the main plot.

## Case studies

Revisit lots of plots and add interesting guides

Do variation of pcp with mini plots of corr 1 0 and -1 ?

## Summary

Guides are really just collections of data symbols, with specific
data values encoded as the visual features to demonstrate
how data values have been encoded.

The text elements within guides are essential because they provide the
only precise decoding of exact data values.  This is what allows the
decoding of the main data symbols within a data visualisation.

---

```{=latex}
\theendnotes
```
