---
title: How Data Visualisation Works
author: Paul Murrell
license: "CC BY"
toc: true
number-sections: true
tbl-cap-location: top
crossref:
  chapters: true
  sec-prefix: Section
bibliography: how-data-vis-works.bib
format: 
  html:
    css: style.css
---

```{r}
#| echo: false
#| message: false
pkgs <- readLines("packages.txt")
invisible(lapply(pkgs, library, character.only=TRUE))
```

```{r}
#| echo: false
figbg <- "#F2F2F2"
highlight <- "#7D12BA" ## Match text code colour (more precise than "purple")
## Set default colour palettes
scale_colour_discrete <- function(...) {
    scale_colour_npg(...)
}
scale_colour_ordinal <- function(...) {
    scale_colour_brewer(palette="Purples", ...)
}
scale_colour_continuous <- function(...) {
    scale_colour_distiller(palette="Purples", direction=1, ...)
}
scale_fill_discrete <- function(...) {
    scale_fill_npg(...)
}
scale_fill_ordinal <- function(...) {
    scale_fill_brewer(palette="Purples", ...)
}
scale_fill_continuous <- function(...) {
    scale_fill_distiller(palette="Purples", direction=1, ...)
}
## Set default line width
update_geom_defaults("line", list(linewidth=1))
## Set default theme
theme_set(theme_bw())
## Match the table odd row fill
theme_update(plot.background=element_rect(colour=NA, 
                                          fill=figbg),
             panel.background=element_blank(),
             panel.grid.minor=element_blank(),
             panel.grid.major=element_blank(),
             legend.background=element_blank())
```

```{r}
#| echo: false
source("rwc.R")
source("youth-crime.R")
crimeAgeSimple <- subset(crimeAge, age %in% c("14", "15", "16"))
```

```{r}
#| echo: false
#| purl: false
knitr::opts_chunk$set(dev='svg', 
                      dev.args=list(bg=figbg),
                      fig.width=8, fig.height=4,
                      out.width="100%")
```

# Executive Summary {#sec-exec}

::: {.actual-quote}

*Visualization is essentially a mapping process from computer representations
to perceptual representations, choosing encoding techniques to maximize human
understanding and communication.*[^definition]
:::

[^definition]:
  @owen1999

```{r}
#| echo: false
#| purl: false
enc <- function(x) {
    paste0('<span class="encode">', x, '</span>')
}
dec <- function(x) {
    paste0('<span class="decode">', x, '</span>')
}
```

::: {.exec-summary}

* Creating a data visualisation involves an 
  `r enc("**encoding**")` of
  <span style="color: #1A6000">**data values**</span> 
  as <span style="color: #005A8F">**visual features**</span>.

  For example, numeric <span style="color: #1A6000">**counts**</span> 
  are `r enc("**encoded**")` 
  as the <span style="color: #005A8F">**heights of bars**</span>
  to create a bar plot.

* Reading a data visualisation involves a 
  `r dec("**decoding**")` of
  <span style="color: #005A8F">**visual features**</span> 
  to obtain <span style="color: #1A6000">**data values**</span>.

  For example, the <span style="color: #005A8F">**heights**</span> of bars
  in a bar plot
  are `r dec("**decoded**")` 
  to obtain <span style="color: #1A6000">**counts**</span>.

* The <span style="color: #8D324A">**effectiveness**</span> 
  of a data visualisation depends on the
  effectiveness of the `r dec("**decoding**")`.

  For example, a bar plot is effective for comparing 
  <span style="color: #1A6000">**counts**</span>
  because the visual system is effective at comparing the 
  <span style="color: #005A8F">**heights**</span> of bars.

* More `r enc("**complex encodings**")` produce more 
  <span style="color: #005A8F">**complex visual features**</span>.

  For example, a scatter plot `r enc("**encodes**")`
  <span style="color: #1A6000">**two numeric variables**</span>
  as the x- and y-locations of points, 
  which produces the 
  <span style="color: #005A8F">**position in space**</span> of points.

* More <span style="color: #005A8F">**complex visual features**</span> 
  lead to more `r dec("**complex decodings**")`.

  For example, the 
  <span style="color: #005A8F">**position in space**</span> of points
  in a scatter plot
  is `r dec("**decoded**")` 
  as the <span style="color: #1A6000">**correlation**</span> 
  between two numeric variables.
  
* An <span style="color: #8D324A">**effective data visualisation**</span>
  <span class="encode">**encodes**</span> data values to
  visual features that facilitate an effective 
  <span class="decode">**decoding**</span> from visual features
  back to information about the data.

:::


{{< include intro.qmd >}}

{{< include visual-model.qmd >}}

{{< include mapping-visual-features.qmd >}}

{{< include quantitative.qmd >}}

{{< include qualitative.qmd >}}

{{< include congruence.qmd >}}

{{< include summaries.qmd >}}

{{< include combine.qmd >}}

{{< include visual-shapes.qmd >}}

{{< include multiple.qmd >}}

{{< include visual-objects.qmd >}}

{{< include text.qmd >}}

{{< include design.qmd >}}

{{< include review.qmd >}}

# Acknowledgements

Lots of citations in footnotes

Have to throw bouquet at Quarto.

# TODO

+ shrink lots of plots into .9x.9 viewport ?

  Also look at adding titles and tweaking axis titles everywhere
  (OR rely on figure captions instead of titles?)

+ add review at end of each section and/or preview to start of 
  each section to reinforce the building concepts.

+ Check that every visual system topic is connected to a visual decoding topic.
  Possibly rename sections, e.g., "visual differences" -> "decoding differences"

+ Flesh out all footnotes

+ Add alt text to ALL figures (fig-alt)

+ Check that colour palettes are ok for CVD

+ review https://friendly.github.io/6135/lectures/GraphPerception.pdf
  from https://friendly.github.io/6135/index.html
  (especially the stuff on basic models of visual perception
  and cognition!)
  to see that (i) I am covering everything and (ii) I am not
  telling too many lies.

* Add another apology (in scope?) that I am ignoring the "fun" or "joy" of
  data vis and point to some people like Jonathon Schwabish and
  Flowing Data

* When talk about basic data types and what you can do with them,
  maybe introduce the explicit idea of "comparison" and all tasks
  basically being based on comparison (with some examples of breaking
  down tasks).
  There is a nice simple list of tasks (rather than some overblown
  task taxonomy) in 
  "Low-Level Components of Analytic Activity in Information Visualization"
  Amar et al (2005)
  which describes comparison as a "fundamental cognitive action" that
  underlies all data vis tasks.

* Check whether HDVW covers all of the types of tasks in 
  "Low-Level Components of Analytic Activity in Information Visualization"
  Amar et al (2005)

* In visual-model, make a BRIEF statement at the end of each section 
  (maybe in black box) about the implications for data vis. 

* Add gestalt simplicity and/or symmetry in visual-model?
  tend to see symmetric elements as part of the same group.
  tend to see the simplest explanation (simple shapes combined rather
  than some complex mess).
  tendency to prefer order and simplicity.
  and link back to it from simple vs complex data symbols and a layout section
  and discussions of ordering variables (to make simpler shapes)?
  gestalt continuity could go in too?

+ "Rethinking the Ranks of Visual Channels"
  McColeman et al (2021)
  Make a bigger deal (at least in footnotes) of the lack of experimental
  results for tasks other than A:B (especially with 
  higher-level visual objects).
  
* "Low-Level Components of Analytic Activity in Information Visualization"
  Amar et al (2005)
  Set of 10 data visualisation tasks
  I like the way these all correspond to a numerical summary.
  Could use that in my data summaries/visual summaries section?
 
* Point out early in visual features that a bar plot is quite simple
  The differences are in the bars so those differences are easy to spot
  Refer back to visual model
  Also refer back from same is same
  Also refer back from wherever discuss data ink ratio perhaps not by that name

* memory limits:  wherever mention, add footnote that acknowledges 
  some controversy (different models and number of items limit
  vs continuous resource limit), with ref to e.g.,
  "Psychophysical scaling reveals a unified theory of visual memory strength"
  Schurgin et al (2020)
  and the binary map model 

* MUST acknowledge the mapping of differences to angles in a line plot
  (essentially an emergent feature!?).

* In colour, mention "In Color Perception, Size Matters"
  Stone (2012)
  "In designing colors for digital visualization systems. one of the most
   critical factors is the interaction between size and color appearance."

* Add Adobe aesthetic colour palette guidelines.  https://color.adobe.com/
  Add order of colours in color palette for bar plot. Bartolomeo et al (2025)
  To colour section.
  Or to design section if that makes more sense.

* Adam's section 4.4 Scaling and Encoding makes important points
  about the mapping from data values to visual properties
  (particularly mappings to positions). He breaks the transformation
  into three separate functions!  So something needs to be said
  about the complexity of some mappings.  Note that Adam is focused
  on efficiency - e.g., his three functions are less complex than grid's
  nested viewports.

* Move the rwcAll table from jittering to more is more

* REMOVE the `code` formatting of data variable names in 
  mapping-visual-features.qmd
  (and anywhere else)

* add link (in footnote) to Pinker 1990, p. 111 about 
  the importance of data type (from section on Quantitative features)

* Add the purpose of each type of plot as they are introduced so that 
  explanation of effectiveness is relative to it's purpose
  Eg a bar plot is effective at... Because... 

* add footnote link to Simkin and Hastie (1987) to part-to-whole for pie charts
  (they found empirical evidence that people judge proportions better on 
  pie charts than bar charts)

* Add citations to MacEachren and Bertin (e.g., in choropleth maps)

* Add (at least in footnotes) the 
  data > info > knowledge > understanding > wisdom 
  hierarchy (Ackoff, 1989).
  It corresponds quite nicely to the 
  visual feature > shape > object hierarchy ?
  Do this in text.qmd ?
  => text can be used for range of things, not just data or metadata

* acknowledge (in footnote) 
  that there is not a one-to-one correspondence between
  visual features that are initially perceived by visual system and
  the visual channels of Cleveland and Mcgill, but it is a useful
  simplification to elide them

* Add case study: interaction plots to visual-shapes ?
  => get to talk about line plots actually being slopes/angles (emergent
     features) and can add links to carswell and wickens (1990) that
     line plots (between two lines) are configural

* Interesting case study might be matrix versus two columns with 
  connecting lines.
  I find the latter better but can I explain why? 
  Example from the brath and banissi article linking text aesthetics to 
  standard aesthetics like length

* somehow mention primacy of position in space in visual-model (with cite) ?

* add mention in multiple mappings about implicit borders around
  small multiples and add footnote link to gestalt closure.
  OR add gestalt closure to visual-model and refer back to it
  AND refer back to it from the jittering case study ?

* add citation for extinction illusion

* add citation to MASSVIS for fixation images

* Add mention that mosaic plots can easily show independence or lack of ?

* add mention of anscombe's quartet to summaries.qmd to compare
  numerical with visual summaries.

* Need to talk about "mapping importance" - use basic ideas of visual
  attention, e.g., colour, to make sure that the most important items get the
  most attention.

* add plot of differences between lines in crimegroup line plot ?
  (do NOT perceive vertical difference)
  citation?  see franconeri 2021

* add footnote about visual feature vs visual channel vs visual dimension
  terminology soup.

* Lines imply continuity (at least) if not linear interpolation
  (density plot this is absolutely ok)

* need citations for more is more and dark is more and part-to-whole.

* Mention that density plots are congruent because the peaks
  are like mountain ranges ?

* Mention memorability of data vis somewhere;
  possibly only needs to be a footnote addition to, e.g., intro
  to congruence.Rmd, that acknowledges some other things that
  are important beyond accuracy (with cites to some of the studies
  on memorability - which are fairly few I think).
  Another place would be in text.Rmd

* Need some citations to the literature in visual-shapes.qmd

  + "Understanding Charts and Graphs"
  Kosslyn (1989)
    [somewhere after p 205]
    + Lines are objects
    "For example, if one wants a reader to see an interaction, a line graph is
     better than a bar graph: not only is a line a single perceptual unit, and
     hence there will be less to hold in short-term memory, but we are good at
     detecting slope differences-which convey the relevant information (see
     Pinker, 1988)."

  Link to "emergent features" and "gestalts" in, e.g.,  
  Pomerantz and Cragin (2014) ?

  Also link to "configural displays" (David Peebles (2008)) and
  "display proximity" (Barnett & Wickens, 1988; Carswell & Wickens, 1987;
  Wickens & Andre, 1990)
  "According to the proximity compatibility principle (Wickens &
  Carswell, 1995), representations with high display proximity fa-
  cilitate high mental proximity tasks requiring the integration of
  information from multiple sources. The relationship between dis-
  play and mental proximity has been revealed in several studies
  (e.g., Barnett & Wickens, 1988; Bennett & Flach, 1992; Bennett,
  Toms, & Woods, 1993; Carswell & Wickens, 1987; Wickens &
  Andre, 1990; Wickens & Carswell, 1995) and it is argued that the
  advantage comes when emergent features (Pomerantz & Pristach,
  1989) of the representation indicate additional information about
  the domain."
  He also mentions line plots as being in between separable bar plots
  and configural radar plots.

  Some useful stuff in Franconeri et al (2021) ?

* how mappings can go wrong?  
  As an umbrella for some topics below, e.g., visual illusions?
  The main point is that an effective data visualisation will AVOID
  the things that the visual system is bad at.

* Must talk about the role of **scales** in encodings
  (not all encodings are linear, not all encodings are obvious!)

* add types of cones to colour chapter?

* CVD

* visual illusions (put in a section on "how mappings go wrong"?)
  OR add these AFTER dissonant mappings (because they are examples
  of our visual system telling us something that is at odds with
  the explicit encoding!?)
  Refer back to luminance and colour contrast.

* redundant encodings
  Use same diagram as congruent mappings

  "Clusters Beat Trend!? Testing Feature Hierarchy in Statistical Graphics"
  VanderPlas and Hofmann (2017)
  says redundancy helps

  "Beyond Memorability: Visualization Recognition and Recall"
  Borkin et al (2015)
  also says redundancy helps

  "The Science of Visual Data Communication: What Works"
  Franconeri et al (2021)
  says it doesn't (though only with expert opinion like Tukey to back it up.
  Exception is to allow for colour-blind viewers.

* non-linear scales.
  separate "position" section?
  NOTE that there ARE some situations where non-linear position.
  scales are ok, e.g., log-log.
  Length is still dodgy. 
  You DO still get ordinal info, even if cannot decode sizes of differences.
  Non-linearity might actually be ok on non-position scales, e.g., 
  a colour scale for numeric data, where differences are smaller at
  one end;  all you are after, if you are using colour, is ordinal,
  so changing colours more swiftly at one end of the scale is ok (?)
  At least some 'colorspace' functions allow non-linear palettes.

* Example where using legend is BETTER than direct labelling:
  facetted bar plot.  Instead of multiple repeated sets of axis labels,
  just use coloured bars, plus single legend.

* big data (or even medium data) - need better name
  => overlapping and summary plots like boxplots (then violin plots)

* use of white space around points in scatter plot

* visual objects
  pictures as data points
  icons 
  higher-level => gain more complex information transfer
  BUT lose visual summaries (cos not just simple shapes with simple features)?
  case study: chernoff faces

* deliberately losing information (e.g., luminance scale for quant)

  "Why Shouldn’t All Charts Be Scatter Plots?
   Beyond Precision-Driven Visualizations"
  Bertini et al (2021)
  Claims that ordinal comparisons more common in practice than
  quantitative comparisons (?)



* clutter
  link back to visual objects ?
  put within visual-objects?

* Layout/graphic design as Mapping Structure
  Case study crap design
  
  Highlighting as Mapping Importance

* layout (arrangement of elements of plot)
  come at it from visual-model perspective.
  alignment, proximity, repetition are all gestalt
  (contrast is just make things different that should be different + vice versa)
  overriding theme is organisation and visual simplicity
  alignment is continuity ?

  One way to look at this is that, when you map data values to a data symbol
  you leave some (often several) visual features of the data symbol 
  undetermined.  How to choose values for those other visual features ?

  Could also have a bit in here about highlighting to direct attention
  and back ref to visual attention in visual-model

  As example do facetted crime district line plots with panel per year
  (or even just single line plot)
  PLUS map of regions PLUS colour scale to match panel strip fill colour
  (or line colours) with region fill colour on map

  Another example is @fig-junk where the left edge and right and bottom
  edges of the criminal are aligned with the left and right and bottom edges 
  of the regions AND the borders on the regions are thicker and black
  so that the criminals and the regions match each other better.
  ALSO the legend has a border so ALL lines are the same
  (keep everything the same that should NOT be attended to;
   differences are seen as significant).

* somewhere in intro (?) as a footnote (?)
  cite Sonning (2022), which goes into great detail about how
  to do a line plot right;  this document aims to provide a simpler
  set of concepts that can be applied to any data vis by a less
  expert reader.

* Mention that little is known about the accuracy of visual objects
  (though they must have good capacity because they are more abstract, right?)

  In general, there needs to be more explicit acknowledgement of the lack
  of experimental results for complex data vis scenarios.
  c.f., "Testing Statistical Charts: What Makes a Good Graph?"
  Vanderplas, Cook, and Hofmann (2020) p 71.


* aspect ratio section (in combine?)

* bubble plots (area not effective => something better?)

* Add citation for distance effect

* Add back-ref in Labels in Text to visual attention in visual-model 

* Add annotation example in text that just labels with text each country
  in @fig-scatter 

* Need citation for more-is-more (kosslyn?)
  AND for representing negative values
  AND for dark is more
  AND for same is same (Principle of unambiguity ?)

* Dark is more

  @fig-neg-tile also demonstrates another example of visual congruence because
  darker colours are naturally associated with larger data 
  values.

* Proper citation for stroop effect ?

* Use reordering of @fig-side-bar-year (so that they start tallest to smallest)
  THEN should see change in that pattern later on more easily?
  Use this as an example of "visual shape" in a bar plot.
  Revisit heatmaps - regions/shapes of colours?

* Cite "Testing Statistical Charts: What Makes a Good Graph?"
  Vanderplas, Cook, and Hofmann (2020) 
  when mention that there are many fewer studies of higher-level
  mappings between data and data vis
  (partly just because it is harder to study!).

* Could also mention in the "scope" that we are assuming that the data
  we have is sufficient for the question(s) of interest and contains
  no "substantive" problems (in the sense of Healy Chapter 1).

* Need citation for contrast effect

* Need citation for chernoff faces

* case study of density plots
  (revisit histograms in terms of shape as well; point out that 
   density plots are preferred to histograms?
   revisit boxplots too and talk about violin plots?)

* check have not missed any major pros or cons for choropleth maps
  https://www.practicalreporting.com/blog/2024/8/21/ive-stopped-using-colored-region-choropleth-maps-should-you

* add citation for choropleth maps

* case study for spinograms

* add mention of underestimation of correlation in scatter plots
  (with citation) in a footnote

* add examples of making the WRONG thing important (e.g., use thick lines
  on bar borders or bright colours on unimportant elements) ?

* Case study: Heatmaps

Heatmaps
are a common way of representing three-dimensional data where 
two of the data variables are either qualitative 
or have been measured on a regular grid.
The latter two variables naturally map to horizontal and vertical
**position** to produce an array of
rectangles, while the third variable maps to **colour**
(see @fig-mappings-2).

The mappings from data values to visual features for a heatmap
are ...

colour is not quant but can convey trends

get contrast effect but again not typically trying to match cells

get regions/shapes of colour => higher-level structures

talk about colour palettes that have deliberate colour boundaries
to emphasise structure - combination of order and category

* citation for data statistics to GoG

* citation for ensembles in statistics.qmd

* citation for "stacked dot plot" in statistics.qmd
  and/or use a different name

* case study: barycentric plots (ternary plots)

* Case study: Line of best fit

  Nice example of a complex calculation, the results of which are what gets
  visualised.  Also example of performing the calculation numerically rather
  than asking the viewer to perform it visually.  This was in summaries.qmd,
  but it cannot be there because it requires scatter plots, which come in
  the next section.
  Add this after mention of visualising models?

  Add footnote to Hadley's more complex model visualisation paper
  https://vita.had.co.nz/papers/model-vis.pdf
  https://dl.acm.org/doi/10.1002/sam.11271

* Case study: SinaPlots

A SinaPlot is a nice combination of decode for raw data and for data summary.
[^sina]
@fig-sina is actually a combination of a SinaPlot and a violin plot.

[^sina]:
  @Sidiropoulos03072018

```{r}
#| echo: false
#| label: fig-sina
#| fig-cap: A SinaPlot of the points scored per game by Tier One nations at Rugby World Cups.
gg <- ggplot(rwcAll, aes(scored, y="")) +
    geom_violin() +
    geom_sina() +
    scale_x_continuous(name="points scored")
pushViewport(viewport(height=.8, width=.8))
print(gg, newpage=FALSE)
popViewport()
```

* attributions for data sources

* add example (section?) to combine.qmd 
  where either separable is NOT appropriate
  or integral is NOT appropriate.  In the mosaic plot case, the interaction
  is beneficial.
  See "dangers of combining mappings"

* do we mention effectiveness of luminance in heatmap for ensemble perception
  of regions?

* must say that chroma and luminance and hue are all integral ?

* add direct labelling to text chapter?
  some sort of ggrepel version of the flags example in visual-objects?
  relate visual objects (other than text) to direct labelling?

* model fits like lm straight line and ref Hadley's data-space vs model-space
  "data models" to mirror "data statistics"
  (revisit box plots as VERY simple models? also talk about failing
   assumptions of these models, like unimodal for boxplot)

* kiviat plots or radar plots or spider plots
  (David Peebles (2008))

* End game/summary: 
  Present process by which a data vis can be broken down.
  What data symbols are being used?  (simple or complex)
  What mappings are being used? 
  What do we want to decode? (data or summaries)
  If mapping back to raw data, are the mappings accurate?
  What can be mapped back to?
  Are mappings congruent with data?
  ...

  * Use heatmap as case study?

  * do the side-by-side pie chart idea as an example of reasoning your
    way to a new visualisation.

  * Move the discussion/revisit of tables from text section to here ?

  * Another possible application of data vis knowledge:
    Analyse one of the weird boxplot variations that look a bit like a ziggurat?

* Another example of deliberate reduction in visual feature expressiveness
  is categorical colour scheme in heat map (so at least some features
  are more obvious).

* Add reference to Ehrenberg (1977) advice on tables in footnote
  in case study on tables (in text topic)
  PLUS reference to 
  "UNDERSTANDING GRAPHS AND TABLES"
  Wainer (1992)

* Smooth lines through scatter plot case study ?
  footnote cite to Tukey for visualising phenomena rather than raw data?
  
* This nice quote ...

  > Graphics are for comparison - comparison of one kind or another - not for
    access to individual amounts. (Tukey 1993)

  ... points out that reading individual values off a plot is not usually
  the point of a data vis, though it is still a precursor to making
  a comparison.  In some ways, ordinal comparisons are more important
  than interval or ratio?


* Answer why "a picture is worth a thousand words"
  Because 1000 words is 1000 fixations, identifying 1000 visual objects,
  and then interpreting their combined meaning ...

* citations to stuff on pictographs and glyphs and icon-based plots
  (in clutter section?)

* add citations for R packages used in plots and diagrams.
  General one for R and ggplot2, then specific ones for specific plots
  Add citation to {ggsci} for default (discrete) palette

* Add citation for surjective, injective, bijective in a footnote 
  somewhere

  Adam has a nice example in his thesis Figure 4.5, a simpler 
  version of which is a line or bar plot that contains a "total"
  line or bar - the visual objects do not map neatly, and disjointly
  to subsets of the data - there is overlap.  This makes mapping 
  back to the data harder or at least more ambiguous ?

* Get visualisation of uncertainty in somewhere ?
  Variability is a data summary, which we map to a visual feature.
  Box plots already do this.
  Only opportunity to demonstrate gradients as a visual feature?

* Use the Data Rocks advice to generate a colour palette for this document ?
  
  https://buttondown.com/datarocksnz/archive/design-matters-6-the-ultimate-dashboard-colour/

* An index that links to all of the places that different sorts of plots
  are discussed ?

* Check all references, especially the ones generated by chatgpt !
&nbsp;

* add how-to-cite

* add docker container to build in
