---
title: How Data Visualisation Works
author: Paul Murrell
toc: true
number-sections: true
tbl-cap-location: top
crossref:
  chapters: true
  sec-prefix: Section
bibliography: how-data-vis-works.bib
format: 
  html:    
    css: style.css
---

```{r}
#| echo: false
#| message: false
library(dplyr)
library(knitr)
library(kableExtra)
library(reshape2)
library(ggplot2)
library(scales)
library(colorspace)
library(ggsci)
library(grid)
```

```{r}
#| echo: false
figbg <- "#F2F2F2"
highlight <- "#7D12BA" ## Match text code colour (more precise than "purple")
## Set default colour palettes
scale_colour_discrete <- function(...) {
    scale_colour_npg(...)
}
scale_fill_discrete <- function(...) {
    scale_fill_npg(...)
}
scale_fill_ordinal <- function(...) {
    scale_fill_brewer(palette="Purples", ...)
}
scale_fill_continuous <- function(...) {
    scale_fill_distiller(palette="Purples", ...)
}
## Set default line width
update_geom_defaults("line", list(linewidth=1))
## Set default theme
theme_set(theme_bw())
## Match the table odd row fill
theme_update(plot.background=element_rect(colour=NA, 
                                          fill=figbg),
             panel.background=element_blank(),
             panel.grid.minor=element_blank(),
             panel.grid.major=element_blank(),
             legend.background=element_blank())
```

```{r}
#| echo: false
source("rwc.R")
source("youth-crime.R")
crimeAgeSimple <- subset(crimeAge, age %in% c("14", "15", "16"))
```

```{r}
#| echo: false
#| purl: false
knitr::opts_chunk$set(dev='svg', 
                      dev.args=list(bg=figbg),
                      fig.width=8, fig.height=4,
                      out.width="100%")
```

{{< include intro.qmd >}}

{{< include visual-model.qmd >}}

{{< include mapping-visual-features.qmd >}}

{{< include quantitative.qmd >}}

{{< include qualitative.qmd >}}

{{< include statistics.qmd >}}

{{< include combine.qmd >}}

{{< include visual-shapes.qmd >}}

{{< include congruence.qmd >}}

# TODO

* CVD

* visual illusions
* gestalt

* redundant encodings

* Data statistics
* computational statistics versus visual statistics
* big data (or even medium data) 
  => overlapping and summary plots like boxplots

* congruence (reuse pie chart that nicely shows Maori at very close to 50%)
* ensembles (use dot plot of district to demonstrate?   Revisit heatmaps?)
* emergent features

* text
* annotations (labelling of teams in dotplot of points per game might work)

* attention
* pre-attentive pop out
* labels (otp-down)

* jittering
  which is dodgy because it involves a visual change that is NOT related
  to a change in the data.  The hope is that the randomness of the
  visual change reduces the signal.

* deliberately losing information (e.g., luminance scale for quant)

* bubble plots (area not effective => something better?)

* Things that do not change are not salient (e.g., the widths of bars in
  bar plot), BUT things that DO change are salient (so only do this to
  represent changes in the data).  This should be cited to something.
  Principle of unambiguity ?

* Add citation for distance effect

* Use reordering of @fig-side-bar-year (so that they start tallest to smallest)
  THEN should see change in that pattern later on more easily?
  Use this as an example of "visual shape" in a bar plot.
  Revisit heatmaps - regions/shapes of colours?

* Need citation for contrast effect

* A plot with BOTH points and lines is effective because the points
  provide the raw data and the lines provide the shapes

* case study of density plots
  (revisit histograms in terms of shape as well; point out that 
   density plots are preferred to histograms?
   revisit boxplots too and talk about violin plots?)

* case study for spinograms

## Case study: Heatmaps

Heatmaps
are a common way of representing three-dimensional data where 
two of the data variables are either qualitative 
or have been measured on a regular grid.
The latter two variables naturally map to horizontal and vertical
**position** to produce an array of
rectangles, while the third variable maps to **colour**
(see @fig-mappings-2).

The mappings from data values to visual features for a heatmap
are ...

colour is not quant but can convey trends

get contrast effect but again not typically trying to match cells

get regions/shapes of colour => higher-level structures

talk about colour palettes that have deliberate colour boundaries
to emphasise structure - combination of order and category

* citation for data statistics to GoG

* citation for ensembles in statistics.qmd

* citation for "stacked dot plot" in statistics.qmd
  and/or use a different name

* case study of plot with BOTH points and lines;  you get good
  mapping back to raw data with points and you get good mapping
  back to data statistics from the line (shape).

* attributions for data sources

* model fits like lm straight line and ref Hadley's data-space vs model-space
  "data models" to mirror "data statistics"
  (revisit box plots as VERY simple models? also talk about failing
   assumptions of these models, like unimodal for boxplot)



* An index that links to all of the places that different sorts of plots
  are discussed ?

&nbsp;
