---
title: How Data Visualisation Works
author: Paul Murrell
toc: true
number-sections: true
tbl-cap-location: top
crossref:
  chapters: true
  sec-prefix: Section
bibliography: how-data-vis-works.bib
format: 
  html:    
    css: style.css
---

```{r}
#| echo: false
#| message: false
library(png)
library(dplyr)
library(reshape2)
library(tidyr)
library(knitr)
library(kableExtra)
library(reshape2)
library(ggplot2)
library(gggrid)
library(scales)
library(colorspace)
library(rcartocolor)
library(ggsci)
library(grid)
library(ggmosaic)
library(GGally)
library(ggforce)
library(ggChernoff)
library(ggimage)
```

```{r}
#| echo: false
figbg <- "#F2F2F2"
highlight <- "#7D12BA" ## Match text code colour (more precise than "purple")
## Set default colour palettes
scale_colour_discrete <- function(...) {
    scale_colour_npg(...)
}
scale_colour_ordinal <- function(...) {
    scale_colour_brewer(palette="Purples", ...)
}
scale_colour_continuous <- function(...) {
    scale_colour_distiller(palette="Purples", direction=1, ...)
}
scale_fill_discrete <- function(...) {
    scale_fill_npg(...)
}
scale_fill_ordinal <- function(...) {
    scale_fill_brewer(palette="Purples", ...)
}
scale_fill_continuous <- function(...) {
    scale_fill_distiller(palette="Purples", direction=1, ...)
}
## Set default line width
update_geom_defaults("line", list(linewidth=1))
## Set default theme
theme_set(theme_bw())
## Match the table odd row fill
theme_update(plot.background=element_rect(colour=NA, 
                                          fill=figbg),
             panel.background=element_blank(),
             panel.grid.minor=element_blank(),
             panel.grid.major=element_blank(),
             legend.background=element_blank())
```

```{r}
#| echo: false
source("rwc.R")
source("youth-crime.R")
crimeAgeSimple <- subset(crimeAge, age %in% c("14", "15", "16"))
```

```{r}
#| echo: false
#| purl: false
knitr::opts_chunk$set(dev='svg', 
                      dev.args=list(bg=figbg),
                      fig.width=8, fig.height=4,
                      out.width="100%")
```

{{< include intro.qmd >}}

{{< include visual-model.qmd >}}

{{< include mapping-visual-features.qmd >}}

{{< include quantitative.qmd >}}

{{< include qualitative.qmd >}}

{{< include congruence.qmd >}}

{{< include summaries.qmd >}}

{{< include combine.qmd >}}

{{< include visual-shapes.qmd >}}

{{< include multiple.qmd >}}

{{< include visual-objects.qmd >}}

{{< include text.qmd >}}

# TODO

* Need to refine the apologies at the start - trying to provide a simple
  set of principles for explaining data vis.  Some detail smoothed,
  some detail left out.

* MUST acknowledge the mapping of differences to angles in a line plot
  (essentially an emergent feature!?).

* REMOVE the `code` formatting of data variable names in 
  mapping-visual-features.qmd
  (and anywhere else)

* add link (in footnote) to Pinker 1990, p. 111 about 
  the importance of data type (from section on Quantitative features)

* add footnote link to Simkin and Hastie (1987) to part-to-whole for pie charts
  (they found empirical evidence that people judge proportions better on 
  pie charts than bar charts)

* acknowledge (in footnote) 
  that there is not a one-to-one correspondence between
  visual features that are initially perceived by visual system and
  the visual channels of Cleveland and Mcgill, but it is a useful
  simplification to elide them

* Add case study: interaction plots to visual-shapes ?
  => get to talk about line plots actually being slopes/angles (emergent
     features) and can add links to carswell and wickens (1990) that
     line plots (between two lines) are configural

* somehow mention primacy of position in space in visual-model (with cite) ?

* add mention in multiple mappings about implicit borders around
  small multiples and add footnote link to gestalt closure.

* add citation for extinction illusion

* add citation to MASSVIS for fixation images

* Add citation to Wikipedia "Visual system" in visual-model.qmd
 
* Add mention that mosaic plots can easily show independence or lack of ?

* add mention of anscombe's quartet to summaries.qmd to compare
  numerical with visual summaries.

* add plot of differences between lines in crimegroup line plot ?
  (do NOT perceive vertical difference)
  citation?  see franconeri 2021

* add footnote about visual feature vs visual channel vs visual dimension
  terminology soup.

* need citations for more is more and dark is more and part-to-whole.

* Need some citations to the literature in visual-shapes.qmd

  Link to "emergent features" and "gestalts" in, e.g.,  
  Pomerantz and Cragin (2014) ?

  Also link to "configural displays" (David Peebles (2008)) and
  "display proximity" (Barnett & Wickens, 1988; Carswell & Wickens, 1987;
  Wickens & Andre, 1990)
  "According to the proximity compatibility principle (Wickens &
  Carswell, 1995), representations with high display proximity fa-
  cilitate high mental proximity tasks requiring the integration of
  information from multiple sources. The relationship between dis-
  play and mental proximity has been revealed in several studies
  (e.g., Barnett & Wickens, 1988; Bennett & Flach, 1992; Bennett,
  Toms, & Woods, 1993; Carswell & Wickens, 1987; Wickens &
  Andre, 1990; Wickens & Carswell, 1995) and it is argued that the
  advantage comes when emergent features (Pomerantz & Pristach,
  1989) of the representation indicate additional information about
  the domain."
  He also mentions line plots as being in between separable bar plots
  and configural radar plots.

  Some useful stuff in Franconeri et al (2021) ?

* how mappings can go wrong?  
  As an umbrella for some topics below, e.g., visual illusions?
  The main point is that an effective data visualisation will AVOID
  the things that the visual system is bad at.

* add types of cones to colour chapter?

* CVD

* visual illusions (put in a section on "how mappings go wrong"?)

* redundant encodings

* non-linear scales
  separate "position" section?

* big data (or even medium data) - need better name
  => overlapping and summary plots like boxplots (then violin plots)

* use of white space around points in scatter plot

* visual objects
  pictures as data points
  icons 
  higher-level => gain more complex information transfer
  BUT lose visual summaries (cos not just simple shapes with simple features)?
  case study: chernoff faces

* jittering
  which is dodgy because it involves a visual change that is NOT related
  to a change in the data.  The hope is that the randomness of the
  visual change reduces the signal.
  Do jittering as a case study !!! (post same is same congruence)

* deliberately losing information (e.g., luminance scale for quant)

* clutter
  link back to visual objects ?

* layout (arrangement of elements of plot)
  come at it from visual-model perspective.
  alignment, proximity, repetition are all gestalt
  (contrast is just make things different that should be different + vice versa)
  overriding theme is organisation and visual simplicity
  alignment is continuity ?

  One way to look at this is that, when you map data values to a data symbol
  you leave some (often several) visual features of the data symbol 
  undetermined.  How to choose values for those other visual features ?

  Could also have a bit in here about highlighting to direct attention
  and back ref to visual attention in visual-model

* somewhere in intro (?) as a footnote (?)
  cite Sonning (2022), which goes into great detail about how
  to do a line plot right;  this document aims to provide a simpler
  set of concepts that can be applied to any data vis by a less
  expert reader.

* bubble plots (area not effective => something better?)

* Add citation for distance effect

* Add back-ref in Labels in Text to visual attention in visual-model 

* Use reordering of @fig-side-bar-year (so that they start tallest to smallest)
  THEN should see change in that pattern later on more easily?
  Use this as an example of "visual shape" in a bar plot.
  Revisit heatmaps - regions/shapes of colours?

* Need citation for contrast effect

* Need citation for chernoff faces

* case study of density plots
  (revisit histograms in terms of shape as well; point out that 
   density plots are preferred to histograms?
   revisit boxplots too and talk about violin plots?)

* case study for spinograms

* add mention of underestimation of correlation in scatter plots
  (with citation) in a footnote

* Case study: Heatmaps

Heatmaps
are a common way of representing three-dimensional data where 
two of the data variables are either qualitative 
or have been measured on a regular grid.
The latter two variables naturally map to horizontal and vertical
**position** to produce an array of
rectangles, while the third variable maps to **colour**
(see @fig-mappings-2).

The mappings from data values to visual features for a heatmap
are ...

colour is not quant but can convey trends

get contrast effect but again not typically trying to match cells

get regions/shapes of colour => higher-level structures

talk about colour palettes that have deliberate colour boundaries
to emphasise structure - combination of order and category

* citation for data statistics to GoG

* citation for ensembles in statistics.qmd

* citation for "stacked dot plot" in statistics.qmd
  and/or use a different name

* case study: barycentric plots (ternary plots)

* attributions for data sources

* add example (section?) to combine.qmd 
  where either separable is NOT appropriate
  or integral is NOT appropriate.  In the mosaic plot case, the interaction
  is beneficial.

* add direct labelling to text chapter?
  some sort of ggrepel version of the flags example in visual-objects?

* model fits like lm straight line and ref Hadley's data-space vs model-space
  "data models" to mirror "data statistics"
  (revisit box plots as VERY simple models? also talk about failing
   assumptions of these models, like unimodal for boxplot)

* kiviat plots or radar plots or spider plots
  (David Peebles (2008))

* End game/summary: 
  Present process by which a data vis can be broken down.
  What mappings are being used? 
  What can be mapped back to?
  ...

  * Use heatmap as case study?

  * do the side-by-side pie chart idea as an example of reasoning your
    way to a new visualisation.

  * Move the discussion/revisit of tables from text section to here ?

  * Another possible application of data vis knowledge:
    Analyse one of the weird boxplot variations that look a bit like a ziggurat?

* citations to stuff on pictographs and glyphs and icon-based plots
  (in clutter section?)

* add citations for R packages used in plots and diagrams.
  General one for R and ggplot2, then specific ones for specific plots
  Add citation to {ggsci} for default (discrete) palette

* An index that links to all of the places that different sorts of plots
  are discussed ?

* Check all references, especially the ones generated by chatgpt !
&nbsp;

* add how-to-cite

* add docker container to build in
