
```{r}
#| echo: false
#| message: false
source("common.R")
```

# Executive Summary {#sec-exec}

::: {.actual-quote}

*Visualization is essentially a mapping process from computer representations
to perceptual representations, choosing encoding techniques to maximize human
understanding and communication.*[^definition]
:::

[^definition]:
  @owen1999

```{r}
#| echo: false
#| purl: false
enc <- function(x) {
    paste0('<span class="encode">', x, '</span>')
}
dec <- function(x) {
    paste0('<span class="decode">', x, '</span>')
}
```

::: {.exec-summary}

* Creating a data visualisation involves an 
  `r enc("**encoding**")` of
  <span style="color: #1A6000">**data values**</span> 
  as <span style="color: #005A8F">**visual features**</span>.

  For example, numeric <span style="color: #1A6000">**counts**</span> 
  are `r enc("**encoded**")` 
  as the <span style="color: #005A8F">**heights of bars**</span>
  to create a bar plot.

* Reading a data visualisation involves a 
  `r dec("**decoding**")` of
  <span style="color: #005A8F">**visual features**</span> 
  to obtain <span style="color: #1A6000">**data values**</span>.

  For example, the <span style="color: #005A8F">**heights**</span> of bars
  in a bar plot
  are `r dec("**decoded**")` 
  to obtain <span style="color: #1A6000">**counts**</span>.

* The <span style="color: #8D324A">**effectiveness**</span> 
  of a data visualisation depends on the
  effectiveness of the `r dec("**decoding**")`.

  For example, a bar plot is effective for comparing 
  <span style="color: #1A6000">**counts**</span>
  because the visual system is effective at comparing the 
  <span style="color: #005A8F">**heights**</span> of bars.

* More `r enc("**complex encodings**")` produce more 
  <span style="color: #005A8F">**complex visual features**</span>.

  For example, a scatter plot `r enc("**encodes**")`
  <span style="color: #1A6000">**two numeric variables**</span>
  as the x- and y-locations of points, 
  which produces the 
  <span style="color: #005A8F">**position in space**</span> of points.

* More <span style="color: #005A8F">**complex visual features**</span> 
  lead to more `r dec("**complex decodings**")`.

  For example, the 
  <span style="color: #005A8F">**position in space**</span> of points
  in a scatter plot
  is `r dec("**decoded**")` 
  as the <span style="color: #1A6000">**correlation**</span> 
  between two numeric variables.
  
* An <span style="color: #8D324A">**effective data visualisation**</span>
  <span class="encode">**encodes**</span> data values to
  visual features that facilitate an effective 
  <span class="decode">**decoding**</span> from visual features
  back to information about the data.

:::


# Acknowledgements

Lots of citations in footnotes

Have to throw bouquet at Quarto.

# TODO

+ shrink lots of plots into .9x.9 viewport ?

  Also look at adding titles and tweaking axis titles everywhere
  (OR rely on figure captions instead of titles?)

+ add review at end of each section and/or preview to start of 
  each section to reinforce the building concepts.

+ Check that every visual system topic is connected to a visual decoding topic.
  Possibly rename sections, e.g., "visual differences" -> "decoding differences"

+ Flesh out all footnotes

+ Add alt text to ALL figures (fig-alt)

+ Check that colour palettes are ok for CVD

+ review https://friendly.github.io/6135/lectures/GraphPerception.pdf
  from https://friendly.github.io/6135/index.html
  (especially the stuff on basic models of visual perception
  and cognition!)
  to see that (i) I am covering everything and (ii) I am not
  telling too many lies.

* When talk about basic data types and what you can do with them,
  maybe introduce the explicit idea of "comparison" and all tasks
  basically being based on comparison (with some examples of breaking
  down tasks).
  There is a nice simple list of tasks (rather than some overblown
  task taxonomy) in 
  "Low-Level Components of Analytic Activity in Information Visualization"
  Amar et al (2005)
  which describes comparison as a "fundamental cognitive action" that
  underlies all data vis tasks.

* Check whether HDVW covers all of the types of tasks in 
  "Low-Level Components of Analytic Activity in Information Visualization"
  Amar et al (2005)

* In visual-model, make a BRIEF statement at the end of each section 
  (maybe in black box) about the implications for data vis. 

* Add the purpose of each type of plot as they are introduced so that 
  explanation of effectiveness is relative to it's purpose
  Eg a bar plot is effective at... Because... 

* add plot of differences between lines in crimegroup line plot ?
  (do NOT perceive vertical difference)
  citation?  see franconeri 2021

* check have not missed any major pros or cons for choropleth maps
  https://www.practicalreporting.com/blog/2024/8/21/ive-stopped-using-colored-region-choropleth-maps-should-you

* attributions for data sources





* add direct labelling to text chapter?
  some sort of ggrepel version of the flags example in visual-objects?
  relate visual objects (other than text) to direct labelling?

* citations to stuff on pictographs and glyphs and icon-based plots
  (in clutter section?)

* add citations for R packages used in plots and diagrams.
  General one for R and ggplot2, then specific ones for specific plots
  Add citation to {ggsci} for default (discrete) palette

* Get visualisation of uncertainty in somewhere ?
  Variability is a data summary, which we map to a visual feature.
  Box plots already do this.
  Only opportunity to demonstrate gradients as a visual feature?

* Use the Data Rocks advice to generate a colour palette for this document ?
  
  https://buttondown.com/datarocksnz/archive/design-matters-6-the-ultimate-dashboard-colour/

* An index that links to all of the places that different sorts of plots
  are discussed ?

  Use Quarto book format (each .qmd becomes its own chapter) and
  then get search functionality for free (which is essentially 
  an automated index).

  Can also follow instructions to "publish" on GitHub pages

* Check all references, especially the ones generated by chatgpt !

* add how-to-cite

* add docker container to build in
