
# Multiple Mappings

@fig-multiple shows a data visualisation of the performance measures
for teams in the 2023 Rugby World Cup (@tbl-rwc-2023).
This visualisation is different from anything we have seen previously
because it involves multiple mappings.
The data symbol in this case is a simple data point, but we have
mapped data values to *five* different visual features of each data point.
The number of clean breaks is mapped to the horizontal **position** of 
the data points and
the number of tries is mapped to the vertical **position** of the data points.
The number of points scored is mapped to the **size** of the points, the
hemisphere is mapped to the **shape** of the points, and the number of runs
is mapped to the **colour** of the points.

```{r}
#| echo: false
#| label: fig-multiple
#| fig-cap: A data visualisation of the performance measures for teams in the 2023 Rugby World Cup.
ggplot(RWCperGame) + 
    geom_point(aes(x=breaks, y=tries, 
               shape=sphere, size=points, fill=runs)) +
    scale_x_continuous(name="clean breaks") +
    scale_y_continuous(name="tries scored") +
    scale_shape_manual(values=c(21, 24)) +
    scale_size_continuous(guide=guide_legend(position="top")) +
    theme(aspect.ratio=1,
          legend.frame=element_rect(colour="black", linewidth=.2))
```

@fig-multiple is effective in some ways because some of the mappings
are to effective visual features. For example, we can easily map back from
horizontal and vertical positions to the number of clean breaks and the
number of tries.  We can also map back from the emergent feature of **position
in space** (@sec-integral) to a correlation between clean breaks and tries.

However, other mappings are less effective.  For example, it is 
less easy to accurately map back from size 
to the number of points scored and even
harder to map back from colour to the number of runs.

One goal of visualising multiple variables like we have in 
@fig-multiple is to discover or display multivariate relationships
between the variables.  For example, we can see that there are four 
teams in the top-right corner of the plot that are all relatively
large and relatively dark in colour, which suggests that there is a
group of teams that run more, make more clean breaks, score more tries,
and score more points compared to the other teams.

This section looks at some of the different approaches to visualising
several variables at once.  How do we perceive multivariate relationships?
Can we still perceive individual data values at the same time?

## Parallel coordinates plots

One issue with @fig-multiple is that it attempts to map each data variable
to a separate visual feature.  Once we have used both horizontal
and vertical **position**, we have to start using less accurate and less
appropriate visual features, like **area** and **colour** for
quantitative data values (see @sec-mapping-visual-features).
If we attempted to map all of the variables in @tbl-rwc-2023, we could
actually run out of visual features to use (@fig-visual-features).

A different approach is to map more of the data variables
to the same visual feature.  For example, @fig-parallel shows 
a parallel coordinates plot of the data from @fig-multiple.
In this visualisation, all of the quantitative variables have been
mapped to **position**, with hemisphere mapped to colour.

```{r}
#| echo: false
#| label: fig-parallel
#| fig-cap: A data visualisation of the performance measures for teams in the 2023 Rugby World Cup.
gg <- ggparcoord(RWCperGame, 
           columns=c(11, 5, 10, 7), groupColumn="sphere",
           scale="uniminmax", alpha=.5) +
    scale_x_discrete(expand=expansion(.02)) +
    scale_colour_discrete(guide=guide_legend(reverse=TRUE))
pushViewport(viewport(width=.8, height=.8))
print(gg, newpage=FALSE)
popViewport()
```

A parallel coordinates plot maps data summaries rather than raw data values
(@sec-summaries), for example, all data values are standardised to a range
between 0 and 1.  @tbl-parallel-stats shows some of the values that
are actually mapped to visual features in @fig-parallel.
Each variable is mapped to a different 
horizontal **position** and the values within each variable are mapped 
to vertical **position**.  In this way, all quantitative data values
are mapped to **position**.

In addition, the hemisphere is mapped to **colour** and a 
separate line data symbol is drawn for each country.
This means that a visual shape is drawn for each country
(@sec-visual-shapes).

```{r}
#| echo: false
#| message: false
#| label: tbl-parallel-stats
#| tbl-cap: The data summaries that are mapped to visual features in @fig-parallel.
vars <- subset(RWCperGame, 
               select=c("country", "sphere", 
                        "runs", "breaks", "tries", "points"))
vars[3:6] <- apply(vars[3:6], 2, 
                   function(x) {
                       (x - min(x))/diff(range(x))
                   })
vars$country <- reorder(vars$country, vars$points)
long <- melt(vars, id.vars=1:2)
kable(head(long), digits=2)
```

On one hand, a parallel coordinates plot appears to
makes sense because it maps lots
of data values to **position**, which is an accurate visual feature
for quantitative data.
However, this advantage is immediately squandered because the values that
are mapped are not the raw data values
(@fig-stat-vis-decode), so a parallel coordinates plot
is not effective for mapping back to raw data values.

The effectiveness of a parallel coordinates plot comes from the visual
shapes produced by the lines.  For example, in @fig-parallel,
we are able to identify a group of four lines at the top of the plot
that are separate from the other lines.  This grouping across
at least three variables (clean breaks, tries, and points) is easier
to see compared to @fig-multiple.  We can also more clearly see
that there are two teams that have similar numbers of runs to the top four,
but then are mixed in with the other teams on the other variables.
Furthermore, the fact that many of the lines are reasonably horizontal
rather than criss-crossing each other wildly suggests that there is 
an overall correlation between all four quantitative variables, not
just in the top four teams.

> A parallel coordinates plot can be effective at identifying 
  high-dimensional groups or correlations
  because it maps multiple data values from different variables
  to a single visual shape.

## Small multiples[^small-multiples]

[^small-multiples]:
  Tufte's small multiples

One difficulty with @fig-parallel is that many of the lines overlap
each other, which makes it difficult to separately perceive 
individual visual shapes.  @fig-profile shows an alternative
approach to visualising the same data, 
a profile plot, where a shape is drawn for each country.
The shape in this case is a filled in polygon, effectively the lines
in @fig-parallel with the area below them (down to zero) filled in.

**TODO** relate to over-plotting section (wherever that ends up)

Again, this can be effective because of the overall shapes.
For example, we can see a group of
four shapes that are larger than the others.
We can also see that Argentina and Fiji appear
very similar, which was not apparent in previous plots.

```{r}
#| echo: false
#| label: fig-profile
#| fig-cap: A data visualisation of the performance measures for teams in the 2023 Rugby World Cup.
gg <- ggplot(long) +
    geom_area(aes(as.numeric(variable), value, fill=sphere)) +
    scale_x_continuous(expand=expansion(0)) +
    scale_y_continuous(expand=expansion(0)) +
    coord_cartesian(clip="off") +
    facet_wrap(vars(country), strip.position="bottom") +
    theme(panel.border=element_blank(),
          strip.background=element_blank(),
          axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank(),
          aspect.ratio=1)
pushViewport(viewport(width=1, height=1))
print(gg, newpage=FALSE)
popViewport()
```

@fig-profile uses the same data summaries as @fig-parallel
(@tbl-parallel-stats), and similar mappings: each variable is mapped
to a different horizontal **position** and the values within each variable
are mapped to vertical **position** and hemisphere is mapped to 
**colour**.  The data symbol is also similar:  multiple data values 
are mapped to a single data symbol, either a line or a polygon.

The difference in @fig-profile is that, rather than having all of the
data symbols on the same plot, each data symbol is drawn in its
own **position in space**.  This makes it easy to view each distinct
data symbol.

The group of large shapes in @fig-profile is easier to see because
the countries have been ordered by the number of points scored 
(the right-hand height of the shape).  As we have seen before,
ordering data symbols can make a big difference (@sec-distance-effect).

## Facetting[^facetting]

[^facetting]: 
  trellis multi-panels, ggplot2 facets, GoG ??

@fig-facet-before shows a line plot of the number of youth offenders per year
in different Police Districts of New Zealand.
This visualisation has some good mappings---we map 
year to horizontal **position** and count to 
vertical **position**---but also a weaker mapping:  the Police District.
We map districts to **colour**, but because there are 12 different districts,
we are exceeding the capacity of colour (@sec-capacity).  It is not
easy to differentiate all of the different colours (not helped by 
overplotting of the lines).

A better mapping for Police District would be **position**, which has
a much higher capacity, but position
has already been taken by year and count.

```{r}
#| echo: false
#| label: fig-facet-before
#| fig-cap: A data visualisation of the number of offenders per year in different Police Districts of New Zealand, from 2011 to 2021.
ggplot(crimeDistrict) +
    geom_line(aes(year, count, colour=district)) +
    scale_colour_manual(values=carto_pal(12, "Safe")) +
    theme(aspect.ratio=1)
```

@fig-facet-after shows one solution: a facetted plot.
Each district has been mapped to a **position in space**
and we draw a separate plot for each district.  This makes
it very easy to distinguish the individual districts.

One downside is that comparisons between districts are more difficult
because of counts and/or years are **unaligned** (@sec-unaligned).
In this case, we have mitigated that problem by drawing a grey line
for all districts in each plot.  Similar to grid
lines, this makes it easier to compare the line for each district
to other districts (@sec-perception-relative).

```{r}
#| echo: false
#| label: fig-facet-after
#| fig-cap: A data visualisation of the number of offenders per year in different Police Districts of New Zealand, from 2011 to 2021.
crimeDistrict$district <- 
    reorder(crimeDistrict$district, crimeDistrict$count, max)
crime2 <- crimeDistrict
crime2$district2 <- crime2$district
crime2$district <- NULL
ggplot(crimeDistrict) +
    geom_line(aes(year, count, group=district2), data=crime2, 
              col="grey", linewidth=.2) +
    geom_line(aes(year, count)) +
    facet_wrap(vars(district)) +
    theme(aspect.ratio=.8)
```

We have again ordered the panels so that districts are near to other districts
with a similar starting count of offenders (@sec-distance-effect).

## Case study: Scatterplot matrices

A form of facetting - get to see the correlation between all four variables
like in @fig-parallel.

This tells a nice story ...

* Going down the diagonal: the number of runs is correlated loosely with
  the number of clean breaks; the number of clean breaks is more
  strongly correlated with the number of tries; and the number of tries is
  very strongly correlated with the number of points.

* Going down the columns:  the number of runs is correlated loosely
  with the number of clean breaks, a little more loosely with the number
  of tries and a little more loosely still with the number of points;
  the number of clean breaks is reasonably correlated with the number
  of tries and a little more loosely with the number of points.

```{r}
#| echo: false
#| label: fig-splom
#| fig-cap: A scatter plot matrix of the quantitative performance measures for teams in the 2023 Rugby World Cup.
quants <- c("runs", "breaks", "tries", "points")
mat <- subset(RWCperGame, 
              select=c(quants, "sphere"))
## https://stackoverflow.com/questions/3735286/create-a-matrix-of-scatterplots-pairs-equivalent-in-ggplot2
mat <- mutate(mat, id = rownames(mat)) 

## Prepare data to be plotted on the x axis
x_vars <- pivot_longer(data = mat,
             cols = 1:4,
             names_to = "variable_x",
             values_to = "x")

## Prepare data to be plotted on the y axis  
y_vars <- pivot_longer(data = mat,
                       cols = 1:4,
                       names_to = "variable_y",
                       values_to = "y") 

## Join data for x and y axes and make plot
all <- full_join(x_vars, y_vars, 
          by = c("id", "sphere"),
          relationship = "many-to-many")
all$variable_x <- factor(all$variable_x, levels=quants)
all$variable_y <- factor(all$variable_y, levels=quants)
all <- subset(all, 
              variable_x != variable_y &
              variable_x != "points" &
              variable_y != "runs" &
              !(variable_x == "tries" & variable_y == "breaks"))
ggplot(all) + 
    aes(x = x, y = y, color = sphere) +
    geom_point() +
    facet_grid(variable_y ~ variable_x, scales="free") +
    theme(axis.title=element_blank(),
          aspect.ratio=1)
```

## Summary

Mention that multivariate visualisation is an area where dynamic and
interactive graphics can be useful, e.g., GGobi ?
