
# Multiple Mappings

@fig-multiple shows a data visualisation of the performance measures
for teams in the 2023 Rugby World Cup (@tbl-rwc-2023).
This visualisation is different from anything we have seen previously
because it involves multiple mappings.
The data symbol in this case is a simple data point, but we have
mapped data values to *five* different visual features of each data point.
The number of clean breaks is mapped to the horizontal **position** of 
the data points and
the number of tries is mapped to the vertical **position** of the data points.
The number of points scored is mapped to the **size** of the points, the
hemisphere is mapped to the **shape** of the points, and the number of runs
is mapped to the **colour** of the points.

```{r}
#| echo: false
#| label: fig-multiple
#| fig-cap: A data visualisation of the performance measures for teams in the 2023 Rugby World Cup.
ggplot(RWCperGame) + 
    geom_point(aes(x=breaks, y=tries, 
               shape=sphere, size=points, fill=runs)) +
    scale_x_continuous(name="clean breaks") +
    scale_y_continuous(name="tries scored") +
    scale_shape_manual(values=c(21, 24)) +
    scale_size_continuous(guide=guide_legend(position="top")) +
    theme(aspect.ratio=1,
          legend.frame=element_rect(colour="black", linewidth=.2))
```

@fig-multiple is effective in some ways because some of the mappings
are to effective visual features. For example, we can easily map back from
horizontal and vertical positions to the number of clean breaks and the
number of tries.  We can also map back from the emergent feature of **position
in space** (@sec-integral) to a correlation between clean breaks and tries.

However, other mappings are less effective.  For example, it is 
less easy to accurately map back from size 
to the number of points scored and even
harder to map back from colour to the number of runs.

One goal of visualising multiple variables like we have in 
@fig-multiple is to discover or display multivariate relationships
between the variables.  For example, we can see that there are four 
teams in the top-right corner of the plot that are all relatively
large and relatively dark in colour, which suggests that there is a
group of teams that run more, make more clean breaks, score more tries,
and score more points compared to the other teams.

This section looks at some of the different approaches to visualising
several variables at once.  How do we perceive multivariate relationships?
Can we still perceive individual data values at the same time?

## Parallel coordinates plots

One issue with @fig-multiple is that it attempts to map each data variable
to a separate visual feature.  Once we have used both horizontal
and vertical **position**, we have to start using less accurate and less
appropriate visual features, like **area** and **colour** for
quantitative data values (see @sec-mapping-visual-features).
If we attempted to map all of the variables in @tbl-rwc-2023, we could
actually run out of visual features to use (@fig-visual-features).

A different approach is to map more of the data variables
to the same visual feature.  For example, @fig-parallel shows 
a parallel coordinates plot of the data from @fig-multiple.
In this visualisation, all of the quantitative variables have been
mapped to **position**, with hemisphere mapped to colour.

```{r}
#| echo: false
#| label: fig-parallel
#| fig-cap: A data visualisation of the performance measures for teams in the 2023 Rugby World Cup.
gg <- ggparcoord(RWCperGame, 
           columns=c(11, 5, 10, 7), groupColumn="sphere",
           scale="uniminmax", alpha=.5) +
    scale_x_discrete(expand=expansion(.02)) +
    scale_colour_discrete(guide=guide_legend(reverse=TRUE))
pushViewport(viewport(width=.8, height=.8))
print(gg, newpage=FALSE)
popViewport()
```

A parallel coordinates plot maps data summaries rather than raw data values
(@sec-summaries), for example, all data values are standardised to a range
between 0 and 1.  @tbl-parallel-stats shows some of the values that
are actually mapped to visual features in @fig-parallel.
Each variable is mapped to a different 
horizontal **position** and the values within each variable are mapped 
to vertical **position**.  In this way, all quantitative data values
are mapped to **position**.

In addition, the hemisphere is mapped to **colour** and a 
separate line data symbol is drawn for each country.
This means that a visual shape is drawn for each country
(@sec-visual-shapes).

```{r}
#| echo: false
#| message: false
#| label: tbl-parallel-stats
#| tbl-cap: The data summaries that are mapped to visual features in @fig-parallel.
vars <- subset(RWCperGame, 
               select=c("country", "sphere", 
                        "runs", "breaks", "tries", "points"))
vars[3:6] <- apply(vars[3:6], 2, 
                   function(x) {
                       (x - min(x))/diff(range(x))
                   })
vars$country <- reorder(vars$country, vars$points)
long <- melt(vars, id.vars=1:2)
kable(head(long), digits=2)
```

On one hand, a parallel coordinates plot appears to
makes sense because it maps lots
of data values to **position**, which is an accurate visual feature
for quantitative data.
However, this advantage is immediately squandered because the values that
are mapped are not the raw data values
(@fig-stat-vis-decode), so a parallel coordinates plot
is not effective for mapping back to raw data values.

The effectiveness of a parallel coordinates plot comes from the visual
shapes produced by the lines.  For example, in @fig-parallel,
we are able to identify a group of four lines at the top of the plot
that are separate from the other lines.  This grouping across
at least three variables (clean breaks, tries, and points) is easier
to see compared to @fig-multiple.  We can also more clearly see
that there are two teams that have similar numbers of runs to the top four,
but then are mixed in with the other teams on the other variables.
Furthermore, the fact that many of the lines are reasonably horizontal
rather than criss-crossing each other wildly suggests that there is 
an overall correlation between all four quantitative variables, not
just in the top four teams.

> A parallel coordinates plot can be effective at identifying 
  high-dimensional groups or correlations
  because it maps multiple data values from different variables
  to a single visual shape.

## Small multiples[^small-multiples]

One difficulty with @fig-parallel is that many of the lines overlap
each other, which makes it difficult to separately perceive 
individual visual shapes.

**TODO** relate to over-plotting section (wherever that ends up)

```{r}
#| echo: false
#| label: fig-profile
#| fig-cap: A data visualisation of the performance measures for teams in the 2023 Rugby World Cup.
gg <- ggplot(long) +
    geom_area(aes(as.numeric(variable), value, fill=sphere)) +
    scale_x_continuous(expand=expansion(0)) +
    scale_y_continuous(expand=expansion(0)) +
    coord_cartesian(clip="off") +
    facet_wrap(vars(country), strip.position="bottom") +
    theme(panel.border=element_blank(),
          strip.background=element_blank(),
          axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank(),
          aspect.ratio=1)
pushViewport(viewport(width=1, height=1))
print(gg, newpage=FALSE)
popViewport()
```

profile plots

star plots

## Facetting

Start again with new multivariate plot of crime data

[^small-multiples]:
  Tufte's small multiples, Trellis multi-panel plots, facetting.

## Case study: Scatterplot matrices

A form of facetting - get to see the correlation between all four variables
like in @fig-parallel.

```{r}
#| echo: false
#| label: fig-splom
#| fig-cap: A scatter plot matrix of the quantitative performance measures for teams in the 2023 Rugby World Cup.
quants <- c("runs", "breaks", "tries", "points")
mat <- subset(RWCperGame, 
              select=c(quants, "sphere"))
## https://stackoverflow.com/questions/3735286/create-a-matrix-of-scatterplots-pairs-equivalent-in-ggplot2
mat <- mutate(mat, id = rownames(mat)) 

## Prepare data to be plotted on the x axis
x_vars <- pivot_longer(data = mat,
             cols = 1:4,
             names_to = "variable_x",
             values_to = "x")

## Prepare data to be plotted on the y axis  
y_vars <- pivot_longer(data = mat,
                       cols = 1:4,
                       names_to = "variable_y",
                       values_to = "y") 

## Join data for x and y axes and make plot
all <- full_join(x_vars, y_vars, 
          by = c("id", "sphere"),
          relationship = "many-to-many")
all$variable_x <- factor(all$variable_x, levels=quants)
all$variable_y <- factor(all$variable_y, levels=quants)
all <- subset(all, 
              variable_x != variable_y &
              variable_x != "points" &
              variable_y != "runs" &
              !(variable_x == "tries" & variable_y == "breaks"))
ggplot(all) + 
    aes(x = x, y = y, color = sphere) +
    geom_point() +
    facet_grid(variable_y ~ variable_x, scales="free") +
    theme(axis.title=element_blank(),
          aspect.ratio=1)
```

## Summary

Mention that multivariate visualisation is an area where dynamic and
interactive graphics can be useful, e.g., GGobi ?
