
```{r}
#| echo: false
#| message: false
source("common.R")
```

# Multiple Mappings

@fig-multiple shows a data visualisation of the performance measures
for teams in the 2023 Rugby World Cup (@tbl-rwc-2023).
This visualisation is different from anything we have seen previously
because it involves **multiple mappings**.
The data symbol in this case is a simple data point, but we have
mapped data values to *five* different visual features of each data point.

The number of clean breaks is encoded as the horizontal **position** of 
the data points and
the number of tries is encoded as the vertical **position** of the data points.
The number of points scored is encoded as the **area** of the data points, the
hemisphere is encoded as the **shape** of the data points, 
and the number of runs
made is encoded as the **colour** of the data points.

```{r}
#| echo: false
#| label: fig-multiple
#| fig-cap: A data visualisation of the performance measures for teams in the 2023 Rugby World Cup.
ggplot(RWCperGame) + 
    geom_point(aes(x=breaks, y=tries, 
               shape=sphere, size=points, fill=runs)) +
    scale_x_continuous(name="clean breaks") +
    scale_y_continuous(name="tries scored") +
    scale_shape_manual(values=c(21, 24)) +
    scale_size_continuous(guide=guide_legend(position="top")) +
    theme(aspect.ratio=1,
          legend.frame=element_rect(colour="black", linewidth=.2))
```

@fig-multiple is effective in some ways because some of the mappings
involve effective visual features. For example, we can easily decode
the number of clean breaks and the
number of tries from
horizontal and vertical positions.
We can also decode a correlation between clean breaks and tries
from the emergent feature of **position
in space** (@sec-integral).

However, other mappings are less effective.  For example, it is 
less easy to accurately decode number of points scored from size 
and even
harder to decode the number of runs made from colour.
These are poor encodings for quantitative data values.

In addition, we are hampered by the fact that many different
visual features are changing at once, so it is harder
to decode changes in any one visual feature (@sec-differences),
plus there are interactions between some visual features,
which can affect the decoding of individual features 
(@sec-configural).  For example, the comparison of data points
in terms of their **size** is made harder by the concurrent changes 
in **pattern**;  it is difficult to tell whether a triangle has the
same area as a circle.

One goal of visualising multiple variables like we have in 
@fig-multiple is to discover or display multivariate relationships
between the variables.  For example, we can see that there are four 
teams in the top-right corner of the plot that are all relatively
large and relatively dark in colour, which suggests that there is a
group of teams that run more, make more clean breaks, score more tries,
and score more points compared to the other teams.
However, the decoding of multivariate relationships
is not necessarily immediate
nor effortless from @fig-multiple.

This chapter looks at some of the different approaches to visualising
several variables at once.  How do we encode multiple
variables so that we can decode multivariate relationships?
And can we still decode individual data values from
individual variables at the same time?

## Scatter plot matrices {#sec-splom}

One issue with @fig-multiple is that it attempts to map each data variable
to a separate visual feature.  Once we have used both horizontal
and vertical **position**, we have to start using less accurate and less
appropriate visual features, like **area** and **colour** for
quantitative data values (@sec-accuracy).
If we attempted to map all of the variables in @tbl-rwc-2023, we could
actually run out of visual features to use (@fig-visual-features).

An alternative approach is to map more of the data variables
to the *same* visual feature. 
In particular, we can map more variables to **position**.[^reuse-pos]

[^reuse-pos]:
    The reuse of position is handled in a variety of ways within
    data visualisation frameworks.  For example, the 
    Grammar of Graphics [@wilkinson2005] allows for *collision modifiers*
    to solve overlapping
    and separately allows for *facets*.
    GoFish [@2026-gofish] introduces the idea of *geometric operators*, such as 
    stacking, to the visualisation grammar.

    In this book, we will just talk about encoding more than two
    variables to position.

@fig-splom shows a scatter plot matrix 
of the performance measures
for teams in the 2023 Rugby World Cup (@tbl-rwc-2023).
These are the same data values that are encoded in the complicated 
scatter plot in @fig-multiple.
Apart from hemisphere,
each different variable is encoded as a horizontal and vertical position
and a scatter plot is created for each combination of variables.
For example, the number of runs is encoded as the top row and the left 
column of scatter plots.
Within each scatter plot, two variables are encoded as horizontal
and vertical position.
For example, the second plot in the first row is a scatter plot
of the number runs versus the number of clean breaks.

We can decode which variables are being combined because this is
just a qualitative decoding from position.
Within each scatter plot, 
the encoding of data values as position 
allows us to decode and compare individual data values
and this provides a superior decoding for most variables compared 
to @fig-multiple.
For example, it is easy to decode from the position of data points in
the second scatter plot on the top row that the lowest number of runs 
made was about 80.  This decoding is much more difficult from the
colour of the data points in @fig-multiple.
We can also identify clusters and outliers within each scatter plot,
which is another improvement over @fig-multiple.

```{r}
#| echo: false
#| label: fig-splom
#| fig-height: 5
#| fig-keep: last
#| fig-cap: A scatter plot matrix of the quantitative performance measures for teams in the 2023 Rugby World Cup.
quants <- c("runs", "breaks", "tries", "points")
mat <- subset(RWCperGame, 
              select=c(quants, "sphere"))
## https://stackoverflow.com/questions/3735286/create-a-matrix-of-scatterplots-pairs-equivalent-in-ggplot2
mat <- mutate(mat, id = rownames(mat)) 

## Prepare data to be plotted on the x axis
x_vars <- pivot_longer(data = mat,
             cols = 1:4,
             names_to = "variable_x",
             values_to = "x")

## Prepare data to be plotted on the y axis  
y_vars <- pivot_longer(data = mat,
                       cols = 1:4,
                       names_to = "variable_y",
                       values_to = "y") 

## Join data for x and y axes and make plot
all <- full_join(x_vars, y_vars, 
          by = c("id", "sphere"),
          relationship = "many-to-many")
all$variable_x <- factor(all$variable_x, levels=quants)
all$variable_y <- factor(all$variable_y, levels=quants)
all <- subset(all, 
              variable_x != variable_y) ## &
              ## variable_x != "runs" &
              ## variable_y != "points" &
              ## !(variable_y == "tries" & variable_x == "breaks"))
ggplot(all, aes(x = x, y = y, color = sphere)) +
    geom_point() +
    scale_x_continuous(expand=expansion(.2)) +
    scale_y_continuous(expand=expansion(.2)) +
    facet_grid(variable_y ~ variable_x, scales="free") +
    theme(axis.title=element_blank(),
          aspect.ratio=1,
          strip.background=element_blank(),
          strip.text=element_blank())
grid.force()
for (i in 1:4) {
    downViewport(grid.grep(paste0("panel-", i, "-", i), 
                           grep=TRUE, viewports=TRUE))
    grid.text(quants[i])
    upViewport(0)
}
```

The emergent feature of position in
space within each scatter plot allows us to decode the correlation between each
pair of variables (@sec-integral) and
we can also get a glimpse of 
multivariate correlations by comparing multiple scatterplots.

For example,
looking down the scatter plots just above the diagonal of the matrix,
we can see that: the number of runs is correlated loosely with
the number of clean breaks; the number of clean breaks is more
strongly correlated with the number of tries; and the number of tries is
very strongly correlated with the number of points.
This suggests a multivariate correlation between all four of these 
variables.

If we look down the first column of the matrix, we
can see that:  the number of runs is correlated loosely
with the number of clean breaks, a little more loosely with the number
of tries, and a little more loosely still with the number of points.

One negative, however, is that this decoding of multivariate relationships
requires significant cognitive effort (@sec-tasks).
A scatter plot matrix is not particularly effective at allowing
us to *effortlessly* decode multivariate relationships.

> A scatter plot matrix is effective for decoding individual data values
  from multiple variables at once, plus pairwise correlations.
  However, scatter plot matrices are less effective for 
  decoding multivariate relationships.

## Facetting {#sec-facetting}

@fig-facet-before shows a line plot of the number of youth offenders per year
in different Police Districts of New Zealand.
This visualisation makes use of several effective mappings. 
For example, we encode
year as horizontal position and we encode count as
vertical position, although the line data symbols are 
**visual shapes** (@sec-line-plots), so the most effective decoding 
is to **data summaries**, like the overall downward trend for all
districts.

The encoding of police district in @fig-facet-before
is less effective.
The district is 
encoded as **colour**, but because there are 12 different districts,
we are exceeding the capacity of colour (@sec-capacity).  It is not
easy to identify all of the different districts from the 
colours of the lines (not helped by the
overplotting of the lines).

A better encoding for Police District would be **position**, which has
a much higher capacity, but, similar to @fig-multiple, we have been 
forced to use colour because position
has already been taken by year and the number of offenders.

```{r}
#| echo: false
#| label: fig-facet-before
#| fig-cap: A data visualisation of the number of offenders per year in different Police Districts of New Zealand, from 2011 to 2021.
ggplot(crimeDistrict) +
    geom_line(aes(year, count, colour=district)) +
    scale_x_continuous(breaks=seq(2012, 2020, 2)) +    
    scale_colour_manual(values=carto_pal(12, "Safe")) +
    theme(aspect.ratio=1)
```

@fig-facet-after shows a solution---a **facetted plot**---that, like
a scatter plot matrix, 
reuses position to encode additional variables.[^facetting]
In @fig-facet-after, each district is encoded as a **position in space**
and we draw a separate plot for each district.  This makes
it very easy to decode the individual districts, much easier
than it is with @fig-facet-before.

[^facetting]: 
    This method of producing a sub-plot for each level of 
    a qualitative variable (or for each combination of levels
    for multiple qualitative variables) is called *facetting*
    in the Grammar of Graphics [@wilkinson2005], but
    it is also known as *multipanel* *conditioning* plots
    or *trellis plots* [@Becker01061996].

The plot for each district also has an enclosing border,
which helps to visually differentiate between districts
(@sec-visual-model-groups) and
we have ordered the panels so that districts are near to other districts
with a similar starting count of offenders (@sec-distance-effect).

```{r}
#| echo: false
#| label: fig-facet-after
#| fig-cap: A data visualisation of the number of offenders per year in different Police Districts of New Zealand, from 2011 to 2021.
crimeDistrict$district <- 
    reorder(crimeDistrict$district, crimeDistrict$count, max)
crime2 <- crimeDistrict
crime2$district2 <- crime2$district
crime2$district <- NULL
ggplot(crimeDistrict) +
    geom_line(aes(year, count, group=district2), data=crime2, 
              col="grey", linewidth=.2) +
    geom_line(aes(year, count)) +
    scale_x_continuous(breaks=seq(2012, 2020, 4)) +    
    facet_wrap(vars(district)) +
    theme(aspect.ratio=.8)
```

One downside to @fig-facet-after
is that decoding and comparison of the number of offenders
between districts and/or years is more difficult
because the vertical and horizontal positions
are **unaligned** for some comparisons (@sec-unaligned-pos).
In this case, we have mitigated that problem by drawing a grey line
for all districts in each plot.  Similar to grid
lines, this makes it easier to compare the line for each district
to other districts (@sec-perception-relative).

## Case study: Bar plots

@fig-bar-pos-stack shows a stacked bar plot of the youth crime rate
in New Zealand over time, broken down by age group.
The data symbols are bars.
The crime rate is encoded as the (vertical) **length** of the bars
and the year is encoded as the horizontal **position** of the bars.
These mappings are supplemented by an additional encoding of
age as the vertical **position** of the bars.
In effect, the final vertical position of each bar is
a complicated encoding of crime rate
plus age group.  

```{r}
#| echo: false
#| label: fig-bar-pos-stack
#| fig-cap: A bar plot of the number of offenders in each age group over time.
gg <- ggplot(crimeAge) +
    geom_col(aes(year, rate, fill=ageFactor, group=factor(year)), 
             position="stack") +
    scale_x_continuous(breaks=seq(2012, 2020, 2), name=NULL) +
    scale_y_continuous(expand=expansion(c(0, .05)), name=NULL) +
    scale_fill_discrete(name="age") +
    crimeAgeLineTitle
pushViewport(viewport(width=.9, height=.9))
print(gg, newpage=FALSE)
popViewport()
```

An alternative is shown in @fig-bar-pos-dodge, where we have the
same data, but in the form of a side-by-side bar plot.
This simple-seeming plot demonstrates another example of 
reusing position to encode multiple data values.

The crime rate is again encoded as the lengths of the bars and
the year is encoded as the horizontal position of the bars.
These mappings are supplemented by an additional encoding of
age as the horizontal position of the bars.

```{r}
#| echo: false
#| label: fig-bar-pos-dodge
#| fig-cap: A bar plot of the number of offenders in each age group over time.
gg <- ggplot(crimeAge) +
    geom_col(aes(year, rate, fill=ageFactor), 
             colour=figbg,
             position="dodge") +
    scale_x_continuous(breaks=seq(2012, 2020, 2), name=NULL) +
    scale_y_continuous(expand=expansion(c(0, .05)), name=NULL) +
    scale_fill_discrete(name="Age") +
    crimeAgeLineTitle
pushViewport(viewport(width=.9, height=.9))
print(gg, newpage=FALSE)
popViewport()
```

In effect, the side-by-side bar plot is like a facetted plot, 
where each year is its own bar plot, with age group encoded
as horizontal position and crime rate encoded as length.
For example,
@fig-bar-pos-facet shows the same data in a plot with one facet
per year.

```{r}
#| echo: false
#| label: fig-bar-pos-facet
#| fig-cap: A bar plot of the number of offenders in each age group over time.
gg <- ggplot(crimeAge) +
    geom_col(aes(ageFactor, rate, fill=ageFactor), 
             colour=figbg) +
    scale_x_discrete(breaks=NULL, name=NULL) +
    scale_y_continuous(expand=expansion(c(0, .05)), name=NULL) +
    scale_fill_discrete(name="Age") +
    crimeAgeLineTitle +
    facet_wrap(vars(year), nrow=1) +
    theme(panel.spacing.x=unit(0, "mm"))
pushViewport(viewport(width=.9, height=.9))
print(gg, newpage=FALSE)
popViewport()
```

## Non-cartesian mappings {#sec-non-cartesian}

Another way to encode more variables using **position** is to
abandon the standard cartesian coordinate system (@sec-cartesian).
For example, @fig-3d-scatter shows a 3D scatter plot of three of 
the performance measures for teams in the 2023 Rugby World Cup (@tbl-rwc-2023).

@fig-3d-scatter takes advantage of the visual system's ability to
interpret a 2D image as a 3D scene (@sec-illusion).
For example,
we can see that the spheres in @fig-3d-scatter appear to lie roughly along
a line in space and this decodes to a multivariate correlation
between the three variables.[^3d-interactive]
It is also possible to see that there are four teams somewhat separate
in 3D space (not just in a 2D plane, as we saw in @fig-scatter).

[^3d-interactive]:
    This visual shape is more apparent if we are able to animate 
    the image.  In the online version of this book it should be possible
    to click and drag the mouse to rotate the image in space.

```{r}
#| echo: false
#| label: fig-3d-scatter
#| results: hide
#| snapshot: true
#| fig-cap: A data visualisation of three RWC variables
setupKnitr(autoprint=TRUE)
cols <- pal_npg()(2)
mat <- matrix(c(0.916227459907532, 0.189996436238289, -0.352744400501251, 
                0, -0.39954274892807, 0.498933911323547, -0.769045233726501, 
                0, 0.0298802461475134, 0.845556795597076, 0.533048689365387, 
                0, 0, 0, 0, 1), ncol=4)
open3d(userMatrix=mat)
plot3d(RWCperGame$breaks, RWCperGame$tries, RWCperGame$points,
       xlab="breaks", ylab="tries", zlab="points",
       type="n")
spheres3d(RWCperGame$breaks, RWCperGame$tries, RWCperGame$points,
          col=cols[(RWCperGame$sphere == "North") + 1])
```

The value of a 3D scatter plot lies in this ability to decode 
multivariate **data summaries** from 3D 
**visual shapes**.
Decoding individual data values is extremely difficult from 
a 3D scatter plot because data values are encoded as positions
along dimensions that are not perpendicular (@sec-cartesian), 
may not be linear (@sec-non-linear-pos),
and are not aligned (@sec-unaligned-pos).
However, that should not be the goal of such a data visualisation;
a 3D scatter plot is effective for decoding data summaries rather than
raw data values.

@fig-3d-heatmap shows another example of a 3D plot.
This is a 3D version of the heatmap from @fig-crime-heatmap,
which shows the rate of offending for different age groups 
across different years.
The crime rate is encoded both as colour and as the height of a
3D bar.

Again, the decoding of positions to individual data values is 
not very accurate with this sort of 3D plot,
although because there is a regular grid of bars, and there are not
too many bars, we do not require much accuracy to decode year and age.

The decoding of the bar heights is also inaccurate, again because 
of non-linear and unaligned scales, but also because many bars
are partially obscured.[^3d-obscure] 
This means that
we cannot accurately decode the ratio of the highest bar to the heights
of other bars;  we are limited to ordinal comparisons at best.
However, data summaries can be decoded from the overall visual shapes.
For example, the general downward trend for all ages over time
is much clearer in @fig-3d-heatmap than it was in @fig-crime-heatmap.
The implicit visual congruence with a descending stairway 
(@sec-implicit) strongly
supports the decoding of decreasing crime rates over time.

[^3d-obscure]:
    The problem of elements being obscured is a significant one.
    The angle of view for @fig-3d-heatmap and the placement of the plot
    elements was carefully chosen.
    As with the 3D scatter plot, this problem can be removed by allowing
    the 3D scene to be rotated interactively.

```{r}
#| echo: false
#| label: fig-3d-heatmap
#| fig-height: 4
#| fig-cap: A data visualisation of youth crime by age
gg <- ggplot(crimeAgeSimple) +
    geom_tile(aes(year, age, fill=rate), colour=NA) +
    ggtitle("Youth Crime in New Zealand",
            "Distinct offenders per 10,000 pop.") +
    scale_x_continuous(expand=expansion(0), breaks=seq(2012, 2020, 2)) + 
    scale_y_continuous(expand=expansion(0), breaks=14:16, position="right") +
    theme(aspect.ratio=1,
          plot.title=element_text(hjust=1),
          plot.subtitle=element_text(hjust=1, margin=margin(0, 0, 20, 0)),
          plot.title.position="plot",
          axis.title.y.right=element_text(angle=90),
          legend.position="bottom")
plot_gg(gg, width = 5, height = 5, multicore = TRUE, scale = 150, 
        zoom = .6, theta = 30, phi = 35, windowsize = c(1600, 800))
render_snapshot()
```

> A 3D plot is not effective for decoding raw data values, but
  it may be effective because the data symbols generate visual shapes
  in space, which can decode to three-dimensional data summaries,
  such as clusters, trends, and multivariate correlations.

A significant limitation of 3D plots is that they can only encode
three variables using position.  Although it is possible to
produce views of more than three dimensions on a 2D page or screen,
the implicit decoding of visual shapes is limited to
three dimensions.[^projection]

[^projection]:
    A 3D plot is just a *projection* of three dimensions into
    2D.  More than three dimensions can also be projected to 2D and
    this can result in recognisable visual shapes in 2D or 3D, 
    but decoding a meaningful data summary from those shapes 
    becomes more difficult.
    This is another case where dynamic and interactive 
    data visualisation can be very helpful
    [@grand-tour-projection-pursuit].

## Case study: Parallel coordinates plots {#sec-parallel}

@fig-parallel shows 
a parallel coordinates plot of the data from @fig-multiple.
This is another example of a non-cartesian coordinate system.
In @fig-parallel, all of the quantitative variables have been
encoded as **position**, but those positions are along a series
of vertical parallel axes, rather than just the normal two perpendicular
axes.[^parallel]

[^parallel]:
    At a deeper level, a parallel coordinates plot represents
    a single *point* in $n$-dimensional space with a *line*
    through $n$ vertices on the parallel coordinates plot
    and several points that lie on the same line in $n$-dimensional space
    translate to several lines that meet at the same point on
    a parallel coordinate plot (or a set of lines that are parallel).

    @wegman-parallel provides a discussion from the
    point of view of statistical analysis.

    @Inselberg2009 provides an extremely thorough explanation, derivation,
    and series of demonstrations, from a more abstract
    computational geometry perspective.  
    The early parts of Chapter 10 provide some accessible comments
    relevant to data visualisation.  

```{r}
#| echo: false
#| label: fig-parallel
#| fig-cap: A parallel coordinates plot of the performance measures for teams in the 2023 Rugby World Cup.  The data values from five quantitative variables are encoded as positions along five vertical, parallel axes.  For each team, the data symbol is a line connecting the positions on the axes.
gg <- ggparcoord(RWCperGame, 
           columns=c(9, 11, 5, 10, 7), groupColumn="sphere",
           scale="uniminmax", alpha=.5) +
    scale_x_discrete(expand=expansion(.02)) +
    scale_y_continuous(guide=guide_none()) +
    scale_colour_discrete(guide=guide_legend(reverse=TRUE),
                          name="hemisphere") +
    theme(axis.title=element_blank(),
          panel.grid.major.x=element_line())
grid.newpage()
pushViewport(viewport(width=.9, height=.8))
print(gg, newpage=FALSE)
popViewport()
```

The parallel coordinates plot in @fig-parallel
maps data summaries rather than raw data values
(@sec-summaries).  All data values are standardised to a range
between 0 and 1.[^parallel-coords]
@tbl-parallel-stats shows some of the values that
are actually mapped to visual features in @fig-parallel.
Each variable is encoded as a different 
horizontal **position** and the values within each variable are encoded as
vertical **position**.  In this way, all quantitative data values
are encoded as **position**.

[^parallel-coords]:
    It is possible to use unscaled axes on a parallel coordinates plot,
    though that becomes untenable if the range of the values on different
    variables varies widely.
    There are also other ways to perform the scaling, for example,
    standardisation (in the sense of subtracting the mean and dividing
    by the standard deviation).

In addition, the hemisphere is encoded as **colour** and a 
separate line data symbol is drawn for each country.
This means that a visual shape is drawn for each country
(@sec-visual-shapes).

```{r}
#| echo: false
#| message: false
#| label: tbl-parallel-stats
#| tbl-cap: The data summaries that are mapped to visual features in @fig-parallel.
vars <- subset(RWCperGame, 
               select=c("country", "sphere", 
                        "runs", "breaks", "tries", "points"))
vars[3:6] <- apply(vars[3:6], 2, 
                   function(x) {
                       (x - min(x))/diff(range(x))
                   })
vars$country <- reorder(vars$country, vars$points)
long <- melt(vars, id.vars=1:2)
kable(head(long), digits=2)
```

On one hand, a parallel coordinates plot appears to
makes sense because it encodes multiple
data values to **position**, which is an accurate visual feature
for quantitative data.
However, this advantage is immediately squandered because the values that
are encoded are not the raw data values
(@fig-stat-vis-decode), so a parallel coordinates plot
is not effective for decoding raw data values from the positions of the lines.
Furthermore, the data symbols are lines, so each data symbol encodes
multiple data values (@sec-line-plots), which makes it difficult to decode
even individual data summary values from the lines.

The effectiveness of a parallel coordinates plot comes from the visual
shapes of the line data symbols, which allow us to decode
data summaries.  For example, in @fig-parallel,
we are able to identify a group of four lines at the top of the plot
that are separate from the other lines, plus a group of three
lines at the bottom.  This grouping across
at least three variables (clean breaks, tries, and points) is easier
to see compared to @fig-multiple.  We can also more clearly see
that there are two teams that have similar numbers of runs to the top four,
but then are mixed in with the other teams on the other variables.
Furthermore, the fact that many of the lines are reasonably horizontal
rather than criss-crossing each other suggests that there is 
an overall correlation between all four quantitative variables,
especially the last three variables, and not
just amongst the top four teams.

Compared to @fig-splom, it is easier to decode multivariate
correlations from @fig-parallel and it is easier to decode
multivariate clusters and outliers.
A parallel coordinates plot can also cope with a larger number
of variables;  we just add another vertical axis rather than
having to add an entire new row and column to a scatter plot matrix.

> A parallel coordinates plot can be effective at identifying 
  high-dimensional groups or correlations
  because it maps multiple data values from different variables
  to a single visual shape.

One major issue with a parallel coordinates plot is that the
visual shapes created by the lines are extremely dependent on
the order of the variables along the x-axis.
For example, @fig-parallel-order shows exactly the same
data as @fig-parallel, but with a different ordering of the
variables.
It is now much harder to identify a group of four teams at the top
and a group of three teams at the bottom.
There is also now no clear decoding of parallel lines between variables,
so no sense of correlation between variables
(all neighbouring pairs of variables are less correlated).

```{r}
#| echo: false
#| label: fig-parallel-order
#| fig-cap: A data visualisation of the performance measures for teams in the 2023 Rugby World Cup.  The order of the axes matters.
gg <- ggparcoord(RWCperGame, 
           columns=c(5, 11, 10, 9, 7), groupColumn="sphere",
           scale="uniminmax", alpha=.5) +
    scale_x_discrete(expand=expansion(.02)) +
    scale_y_continuous(guide=guide_none()) +
    scale_colour_discrete(guide=guide_legend(reverse=TRUE),
                          name="hemisphere") +
    theme(axis.title=element_blank(),
          panel.grid.major.x=element_line())
grid.newpage()
pushViewport(viewport(width=.8, height=.8))
print(gg, newpage=FALSE)
popViewport()
```

We can also have a large influence on the shapes in a parallel
coordinates plot by inverting the direction of the axes.
For example, @fig-parallel-neg shows the same data as @fig-parallel
and with the same ordering of the variables, but with the axes
for line breaks and 
points inverted so that higher values are at the bottom and lower
values are at the top.
In other words, the positive relationships between runs and line breaks,
between line breaks and tries,
and between tries and points are now negative relationships
(a higher number of tries correlates with a lower number of negative points).

The arrangement in @fig-parallel-neg shows that a negative correlation between
two adjacent variables shows up as many crossings of lines
at a very tight location (rather than lots of mostly parallel lines
with few crossings for positive correlation).
This demonstrates that 
it can be easier to identify a negative relationship than a
positive one with a parallel coordinates plot.
For example, the tight crossing of lines between runs and (negative) 
line breaks 
and between (negative) line breaks and tries stands out
more in this plot than the roughly parallel lines for the same
pairs of variables in @fig-parallel.[^parallel-perception]
On the downside, it is now much harder to identify clusters.
For example, it is more difficult to identify 
the group of four teams that all have
high numbers of line breaks, tries, and points.

[^parallel-perception]:
    There are many variations on parallel coordinates plots that
    attempt to mitigate some of the visual artifacts that can occur.
    For example, @slope-dependent-parallel
    suggest a way to reduce the 
    higher visual impact of steeper, intersecting lines in a parallel 
    coordinates plot.

```{r}
#| echo: false
#| label: fig-parallel-neg
#| fig-cap: A data visualisation of the performance measures for teams in the 2023 Rugby World Cup.  The direction of the axes matters.
temp <- RWCperGame
temp$negpoints <- -temp$points
temp$negbreaks <- -temp$breaks
gg <- ggparcoord(temp, 
           columns=c(9, 11, 13, 10, 12), groupColumn="sphere",
           scale="uniminmax", alpha=.5) +
    scale_x_discrete(expand=expansion(.02)) +
    scale_y_continuous(guide=guide_none()) +
    scale_colour_discrete(guide=guide_legend(reverse=TRUE),
                          name="hemisphere") +
    theme(axis.title=element_blank(),
          panel.grid.major.x=element_line())
grid.newpage()
pushViewport(viewport(width=.8, height=.8))
print(gg, newpage=FALSE)
popViewport()
```

Parallel coordinates plots are often used on much larger data sets,
in which case overplotting becomes a serious problem 
(@sec-overplot).  The use of colour for different groups 
and the use of semitransparent colours for the lines can
be helpful.[^parallel-interactive]

[^parallel-interactive]:
    Parallel coordinates plots are an example of a data visualisation
    that benefits a lot from interactivity.  The ability to select 
    subsets of the lines can help with overplotting and the ability to
    rapidly experiment with different orderings 
    and directions of the axes can help with determining the
    most effective view of the data.

## Case study: Profile plots

One difficulty with @fig-parallel is that many of the lines overlap
each other, which makes it difficult to decode the separate
visual shapes.  @fig-profile shows an alternative
approach to visualising the same data, called 
a profile plot.[^small-multiples]  Similar to a parallel coordinates plot, 
a visual shape is drawn for each country, but each shape
is drawn at its own position in space and
the shape a filled polygon, effectively the lines
in @fig-parallel with the area below them filled in.

[^small-multiples]:
    The idea of drawing lots of small repeated plots,
    with a consistent scale and style is referred to by
    @tufte1983visual as *small multiples*.
    This corresponds strongly with the idea of facetting (@sec-facetting).

Like the parallel coordinates plot, this sort of data visualisation 
can be effective because of the overall shapes.
For example, we can see a group of
four shapes that are larger than the others.
We can also see that Argentina and Fiji appear
very similar, which was not easily apparent in previous plots.

```{r}
#| echo: false
#| label: fig-profile
#| fig-cap: A data visualisation of the performance measures for teams in the 2023 Rugby World Cup.
gg <- ggplot(long) +
    geom_area(aes(as.numeric(variable), value, fill=sphere)) +
    scale_x_continuous(expand=expansion(0)) +
    scale_y_continuous(expand=expansion(0)) +
    coord_cartesian(clip="off") +
    facet_wrap(vars(country), strip.position="bottom") +
    theme(panel.border=element_blank(),
          strip.background=element_blank(),
          axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank(),
          aspect.ratio=1)
pushViewport(viewport(width=1, height=1))
print(gg, newpage=FALSE)
popViewport()
```

@fig-profile uses the same data summaries as @fig-parallel
(@tbl-parallel-stats), and similar mappings: each variable is encoded as
a different horizontal **position** and the values within each variable
are encoded as vertical **position** and hemisphere is mapped to 
**colour**.  The data symbol is also similar:  multiple data values 
are mapped to a single data symbol, in this case a polygon.

The difference in @fig-profile is that, rather than having all of the
data symbols on the same plot, each data symbol is drawn in its
own **position in space**.  This makes it easy to view each distinct
data symbol.

The cluster of four larger shapes in @fig-profile is also easier to see because
the countries have been ordered by the number of points scored 
(the right-hand height of the shape).  As we have seen before,
ordering data symbols can make a big difference (@sec-distance-effect).

A major weakness with a profile plot like @fig-profile is that it
does not scale well to large data sets.  For example,
adding another country to @fig-profile would take up additional space,
but no more space would be necessary for a parallel coordinates plot
like @fig-parallel because we would just add another line.

## Dangers of multiple mappings

@fig-multiple showed that the naive approach of just using 
a separate visual feature for each data variable leads to
complex data visualisations that are difficult to decode.

We have explored some alternative data visualisations
of multiple variables that provide better decoding of
individual data values, like facetting and scatter plot matrices,
but they do not provide effortless decoding of multivariate data summaries,
like correlation and clusters.

The alternatives that provide better decoding of multivariate data summaries,
like 3D plots and parallel coordinate plots, rely on visual shapes that can
be decoded to data summaries, but sacrifice the ability to accurately decode
individual data values.
One danger is that there is still a temptation to
decode individual data values, which can result in erroneous or
misleading conclusions.

For example, in @fig-3d-scatter, we may be tempted to decode and compare
the number of clean breaks for different teams based on the positions
of the data points and the labels on the axes, but that is unlikely
to result in an accurate decoding, even for ordinal comparisons.

It is perhaps safer to make it clear that only visual shapes should be
decoded from the 3D scatter plot by removing the axis scales
altogether, as in @fig-3d-scatter-no-ticks.

```{r}
#| echo: false
#| label: fig-3d-scatter-no-ticks
#| results: hide
#| fig-cap: A data visualisation of three RWC variables
cols <- pal_npg()(2)
mat <- matrix(c(0.916227459907532, 0.189996436238289, -0.352744400501251, 
                0, -0.39954274892807, 0.498933911323547, -0.769045233726501, 
                0, 0.0298802461475134, 0.845556795597076, 0.533048689365387, 
                0, 0, 0, 0, 1), ncol=4)
open3d(userMatrix=mat)
plot3d(RWCperGame$breaks, RWCperGame$tries, RWCperGame$points,
       type="n", decorate=FALSE)
decorate3d(xlab="breaks", ylab="tries", zlab="points",
           axes=FALSE, aspect=1)
bbox3d(xlen=0, ylen=0, zlen=0, color=c("white"))
spheres3d(RWCperGame$breaks, RWCperGame$tries, RWCperGame$points,
          col=cols[(RWCperGame$sphere == "North") + 1])
```

Given the problems with 
decoding individual data values, the classic
objection to 3D plots is justified:  if only two dimensions are
required, then nothing is gained by adding a third dimension.
There are really no arguments in favour of @fig-bar-3d-bad-1
over @fig-bar-3d-bad-2.

```{r}
#| echo: false
#| label: fig-bar-3d-bad
#| layout-ncol: 2
#| fig-width: 4
#| fig-cap: A data visualisation of youth crime by age for 2021
#| fig-subcap:
#|   - 3D bar plot 
#|   - 2D bar plot
notrun <- function() {
    gg <- ggplot(subset(crimeAge, year == 2021)) +
        geom_col(aes(x=ageFactor, y=1, fill=rate)) +
        ggtitle("Youth Crime in New Zealand",
                "Distinct offenders per 10,000 pop.") +
        scale_x_discrete(name="age") + 
        scale_y_continuous(guide=FALSE, name=NULL) +
        scale_fill_continuous(limits=c(0, NA), guide=FALSE) +
        theme(aspect.ratio=1/3,
              plot.subtitle=element_text(margin=margin(0, 0, 50, 0)),
              plot.title.position="plot")
    plot_gg(gg, width = 5, height = 5, multicore = TRUE, scale = 500, 
            zoom = .6, theta = 30, phi = 35, windowsize = c(1600, 800))
    render_snapshot()
}
df <- subset(crimeAge, year == 2021)
cloud(rate ~ factor(age) + factor(year), df, panel.3d.cloud=panel.3dbars,
      zlim=c(0, max(df$rate)), xbase=.5, ybase=.5,
      scales=list(arrows=FALSE), xlab="age", ylab="year",
      col.facet="grey")
ggplot(df) +
    geom_col(aes(age, rate))
```

## Summary

When we visualise multiple variables at once, it is not
effective to map each different variable to a different visual feature
because that usually forces us to make use of a visual feature that is
either inappropriate or inaccurate.

An alternative is to reuse the same visual feature for multiple
different variables.
In particular, we can reuse **position** for more than 
just two variables, for example, to create a scatter plot matrix
or a facetted plot.
This at least allows us to accurately decode individual variables.

Another alternative is to use non-cartesian coordinates, for
example 3D plots or parallel coordinates plots.
These generate visual shapes that allow us to decode multivariate
data summaries, like multivariate clusters and multivariate correlations.
However, in order to gain multivariate data summaries,
we typically have to sacrifice the ability to accurately decode individual
data values.

No matter which approach we use,
there is still a limit to how much information can effectively be displayed at
once within a static data visualisation.
This is one area where dynamic and interactive graphics can be useful
because the viewer is able to rapidly switch between multiple views
of the data.  

---

```{=latex}
\theendnotes
```
