
# Colour

@fig-pie-hue shows a pie chart of the number of offenders for different
levels of crime seriousness (between 2011 and 2021).
In this data visualisation, level, which is a **qualitative** variable,
is mapped to **colour**, which is appropriate for representing
qualitative values (see @sec-qual-features).  It is easy to identify
different crime levels because it is easy to perceive differences between
the colours of the wedges in the pie chart.

```{r}
#| echo: false
#| label: fig-pie-hue
#| fig-cap: A data visualisation of the number of offenders for different levels of crime seriousness.
ggplot(crimeLevelTotal) +
    geom_col(aes(x=total, y="", fill=level)) +
    coord_polar() +
    scale_fill_npg(guide=guide_legend(reverse=TRUE)) +
    theme(axis.title=element_blank(),
          axis.ticks=element_blank())
```

@fig-pie-ordinal show another pie chart of the same data, again mapping
the **qualitative** level to **colour**.
It is again easy to differentiate between different crime levels because we
can easily perceive differences between the colours of the wedges.
However, @fig-pie-ordinal is *more* effective than @fig-pie-hue because
we can *also* perceive an ordering of the wedges from more serious crimes,
which are darker wedges, to less serious crimes, which are lighter wedges.

```{r}
#| echo: false
#| label: fig-pie-ordinal
#| fig-cap: A data visualisation of the number of offenders for different levels of crime seriousness.
ggplot(crimeLevelTotal) +
    geom_col(aes(x=total, y="", fill=level), colour="black") +
    coord_polar() +
    scale_fill_ordinal(guide=guide_legend(reverse=TRUE)) +
    theme(axis.title=element_blank(),
          axis.ticks=element_blank())
```

The fact that there can be two data visualisations,
@fig-pie-hue and @fig-pie-ordinal,
with the same mappings,
the **qualitative**
level of crime maps to the visual feature **colour**,
but the two data visualisations 
have different effectiveness suggests that there is still
more to learn about how different visual features work.

This section covers some more details about how **qualitative** visual
features work.

## Hue, chroma, and luminance {#sec-colour}

We have so far taken a very simplistic approach to **colour**,
treating it as if it is a single **qualitative** visual feature
(see @sec-qual-features).
However, the perception of colour can be divided into 
three separate visual features:  **hue**, **chroma**, and **luminance**
(see @fig-hcl).[^tricolour]
Hue describes what we normally refer to as the colour name: reds, oranges,
yellows, greens, blues, and purples.
Chroma describes the colourfulness, from bright to dull,
and luminance describes the lightness, from dark to light.
Our visual system is capable to perceiving changes in hue and changes in
chroma and changes in luminance.

[^tricolour]:
  The three-dimensional experience of colour arises from the fact that there
  are three different sorts of cones in the retina (see @sec-eye),
  each sensitive to a different range of light wavelengths:
  The L cones respond to long wavelengths (reds); the M cones respond
  to medium wavelengths (greens); and the S cones respond to short
  wavelengths (blues).
  The neural connections that combine the messages from the retina are also 
  divided into three pathways: light-dark, which represents the sum of L, M, 
  and S cones;
  red-green, which represents the difference between L and M cones;
  and blue-yellow, which represents the difference between S cones and the
  sum of L and M cones.
  For more information, see @ware2021information.

We have described colour as a **qualitative** visual feature so far, but that 
really strictly only applies to **hue**.  For example, 
@fig-stacked-bar used different hues to represent different ethnic groups.
Hue is excellent for the perception of differences.

```{r}
#| echo: false
#| label: fig-hcl
#| fig-cap: "The perception of colour can be divided into three components: differences in **hue**, the colours of the rainbow; differences in **chroma**, from bright and colourful to dull and grey; and differences in **luminance**, from light to dark."
vp2 <- viewport(width=.9, y=unit(3, "lines"),
                height=unit(1, "npc") - unit(6, "lines"),
                just="bottom")
pushViewport(viewport(width=.8, height=.8, layout=grid.layout(1, 3)))
pushViewport(viewport(layout.pos.col=1),
             vp2)
pushViewport(viewport(width=unit(1, "snpc"), height=unit(1, "snpc")))
n <- 7
t <- seq(0, 2*pi, length.out=n+1)[-1]
x <- .5 + .3*cos(t)
y <- .5 + .3*sin(t)
grid.circle(x, y, r=.1,
            gp=gpar(col="black", fill=hcl(t/pi*180, 70, 70)))
popViewport()
grid.text("hue", y=unit(-2, "lines"))
popViewport(2)
pushViewport(viewport(layout.pos.col=2),
             vp2)
grid.circle(.5, 0:4/4, r=.1, 
            gp=gpar(col="black", fill=hcl(300, 100*1:5/6, 70))) 
grid.text("chroma", y=unit(-2, "lines"))
popViewport(2)
pushViewport(viewport(layout.pos.col=3),
             vp2)
grid.circle(.5, 0:4/4, r=.1, 
            gp=gpar(col="black", fill=hcl(300, 70, 100*5:1/6))) 
grid.text("luminance", y=unit(-2, "lines"))
popViewport(2)
```

## Ordinal visual features {#sec-ordinal}

**Chroma** and **luminance** are both **qualitative** visual features
in that they can be used to visualise different groups, but they
are also **ordinal** visual features because we can perceive
an ordering from a set of colours that differ in terms of either chroma
or luminance or both.  A set of colours that transitions from darker
to lighter and/or brighter to duller conveys an ordering from larger
to smaller.

This helps to explain why @fig-pie-ordinal is more effective
than @fig-pie-hue.  In @fig-pie-ordinal, the colours differ
mostly in terms of luminance and chroma so we are able to perceive
changes from larger values to smaller values because the
colours transition from darker to lighter.

In this sense, **chroma** and **luminance** are able to convey
more information than **hue** because they can represent **nominal**
differences *and* **ordinal** differences, whereas hue can only
represent nominal differences.  

On the other hand, @fig-lum-cap shows that chroma and luminance are
even more limited than hue in terms of **capacity**.  We can only use
chroma and/or luminance to represent **nominal** data when there
are very few different groups to represent.

```{r}
#| echo: false
#| label: fig-lum-cap
#| fig-height: 3
#| fig-cap: Mapping qualitative values to luminance and chroma works well for representing nominal data values, but only for a very small number of values. The three values on the left are easily distringuishable, but it becomes difficult to clearly distinguish between adjacent colours in the middle and almost impossible for the colours on the right.  This figure can be compared with @fig-colour-cap, which demonstrates the capacity of hue for representing nominal data values.
clpie <- function(n) {
    t <- rev(seq(0, 2*pi, length.out=100*n) + pi/2)
    cols <- hcl.colors(n+1, "Purples")[-(n+1)]
    for (i in 1:n) {
        index <- 1:100 + (i-1)*100
        x <- cos(t[index])
        y <- sin(t[index])
        grid.polygon(c(.5, .5 + .3*x), c(.5, .5 + .3*y), 
                     gp=gpar(col=figbg, lwd=4, fill=cols[i]))
        grid.text(letters[i], .5 + mean(.25*x), .5 + mean(.25*y),
                  gp=gpar(cex=1))
    }
}
pushViewport(viewport(layout=grid.layout(1, 3, respect=TRUE)))
pushViewport(viewport(layout.pos.col=1))
clpie(3)
popViewport()
pushViewport(viewport(layout.pos.col=2))
clpie(7)
popViewport()
pushViewport(viewport(layout.pos.col=3))
clpie(11)
popViewport()
```

## Perception is relative {#sec-colour-relative}

We saw in @sec-perception-relative that perception of **lengths** is 
relative.  For example, smaller differences in length are easier
to see when the lengths themselves are short (see @fig-relative).

There is a similar rule for the perception of **colours**. 
Our perception of a colour is affected by neighbouring 
colours.[^contrast-effect]
This effect is demonstrated in @fig-contrast-effect:
the same colour when surrounded by a darker background appears lighter
than it does when surrounded by a lighter background.

[^contrast-effect]:
  This relative perception of colours is useful for being able
  to adapt to a wide range of lighting conditions.
  See @ware2021information for a more detailed discussion.

```{r}
#| echo: false
#| label: fig-contrast-effect
#| fig-height: 3
#| fig-cap: The perception of colours is relative. In the top row, the inner rectangles are exactly the same shade of grey on the left and exactly the same shade of purple on the right.  The perception of the inner rectangles is affected by the surrounding colour, so the same shade appears darker against a light background and darker against a light background.  The bottom row draws a line connecting the two inner rectangles to show that they really are the same shade.
pushViewport(viewport(layout=grid.layout(2, 2, widths=2, respect=TRUE)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1),
             viewport(width=.8, height=.8))
grid.rect(x=.25, width=.5, gp=gpar(col=NA, fill="grey20"))
grid.rect(x=.25, width=.25, height=.5, gp=gpar(col=NA, fill="grey60"))
grid.rect(x=.75, width=.5, gp=gpar(col=NA, fill="grey80"))
grid.rect(x=.75, width=.25, height=.5, gp=gpar(col=NA, fill="grey60"))
popViewport(2)
pushViewport(viewport(layout.pos.col=1, layout.pos.row=2),
             viewport(width=.8, height=.8))
grid.rect(x=.25, width=.5, gp=gpar(col=NA, fill="grey20"))
grid.rect(x=.25, width=.25, height=.5, gp=gpar(col=NA, fill="grey60"))
grid.rect(x=.75, width=.5, gp=gpar(col=NA, fill="grey80"))
grid.rect(x=.75, width=.25, height=.5, gp=gpar(col=NA, fill="grey60"))
grid.segments(.25, .5, .75, .5, gp=gpar(col="grey60", lwd=20))
popViewport(2)
purples <- pal_brewer(palette="Purples")(4)[-1]
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1),
             viewport(width=.8, height=.8))
grid.rect(x=.25, width=.5, gp=gpar(col=NA, fill=purples[1]))
grid.rect(x=.25, width=.25, height=.5, gp=gpar(col=NA, fill=purples[2]))
grid.rect(x=.75, width=.5, gp=gpar(col=NA, fill=purples[3]))
grid.rect(x=.75, width=.25, height=.5, gp=gpar(col=NA, fill=purples[2]))
popViewport(2)
pushViewport(viewport(layout.pos.col=2, layout.pos.row=2),
             viewport(width=.8, height=.8))
grid.rect(x=.25, width=.5, gp=gpar(col=NA, fill=purples[1]))
grid.rect(x=.25, width=.25, height=.5, gp=gpar(col=NA, fill=purples[2]))
grid.rect(x=.75, width=.5, gp=gpar(col=NA, fill=purples[3]))
grid.rect(x=.75, width=.25, height=.5, gp=gpar(col=NA, fill=purples[2]))
grid.segments(.25, .5, .75, .5, gp=gpar(col=purples[2], lwd=20))
popViewport(2)
popViewport()
```

This effect also applies to the perception of hue.
@fig-pie-contrast shows a pie chart of the number of offenders in each
police district, with police district mapped to the **colour** of the
wedges.  There are two issues with the perception of the wedge colours:
each wedge is affected by its neighbour so that, for example, some of 
the blue wedges look more green at the boundary with their more blue 
neighbour and more blue at the boundary with their more green neighbour;
furthermore, the colours in the legend are perceived differently from
the corresponding colours in the wedges because the legend colours are
surrounded by white.

```{r}
#| echo: false
#| label: fig-pie-contrast
#| fig-cap: This pie chart demonstrates that the perception of hues is relative. The colours within each wedge appear to have a subtle gradient within the wedge because the perception of the wedge colour is affected differently by the neighbouring wedge colours.
ggplot(crimeDistrictTotal) + 
    geom_col(aes(x=total, y="", fill=district)) +
    coord_polar() +
    scale_fill_hue() +
    theme(aspect.ratio=1,
          axis.title.y=element_blank(),
          axis.ticks.y=element_blank())
```

The pie chart in @fig-pie-contrast is *not* an effective data visualisation
because it maps qualitative data values with lots of different groups
to colour hue.  This exceeds the **capacity** of the **hue** visual feature;
it is difficult to clearly distinguish between all of the different hues
(see @sec-capacity).

Although it does not make the use of hue any more appropriate,
for the purposes of demonstration,
we can solve the colour perception problems in @fig-pie-contrast
by adding a white border to all of the wedges
(see @fig-pie-contrast-solved).

```{r}
#| echo: false
#| label: fig-pie-contrast-solved
#| fig-cap: The hues in this pie chart are perceived more accurately than those in @fig-pie-contrast because each wedge is surrounded by a white border.
ggplot(crimeDistrictTotal) + 
    geom_col(aes(x=total, y="", fill=district), colour="white", linewidth=1) +
    coord_polar() +
    scale_fill_hue() +
    theme(aspect.ratio=1,
          axis.title.y=element_blank(),
          axis.ticks.y=element_blank())
```

## Case study: Divergent colour palettes

@fig-neg-tile shows a heatmap of the points differential for Tier One
nations at each Rugby World Cup.  The data symbols are rectangles, or "tiles".
The teams are mapped to the vertical **positions** of the tiles and
the years are mapped to the horizontal **positions** of the tiles.
These are both effective mappings and we can easily map *back* to
teams or years.
The points differentials are mapped to the **colour** of the tiles,
but in a slightly complicated way.  Positive data values map to a shade
of purple, while negative values map to a shade of brown.
In both cases, we have a constant **hue**, with increasing **chroma** and
decreasing **luminance** as the data values get further away from zero.

The use of two **hues** creates two different groups (positive and negative
values), while the changes in **chroma** and **luminance** convey
changes in absolute magnitude (if only for ordinal comparisons).

```{r}
#| echo: false
#| label: fig-neg-tile
#| fig-cap: A heatmap of the total points differentials for Tier One nations at individual Rugby World Cups.  South Africa was excluded from the 1987 and 1991 tournament due to its apartheid policies.[^gggrid-bite]
edge <- function(data, coords) {
    polygonGrob(c(0, 0,
                  coords$x[data$absent][2] + .05, 
                  coords$x[data$absent][2] + .05, 
                  0, 0, 1, 1),
                c(1, 
                  coords$y[data$absent][1] + .05, 
                  coords$y[data$absent][1] + .05, 
                  coords$y[data$absent][1] - .05, 
                  coords$y[data$absent][1] - .05,
                  0, 0, 1),
                gp=gpar(fill=NA))
}
ggplot(rwcAllyear) +
    geom_tile(aes(x=year, y=team, fill=diff)) +
    scale_fill_continuous_diverging(palette="Purple-Brown", rev=TRUE,
                                    na.value="transparent") +
    scale_x_continuous(breaks=unique(rwcAllyear$year), expand=expansion(0)) +
    scale_y_discrete(expand=expansion(0)) +
    grid_panel(edge, aes(x=year, y=team, absent=absent)) +
    theme(panel.border=element_blank(),
          aspect.ratio=1)
```

[^gggrid-bite]:
  The {gggrid} package [@pkg-gggrid]
  was used to take the "bite" out of the panel
  border in @fig-neg-tile where South Africa was excluded.

> A diverging colour palette is effective because it creates two distinct
  ordinal visual features.

## Summary

**Colour** is really three visual features:  **hue**, **chroma**, 
and **luminance**.

**Hue** is excellent for representing **nominal** data values.

**Chroma** and **luminance** can be used to represent **ordinal**
data values, but they have a very limited capacity.



