
# Colour

@fig-pie-hue shows a pie chart of the number of offenders for different
levels of crime seriousness (between 2011 and 2021).
In this data visualisation, the crime level, 
which is a **qualitative** variable,
is encoded as **colour**, which is appropriate for representing
qualitative values (see @sec-qual-features).  It is easy to identify
different crime levels because it is easy to perceive differences between
the colours of the wedges in the pie chart.

```{r}
#| echo: false
#| label: fig-pie-hue
#| fig-cap: A data visualisation of the number of offenders for different levels of crime seriousness.
ggplot(crimeLevelTotal) +
    geom_col(aes(x=total, y="", fill=level)) +
    coord_polar() +
    scale_fill_npg(guide=guide_legend(reverse=TRUE)) +
    theme(axis.title=element_blank(),
          axis.ticks=element_blank())
```

@fig-pie-ordinal shows another pie chart of the same data, again encoding
the **qualitative** level as **colour**.
It is again easy to differentiate between different crime levels because we
can easily perceive differences between the colours of the wedges.
However, @fig-pie-ordinal is *more* effective than @fig-pie-hue because
we can *also* perceive an ordering of the wedges from more serious crimes,
which are darker wedges, to less serious crimes, which are lighter wedges.

```{r}
#| echo: false
#| label: fig-pie-ordinal
#| fig-cap: A data visualisation of the number of offenders for different levels of crime seriousness.
ggplot(crimeLevelTotal) +
    geom_col(aes(x=total, y="", fill=level), colour="black") +
    coord_polar() +
    scale_fill_ordinal(guide=guide_legend(reverse=TRUE)) +
    theme(axis.title=element_blank(),
          axis.ticks=element_blank())
```

The fact that there can be two data visualisations,
@fig-pie-hue and @fig-pie-ordinal,
with the same mapping,
the **qualitative**
level of crime is encoded as the visual feature **colour**,
but the two data visualisations 
have different effectiveness suggests that there is still
more to learn about how different visual features work.

This section covers some more details about how the **colour** visual
feature works.

## Hue, chroma, and luminance {#sec-colour}

We have so far taken a very simplistic approach to **colour**,
treating it as if it is a single **qualitative** visual feature
(@sec-qual-features).
However, the perception of colour can be divided into 
three separate visual features:  **hue**, **chroma**, and **luminance**
(@fig-hcl).[^tricolour]
Hue describes what we normally refer to as the colour name: reds, oranges,
yellows, greens, blues, and purples.
Chroma describes the colourfulness, from bright to dull,
and luminance describes the lightness, from dark to light.
Our visual system is capable to perceiving changes in hue and changes in
chroma and changes in luminance.

[^tricolour]:
  The three-dimensional experience of colour arises from the fact that there
  are three different sorts of cones in the retina (see @sec-eye),
  each sensitive to a different range of light wavelengths:
  The L cones respond to long wavelengths (reds); the M cones respond
  to medium wavelengths (greens); and the S cones respond to short
  wavelengths (blues).
  The neural connections that combine the messages from the retina are also 
  divided into three pathways: light-dark, which represents the sum of L, M, 
  and S cones;
  red-green, which represents the difference between L and M cones;
  and blue-yellow, which represents the difference between S cones and the
  sum of L and M cones.
  For more information, see @ware2021information.

We have described colour as a **qualitative** visual feature so far, but that 
really strictly only applies to **hue**.  For example, 
@fig-stacked-bar used different hues to represent different ethnic groups.
Hue is excellent for the perception of differences.

```{r}
#| echo: false
#| label: fig-hcl
#| fig-cap: "The perception of colour can be divided into three components: differences in **hue**, the colours of the rainbow; differences in **chroma**, from bright and colourful to dull and grey; and differences in **luminance**, from light to dark."
vp2 <- viewport(width=.9, y=unit(3, "lines"),
                height=unit(1, "npc") - unit(6, "lines"),
                just="bottom")
n <- 7
t <- seq(0, 2*pi, length.out=n+1)[-(n+1)]
hue <- t/pi*180
chroma <- 100*1:5/5
lum <- 100*1:5/6
x <- .2 + .3*cos(t + 2*pi/n)
y <- .5 + .3*sin(t + 2*pi/n)
grid.newpage()
grid.rect(width=unit(1, "npc") - unit(0/96, "in"),
          height=unit(1, "npc") - unit(0/96, "in"),
          gp=gpar(col=NA, fill="grey20"))
pushViewport(viewport(x=.55, width=.8, height=.8, layout=grid.layout(1, 3)))
pushViewport(viewport(layout.pos.col=1),
             vp2)
grid.move.to(.5, .5)
popViewport(2)
pushViewport(viewport(layout.pos.col=2),
             vp2)
grid.line.to(.5, .5, gp=gpar(col="grey40", lwd=2))
popViewport(2)
pushViewport(viewport(layout.pos.col=3),
             vp2)
grid.line.to(.5, .5, gp=gpar(col="grey40", lwd=2))
popViewport(2)
pushViewport(viewport(layout.pos.col=1),
             vp2)
pushViewport(viewport(width=unit(1, "snpc"), height=unit(1, "snpc")))
grid.circle(x, y, r=.1,
            gp=gpar(col="grey20", lwd=10))
grid.circle(x, y, r=.1,
            gp=gpar(col=figbg, fill=hcl(hue, chroma[3], lum[3]), 
                    lwd=c(rep(0, n - 1), 0)))
popViewport()
grid.text("hue", x=.2, y=unit(-2, "lines"), gp=gpar(col=figbg, fontface="bold"))
popViewport(2)
pushViewport(viewport(layout.pos.col=2),
             vp2)
grid.circle(.5, 0:4/4, r=.1,
            gp=gpar(col="grey20", lwd=10))
grid.circle(.5, 0:4/4, r=.1, 
            gp=gpar(col=figbg, fill=hcl(hue[7], chroma, lum[3]),
                    lwd=c(0,0,0,0,0)))
grid.text("chroma", y=unit(-2, "lines"), gp=gpar(col=figbg, fontface="bold"))
popViewport(2)
pushViewport(viewport(layout.pos.col=3),
             vp2)
grid.circle(.5, 0:4/4, r=.1,
            gp=gpar(col="grey20", lwd=10))
grid.circle(.5, 0:4/4, r=.1, 
            gp=gpar(col=figbg, fill=hcl(hue[7], chroma[3], lum), 
                    lwd=c(0,0,0,0,0)))
grid.text("luminance", y=unit(-2, "lines"), gp=gpar(col=figbg, fontface="bold"))
popViewport(2)
```

## Ordinal visual features {#sec-ordinal}

**Chroma** and **luminance** are both **nominal** visual features
in that they can be used to visualise different groups, but they
are also **ordinal** visual features because we can perceive
an ordering from a set of colours that differ in terms of either chroma
or luminance or both.  A set of colours that transitions from darker
to lighter and/or brighter to duller conveys an ordering from larger
to smaller.

This helps to explain why @fig-pie-ordinal is more effective
than @fig-pie-hue.  In @fig-pie-ordinal, the colours differ
mostly in terms of luminance and chroma so we are able to perceive
changes from larger values to smaller values because the
colours transition from darker to lighter.

In this sense, **chroma** and **luminance** are able to convey
more information than **hue** because they can represent **nominal**
differences *and* **ordinal** differences, whereas hue can only
represent nominal differences.  

On the other hand, @fig-lum-cap shows that chroma and luminance are
even more limited than hue in terms of **capacity**.  We can only use
chroma and/or luminance to represent **nominal** data when there
are very few different groups to represent.

```{r}
#| echo: false
#| label: fig-lum-cap
#| fig-height: 3
#| fig-cap: Mapping qualitative values to luminance and chroma works well for representing nominal data values, but only for a very small number of values. The three values on the left are easily distringuishable, but it becomes difficult to clearly distinguish between adjacent colours in the middle and almost impossible for the colours on the right.  This figure can be compared with @fig-colour-cap, which demonstrates the capacity of hue for representing nominal data values.
clpie <- function(n) {
    t <- rev(seq(0, 2*pi, length.out=100*n) + pi/2)
    cols <- hcl.colors(n+1, "Purples")[-(n+1)]
    textcols <- c("white", "black")[rep(1:2, c(floor(n/3), n - floor(n/3)))]
    for (i in 1:n) {
        index <- 1:100 + (i-1)*100
        x <- cos(t[index])
        y <- sin(t[index])
        grid.polygon(c(.5, .5 + .3*x), c(.5, .5 + .3*y), 
                     gp=gpar(col=figbg, lwd=4, fill=cols[i]))
        grid.text(letters[i], .5 + mean(.25*x), .5 + mean(.25*y),
                  gp=gpar(cex=1, col=textcols[i]))
    }
}
pushViewport(viewport(layout=grid.layout(1, 3, respect=TRUE)))
pushViewport(viewport(layout.pos.col=1))
clpie(3)
popViewport()
pushViewport(viewport(layout.pos.col=2))
clpie(7)
popViewport()
pushViewport(viewport(layout.pos.col=3))
clpie(11)
popViewport()
```

## Perception is relative {#sec-colour-relative}

We saw in @sec-perception-relative that perception of **lengths** is 
relative.  For example, smaller differences in length are easier
to see when the lengths themselves are short (see @fig-relative).

There is a similar rule for the perception of **colours**. 
Our perception of luminance and chroma is affected by neighbouring 
colours.[^contrast-effect]
This effect is demonstrated in @fig-contrast-effect:
the same colour when surrounded by a darker background appears lighter
than it does when surrounded by a lighter background.

[^contrast-effect]:
  This relative perception of colours is useful for being able
  to adapt to a wide range of lighting conditions.
  See @ware2021information for a more detailed discussion.

```{r}
#| echo: false
#| label: fig-contrast-effect
#| fig-height: 3
#| fig-cap: The perception of colours is relative. In the top row, the inner rectangles are exactly the same shade of grey on the left and exactly the same shade of purple on the right.  The perception of the inner rectangles is affected by the surrounding colour, so the same shade appears darker against a light background and darker against a light background.  The bottom row draws a line connecting the two inner rectangles to show that they really are the same shade.
pushViewport(viewport(layout=grid.layout(2, 2, widths=2, respect=TRUE)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1),
             viewport(width=.8, height=.8))
grid.rect(x=.25, width=.5, gp=gpar(col=NA, fill="grey20"))
grid.rect(x=.25, width=.25, height=.5, gp=gpar(col=NA, fill="grey60"))
grid.rect(x=.75, width=.5, gp=gpar(col=NA, fill="grey80"))
grid.rect(x=.75, width=.25, height=.5, gp=gpar(col=NA, fill="grey60"))
popViewport(2)
pushViewport(viewport(layout.pos.col=1, layout.pos.row=2),
             viewport(width=.8, height=.8))
grid.rect(x=.25, width=.5, gp=gpar(col=NA, fill="grey20"))
grid.rect(x=.25, width=.25, height=.5, gp=gpar(col=NA, fill="grey60"))
grid.rect(x=.75, width=.5, gp=gpar(col=NA, fill="grey80"))
grid.rect(x=.75, width=.25, height=.5, gp=gpar(col=NA, fill="grey60"))
grid.segments(.25, .5, .75, .5, gp=gpar(col="grey60", lwd=20))
popViewport(2)
purples <- pal_brewer(palette="Purples")(4)[-1]
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1),
             viewport(width=.8, height=.8))
grid.rect(x=.25, width=.5, gp=gpar(col=NA, fill=purples[1]))
grid.rect(x=.25, width=.25, height=.5, gp=gpar(col=NA, fill=purples[2]))
grid.rect(x=.75, width=.5, gp=gpar(col=NA, fill=purples[3]))
grid.rect(x=.75, width=.25, height=.5, gp=gpar(col=NA, fill=purples[2]))
popViewport(2)
pushViewport(viewport(layout.pos.col=2, layout.pos.row=2),
             viewport(width=.8, height=.8))
grid.rect(x=.25, width=.5, gp=gpar(col=NA, fill=purples[1]))
grid.rect(x=.25, width=.25, height=.5, gp=gpar(col=NA, fill=purples[2]))
grid.rect(x=.75, width=.5, gp=gpar(col=NA, fill=purples[3]))
grid.rect(x=.75, width=.25, height=.5, gp=gpar(col=NA, fill=purples[2]))
grid.segments(.25, .5, .75, .5, gp=gpar(col=purples[2], lwd=20))
popViewport(2)
popViewport()
```

This effect also applies to the perception of hue.
@fig-pie-contrast shows a pie chart of the number of offenders in each
police district, with police district mapped to the **colour** of the
wedges.  There are two issues with the perception of the wedge colours:
each wedge is affected by its neighbour so that, for example, some of 
the blue wedges look more green at the boundary with their more blue 
neighbour and more blue at the boundary with their more green neighbour;
furthermore, the colours in the legend are perceived differently from
the corresponding colours in the wedges because the legend colours are
surrounded by a light grey, rather than sharing a boundary with
a neighbouring colour.

```{r}
#| echo: false
#| label: fig-pie-contrast
#| fig-cap: This pie chart demonstrates that the perception of hues is relative. The colours within each wedge appear to have a subtle gradient within the wedge because the perception of the wedge colour is affected differently by the neighbouring wedge colours.
ggplot(crimeDistrictTotal) + 
    geom_col(aes(x=total, y="", fill=district)) +
    coord_polar() +
    scale_fill_hue() +
    theme(aspect.ratio=1,
          axis.title.y=element_blank(),
          axis.ticks.y=element_blank())
```

The pie chart in @fig-pie-contrast is *not* an effective data visualisation
because it encodes qualitative data values with many different groups
as colour hue.  This exceeds the **capacity** of the **hue** visual feature;
it is difficult to clearly distinguish between all of the different hues
(see @sec-capacity).

Although it does not make the use of hue any more appropriate,
for the purposes of demonstration,
we can solve the colour perception problems in @fig-pie-contrast
by adding a white border to all of the wedges
(see @fig-pie-contrast-solved).
Now all of the wedges and all of the squares in the legend are
perceived relative to a surrounding white border.

```{r}
#| echo: false
#| label: fig-pie-contrast-solved
#| fig-cap: The hues in this pie chart are perceived more accurately than those in @fig-pie-contrast because each wedge is surrounded by a white border.
ggplot(crimeDistrictTotal) + 
    geom_col(aes(x=total, y="", fill=district), colour="white", linewidth=1) +
    coord_polar() +
    scale_fill_hue() +
    theme(aspect.ratio=1,
          axis.title.y=element_blank(),
          axis.ticks.y=element_blank())
```

## Colour vision deficiency {#sec-CVD}

Roughly 10% of the population, primarily men, suffer from a form of
colour vision deficiency.[^cvd]
In the most common case, the viewer will have difficulty distinguishing 
between red and green hues.
@fig-pie-hue-cvd simulates how the colours in @fig-pie-hue would appear
to a person with severe deuteranopia (red-green colour blindness).  

[^cvd]:
  Provide a cite for those percentages

```{r}
#| echo: false
#| label: fig-pie-hue-cvd
#| fig-cap: A data visualisation of the number of offenders for different levels of crime seriousness.
ggplot(crimeLevelTotal) +
    geom_col(aes(x=total, y="", fill=level)) +
    coord_polar() +
    scale_fill_manual(values=deutan(pal_npg()(5)),
                      guide=guide_legend(reverse=TRUE)) +
    theme(axis.title=element_blank(),
          axis.ticks=element_blank())
```

In order to avoid confusion for viewers with CVD, 
it may be necessary to vary luminance and/or chroma in addition
to or instead of hue.  For example, @fig-pie-ordinal-cvd 
shows how the shades of purple in @fig-pie-ordinal would appear
to a viewer with severe deuteranopia.
The difference in hue
is of no importance in this case.

```{r}
#| echo: false
#| label: fig-pie-ordinal-cvd
#| fig-cap: A data visualisation of the number of offenders for different levels of crime seriousness.
ggplot(crimeLevelTotal) +
    geom_col(aes(x=total, y="", fill=level), colour="black") +
    coord_polar() +
    scale_fill_manual(values=deutan(pal_brewer(palette="Purples")(5)),
                      guide=guide_legend(reverse=TRUE)) +
    theme(axis.title=element_blank(),
          axis.ticks=element_blank())
```

## Size matters

The perception of colours is also affected by the size of the
coloured region.  In particular, small regions of colour 
may blend together with neighbouring colours.  [^stone]
For example, on the left of @fig-colour-size, the yellow and blue
stripes are the same colours both in the top set if wide bars 
and in the bottom set of narrow bars.
The right side of @fig-colour-size shows that the same colours
may appear different when they are used as fill colours for larger regions,
like bars, compared to when they are used for thin lines.

[^stone]:
    "In Color Perception, Size Matters"
    Stone (2012)
    "In designing colors for digital visualization systems. one of the most
    critical factors is the interaction between size and color appearance."
    Talks about "spreading", which means a sort of blending or averaging
    of colours over small areas *by the visual system*.

```{r}
#| echo: false
#| label: fig-colour-size
#| fig-height: 3
#| fig-cap: The perception of colours varies based on the size of the coloured regions.
grid.newpage()
grid.rect(width=unit(1, "npc") - unit(0/96, "in"),
          height=unit(1, "npc") - unit(0/96, "in"),
          gp=gpar(col=NA, fill="grey20"))
cols <- c("#FFF200", "#00ADEF")
pushViewport(viewport(layout=grid.layout(2, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1),
             viewport(width=.8, height=.6, y=.4))
grid.rect(x=0:8/9, width=1/9, just="left", gp=gpar(col=NA, fill=cols))
popViewport(2)
pushViewport(viewport(layout.pos.col=1, layout.pos.row=2),
             viewport(width=.8, height=.6))
grid.rect(x=0:58/59, width=1/59, just="left", gp=gpar(col=NA, fill=cols))
popViewport(2)
x <- rep(1:5/6, each=2)
y <- c(1, 2, 4, 3, 3, 4, 3, 3, 5, 2)/6
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1),
             viewport(width=.8, height=.8))
grid.rect(x=unit(x, "npc") + unit(c(-5, 0), "mm"), width=unit(5, "mm"),
          y=0, height=y, hjust=0, vjust=0,
          gp=gpar(col=NA, fill=cols))
popViewport(2)
pushViewport(viewport(layout.pos.col=2, layout.pos.row=2),
             viewport(width=.8, height=.8))
grid.lines(x[c(TRUE, FALSE)], y[c(TRUE, FALSE)], gp=gpar(col=cols[1]))
grid.lines(x[c(FALSE, TRUE)], y[c(FALSE, TRUE)], gp=gpar(col=cols[2]))
popViewport()
```

## Colour palettes

When we encode nominal or ordinal data values as colours,
we have to choose a set of colours, or a **colour palette**.
For example, in @fig-pie-hue we have to choose a different colour
to represent each level of crime.

This choice is difficult because there are many different colours 
to choose from and often there is no obvious mapping 
from a data value to a particular
colour.  For example, what specific hue, chroma, and luminance
should we choose to represent a `high` crime level?

There are also competing criteria involved in selecting colours.
For example, although the primary goal is to represent different categories 
with colours that are as different from each
other as possible, as we saw in @sec-CVD, we also 
have to take into account that some viewers may have difficulty
perceiving differences between particular colours.

Fortunately, researchers have produced a number of 
predefined colour palettes that we can make use of.
For example, the colours used in @fig-pie-hue
are based on palettes used in journals published by the
Nature Publishing Group.[^ggsci]

[^ggsci]:
  cite ggsci, colour brewer (and others?)

If the data symbols are arranged in a regular pattern,
like in a bar plot,
then comparisons between neighbours are most important,
so we can order the colours to maximise neighbour differences
rather than maximising all pairwise differences.[^bartolomeo]
@fig-pie-order demonstrates this idea with a bar plot of the crime
rate in different districts for both 2011 and 2021.
This bar plot uses the same set of
hues as @fig-pie-contrast-solved, just in a different order
so that adjacent bars have very different colours.

[^bartolomeo]:
  "Reflections on the Uses and Available Choices of Categorical Colorschemes"
  Bartolomeo et al (2025)

```{r}
#| echo: false
#| label: fig-pie-order
#| fig-cap: The hues in this pie chart are ordered so that adjacent hues are distant from each other.
ggplot(subset(crimeDistrict, year %in% c(2011, 2021))) + 
    geom_col(aes(x=year, y=rate, fill=district), position="dodge",
             colour=figbg) +
    scale_fill_manual(values=pal_hue()(12)[t(matrix(1:12, 
                                                    ncol=2))]) +
    scale_x_continuous(breaks=c(2011, 2021)) +
    scale_y_continuous(expand=expansion(c(0, .05))) +
    theme(axis.title.x=element_blank())
```

A similar idea is to just reuse a smaller set of hues, which means
larger differences between adjacent colours, but repeat
the set.   The
regular arrangement of the data symbols means that we can
still differentiate between repeated colours because of the bar positions
(@fig-pie-repeat).  For example, we can differentiate the pink for 
the Wellington district
from the pink for the Eastern district because Wellington is at the end and
Eastern is in the middle.

```{r}
#| echo: false
#| label: fig-pie-repeat
#| fig-cap: The hues in this pie chart are repeated so that adjacent hues are more distant from each other.
ggplot(subset(crimeDistrict, year %in% c(2011, 2021))) + 
    geom_col(aes(x=year, y=rate, fill=district), position="dodge",
             colour=figbg) +
    scale_fill_manual(values=rep(pal_hue()(6), 2)) +
    scale_x_continuous(breaks=c(2011, 2021)) +
    scale_y_continuous(expand=expansion(c(0, .05))) +
    theme(axis.title.x=element_blank())
```

## Case study: Divergent colour palettes

The Rugby World Cup is a competition that is held every four years, with the
first event taking place in 1987.
We have data on the total number of points that were scored and conceded
in World Cup matches
for Tier One nations (the top 10 rugby-playing nations) at each World Cup.
@tbl-rwc-all-year shows a sample of these data.[^rwc]

[^rwc]:
  Attribution to kaggle data set.

```{r}
#| echo: false
#| message: false
#| label: tbl-rwc-all-year
#| tbl-cap: A table of the total number of points scored and conceded by Tier One nations in Rugby World Cup matches at each World Cup.  Each row represents thetotal points scored by one team in one World Cup.  There are 100 rows in total, with just the first 6 rows shown here.
kable(head(rwcAllyear[,c("team", "year", "scored", "conceded", "diff")]), digits=0)
```

@fig-neg-tile shows a heatmap of the points differential 
(points scored for minus points scored against) for Tier One
nations at each Rugby World Cup.  The data symbols are rectangles, or "tiles".
The teams are encoded as the vertical **positions** of the tiles and
the years are encoded as the horizontal **positions** of the tiles.
These are both effective mappings and we can easily **decode** to
teams or years.
The points differentials are encoded as the **colour** of the tiles,
but in a slightly complicated way.  Positive data values map to a shade
of purple, while negative values map to a shade of brown.
In both cases, we have a constant **hue**, with increasing **chroma** and
decreasing **luminance** as the data values get further away from zero.

The use of two **hues** creates two different groups (positive and negative
values), while the changes in **chroma** and **luminance** convey
changes in absolute magnitude (if only for ordinal comparisons).

```{r}
#| echo: false
#| label: fig-neg-tile
#| fig-cap: A heatmap of the total points differentials for Tier One nations at individual Rugby World Cups.  South Africa was excluded from the 1987 and 1991 tournament due to its apartheid policies.[^gggrid-bite]
edge <- function(data, coords) {
    polygonGrob(c(0, 0,
                  coords$x[data$absent][2] + .05, 
                  coords$x[data$absent][2] + .05, 
                  0, 0, 1, 1),
                c(1, 
                  coords$y[data$absent][1] + .05, 
                  coords$y[data$absent][1] + .05, 
                  coords$y[data$absent][1] - .05, 
                  coords$y[data$absent][1] - .05,
                  0, 0, 1),
                gp=gpar(fill=NA))
}
ggplot(rwcAllyear) +
    geom_tile(aes(x=year, y=team, fill=diff)) +
    scale_fill_continuous_diverging(palette="Purple-Brown", rev=TRUE,
                                    na.value="transparent") +
    scale_x_continuous(breaks=unique(rwcAllyear$year), expand=expansion(0)) +
    scale_y_discrete(expand=expansion(0)) +
    grid_panel(edge, aes(x=year, y=team, absent=absent)) +
    theme(panel.border=element_blank(),
          aspect.ratio=1)
```

[^gggrid-bite]:
  The {gggrid} package [@pkg-gggrid]
  was used to take the "bite" out of the panel
  border in @fig-neg-tile where South Africa was excluded.

> A diverging colour palette is effective because it creates two distinct
  visual features:  a nominal hue to differentiate between two groups and 
  changes in chroma and/or luminance to encode ordinal values within each group.

## Summary

**Colour** is really three visual features:  **hue**, **chroma**, 
and **luminance**.

**Hue** is excellent for representing **nominal** data values.

**Chroma** and **luminance** can be used to represent **ordinal**
data values, but they have a very limited capacity.

We should lean on the work done by experts when it comes to choosing
colour palettes.

