
```{r}
#| echo: false
#| message: false
source("common.R")
```

# Length {#sec-quantitative}

@fig-bar-district shows a bar plot of the number of youth offenders
in each police district in New Zealand in 2021 (@tbl-map-data).
For the reasons outlined in @sec-encoding-visual-features
and @sec-position,
this data visualisation is effective for
decoding the number of offenders in different police districts
and making comparisons between them
because it encodes the police districts as the
vertical **positions** of bars and it encodes the number of offenders as the 
**lengths* of bars.

```{r}
#| echo: false
#| label: fig-bar-district
#| fig-cap: A bar plot of the number of offenders for different police districts, with the districts ordered from North to South.
crimeTemp <- crimeDistrict2021
crimeTemp$district <- factor(crimeTemp$district,
		             levels=rev(c("Northland", "Waitematā",
			              "Auckland City",
			              "Counties/Manukau", "Waikato",
				      "Bay of Plenty",
				      "Eastern", "Central",
				      "Wellington",
				      "Tasman", "Canterbury",
				      "Southern")))
ggplot(crimeTemp) +
    geom_col(aes(x=count, y=district), width=.9) +
    scale_x_continuous(expand=expansion(c(0, .05))) +
    theme(aspect.ratio=1)
```

@fig-bar-district-ordered shows a bar plot of the same data, but with the
bars ordered from longest to shortest.  This data visualisation uses
the same encodings as @fig-bar-district, which means that it is also
effective for comparing the number of offenders between 
police districts.
However, @fig-bar-district-ordered is *more* effective than @fig-bar-district
for some comparisons.  For example, it is easier to compare the Northland
district with the Tasman district in @fig-bar-district-ordered.
The fact that two data visualisations with the same encodings
can have a different effectiveness suggests that there is more to know about
how different visual features work.
This chapter covers some more details about how well we can decode
information when we have encoded data values
as **length**.

```{r}
#| echo: false
#| label: fig-bar-district-ordered
#| fig-cap: A bar plot of the number of offenders for different police districts, with the districts ordered from largest number of offenders to smallest.
crimeDistrict2021$districtOrdered <- 
    with(crimeDistrict2021, reorder(district, count))
ggplot(crimeDistrict2021) +
    geom_col(aes(x=count, y=districtOrdered), width=.9) +
    scale_x_continuous(expand=expansion(c(0, .05))) +
    scale_y_discrete(name="district") +
    theme(aspect.ratio=1)
```

## The distance effect {#sec-distance-effect}

In @sec-accuracy, we saw that **position** and **length** were the most
accurate visual features (@fig-accuracy).  However, this accuracy
decreases if the two visual features that are being compared are
not in close proximity.[^distance-effect]
For example, 
it is more difficult to compare the lengths of the two black bars in the
set of bars at the bottom of @fig-distance-effect than it is to compare
the lengths of the two black bars at the top of @fig-distance-effect.

[^distance-effect]:
    The first demonstration of the "distance effect" is attributed to 
    @cleveland1984graphical.  It has since been replicated several times,
    for example, by @talbot2014four
    and @lu2021modeling.

This helps to explain why @fig-bar-district-ordered is more effective
than @fig-bar-district for some comparisons (e.g., Northland versus Tasman).
One reason for ordering the bars in a bar plot from longest to shortest 
is because
the ordering places bars that we want to compare closer 
together.


```{r}
#| echo: false
#| label: fig-distance-effect
#| fig-cap: Two bars are easier to compare if they are close to each other (top row).  If there is a distance between the bars and/or there are distractors (other bars) in between then comparisons are less accurate (bottom row).  On both the top and bottom row, the black bar on the left is slightly higher than the black bar on the right.
pushViewport(viewport(width=.6, layout=grid.layout(2, 1)))
pushViewport(viewport(layout.pos.row=1),
             viewport(width=.8, height=.8))
grid.segments(0, .2, 1, .2, gp=gpar(col="grey"))
grid.rect(.425, .2, .1, .6, just="bottom", gp=gpar(fill="black"))
grid.rect(.575, .2, .1, .55, just="bottom", gp=gpar(fill="black"))
popViewport(2)
pushViewport(viewport(layout.pos.row=2),
             viewport(width=.8, height=.8))
grid.segments(0, .2, 1, .2, gp=gpar(col="grey"))
grid.rect(.125, .2, .1, .6, just="bottom", gp=gpar(fill="black"))
grid.rect(seq(.275, .725, .150), .2, .1, runif(4, .2, .8), 
          just="bottom", gp=gpar(col=NA, fill="grey"))
grid.rect(.875, .2, .1, .55, just="bottom", gp=gpar(fill="black"))
popViewport(2)
```

## Unaligned length {#sec-unaligned}

@fig-unaligned-length shows that it is harder to accurately compare
the **lengths** of two bars if they are **unaligned** and do not share a 
common baseline (similar to the accuracy of comparing unaligned **positions**;
@sec-unaligned-pos).

```{r}
#| echo: false
#| label: fig-unaligned-length
#| fig-cap: It is easier to compare the lengths of the two bars on the left because they are **aligned**;  they share a common base.  The two bars on the right are harder to compare because they are **unaligned**.
pushViewport(viewport(x=0, width=.5, just="left"))
pushViewport(viewport(width=unit(.8, "snpc")))
grid.segments(.2, .1, .2, .9, gp=gpar(col="grey"))
grid.rect(.2, c(.3, .7), c(.5, .6), .2, just="left", gp=gpar(fill="black"))
popViewport(2)
pushViewport(viewport(x=.5, width=.5, just="left"))
pushViewport(viewport(x=.4, width=unit(.8, "snpc")))
grid.segments(.2, .1, .2, .5, gp=gpar(col="grey"))
grid.segments(.4, .5, .4, .9, gp=gpar(col="grey"))
grid.rect(c(.2, .4), c(.3, .7), c(.5, .6), .2, just="left",
          gp=gpar(fill="black"))
popViewport(2)
```

@fig-stacked-bar shows a visualisation of the data on 
the number of offenders in different ethnic groups (@tbl-crime-group-total).
The bar plot of these data that we saw in @fig-bar encoded ethnic groups
as the **positions** of the bars so that the bars were arranged
vertically with a common left-hand edge.
In @fig-stacked-bar, the ethnic groups are 
encoded as different **colours** and the bars are "stacked" 
horizontally.[^stacking-as-position]

[^stacking-as-position]:
  In a stacked bar plot, groups are encoded as the
  positions of the bars and counts are encoded as the lengths
  of the bars, but the positions are in the same
  dimension as the lengths (in this case, both position and length
  are horizontal).  In a side-by-side bar plot,
  the position of the bars is in the opposite dimension to the
  lengths of the bars.

Although the number of offenders has been encoded as the same
visual feature in both @fig-stacked-bar and @fig-bar---in both cases,
the number of offenders is encoded as the **lengths**
of bars---comparing those lengths is much harder in @fig-stacked-bar
because the lengths are **unaligned**.
One reason why stacked bar plots are less effective for some
comparisons is because we are forced to compare the unaligned lengths 
of the bars.

```{r}
#| echo: false
#| label: fig-stacked-bar
#| fig-cap: A stacked bar plot of the number of offenders in each ethnic group.
gg <- ggplot(crimeGroup2021) +
    geom_col(aes(count, y="", fill=group), position="stack", width=.5) +
    scale_x_continuous(expand=expansion(0)) +
    scale_y_discrete(name=NULL, expand=expansion(0)) +
    theme(axis.ticks.y=element_blank())
pushViewport(viewport(width=.8, height=.6))
print(gg, newpage=FALSE)
popViewport()
```

## Case study:  Stacked versus side-by-side bar plots {#sec-stacked-vs-side}

Stacking bars, which creates unaligned lengths, is something that
arises when we use a bar plot to show a quantitative variable broken
down by more than one qualitative variable.
For example, @tbl-crime-group-year shows data on the number of offenders
in different ethnic groups for individual years (from 2011 to 2021).
This is a more detailed version of the data from @tbl-crime-group-total.

```{r}
#| echo: false
#| message: false
#| label: tbl-crime-group-year
#| tbl-cap: A table of the number of offenders per year aged 14 to 16 for different ethnic groups from 2011 to 2021.  There are 44 rows of data, but just the first 6 rows are shown here.
crimeGroupTable <- subset(crimeGroup, select=c("group", "count", "year"))
kable(head(crimeGroupTable), row.names=FALSE)
```

@fig-stacked-bar-year shows the number of offenders (quantitative)
for each ethnic group (qualitative) and
for each year from 2011 to 2021 (treated here as
qualitative, though ordinal).
In this data visualisation, the number of offenders is encoded as
the lengths of the bars, the years are encoded to the horizontal
**position** of the bars, the ethnic groups are encoded to 
the colours of the bars, and the bars are stacked vertically.

The problem that we face in this situation is that there are 
four bars to show for each year. 
If we simply encode the year as the horizontal position of each bar,
the bars will overlap and we will have great difficulty 
decoding data values from the bars (@sec-pos-scarce).

Stacking the bars is one solution, but 
as we have just seen, this makes some comparisons more difficult.
For example, we can easily see that the number of offenders in the 
unknown group increased after 2016, but it is much harder to see what
has happened to the European/Other group since 2016.

```{r}
#| echo: false
#| label: fig-stacked-bar-year
#| fig-cap: A stacked bar plot of the number of offenders in each ethnic group for each year from 2011 to 2021. 
ggplot(crimeGroup) +
    geom_col(aes(year, count, fill=group), position="stack") +
    scale_x_continuous(breaks=seq(2012, 2020, 2)) +    
    scale_y_continuous(expand=expansion(c(0, .05)))
```

An alternative visualisation is shown in @fig-side-bar-year.
This has placed the four ethnic groups side-by-side for each year.
It is now easier to see the changes over time for the European/Other 
group because all bars are now aligned.  Unfortunately, as we saw in
@sec-distance-effect, some comparisons are more difficult
because the bars are now further apart
and have distractors in between.

```{r}
#| echo: false
#| label: fig-side-bar-year
#| fig-cap: A side-by-side bar plot of the number of offenders in each ethnic group for each year from 2011 to 2021. 
ggplot(crimeGroup) +
    geom_col(aes(year, count, fill=group), position="dodge") +
    scale_x_continuous(breaks=seq(2012, 2020, 2)) +    
    scale_y_continuous(expand=expansion(c(0, .05)))
```

Yet another variation is shown in @fig-side-bar-group.
This time we have placed all years side-by-side for each ethnic group.
This provides the clearest view of the changes over time for each ethnic
group because the relevant bars are close together and aligned.
However, there are other comparisons that are considerably more difficult 
because of the distance between the relevant bars, for example,
comparing between ethnicities for a specific year.

```{r}
#| echo: false
#| label: fig-side-bar-group
#| fig-keep: last
#| fig-cap: A side-by-side bar plot of the number of offenders in each ethnic group for each year from 2011 to 2021. 
ggplot(crimeGroup) +
    geom_col(aes(group, count, fill=group, group=year), position="dodge",
             colour=figbg) +
    scale_y_continuous(expand=expansion(c(0, .05))) +
    theme(axis.title=element_blank(),
          legend.position="none")
grid.force()
barPath <- grid.grep("panel::geom_rect", grep=TRUE, viewports=TRUE)
bars <- grid.get(barPath)
downViewport(attr(barPath, "vpPath"))
index <- which(crimeGroup$year == 2011 & crimeGroup$group == "Unknown")
grid.text(" 2011", 
          bars$x[index] + bars$width[index]/2, bars$y[index], 
          just="left", rot=90,
          gp=gpar(fontsize=8))
index <- which(crimeGroup$year == 2021 & crimeGroup$group == "Unknown")
grid.text(" 2021", 
          bars$x[index] + bars$width[index]/2, bars$y[index], 
          just="left", rot=90,
          gp=gpar(fontsize=8))
```

There is no single best visualisation in general for this situation because
different visualisations each have their strengths and weaknesses.
If a particular sort of comparison is to be emphasised, then one 
data visualisation might stand out, but there is also the option of
providing more than one data visualisation in order to cover 
more comparisons.

## Perception is relative {#sec-perception-relative}

@fig-relative shows that the perception of differences between the lengths
of bars is relative;
it is easier to detect smaller differences when the visual features
that we are comparing are smaller.[^Weber]
It is easier to see that the pair of bars on the left have different
lengths and it is harder to see that the pair of bars on the right have 
different lengths.  All three pairs of bars have different lengths,
but the absolute size of the difference is the same, so the size
of the difference relative to the bar lengths gets smaller as the bars
get longer.

[^Weber]:
  The idea of perception of differences being relative is a very old result 
  from Psychology 
  known as Weber's Law [@weber1834tastsinn; @boring1942sensation].

```{r}
#| echo: false
#| label: fig-relative
#| fig-cap: A demonstration that perception of differences is relative. The pairs of bars all differ by the same absolute amount, but the difference is more easily perceived when comparing the shorter bars.  In the shorter bars, the absolute difference between the bars is a larger difference relative to the length of the bars.
colWidth <- .2
offset <- c(0, .05)
pushViewport(viewport(height=.8, layout=grid.layout(1, 3)))
pushViewport(viewport(layout.pos.col=1), 
             viewport(width=.8, height=.8))
grid.rect(c(.35, .65), offset, width=colWidth, height=c(.1, .15), 
          just="bottom", gp=gpar(fill="black"))
popViewport(2)
pushViewport(viewport(layout.pos.col=2), 
             viewport(width=.8, height=.8))
grid.rect(c(.35, .65), offset, width=colWidth, height=c(.5, .55), 
          just="bottom", gp=gpar(fill="black"))
popViewport(2)
pushViewport(viewport(layout.pos.col=3), 
             viewport(width=.8, height=.8))
grid.rect(c(.35, .65), offset, width=colWidth, height=c(.8, .85), 
          just="bottom", gp=gpar(fill="black"))
popViewport(2)
```

@fig-relative-ref shows that we can make it easier to judge
the differences by providing reference marks.  Each pair
of black bars in @fig-relative-ref
is the same and each pair of bars differs in length
by the same small amount.  The difference between the bars in each 
pair is difficult to perceive because the difference is small relative
to the lengths of the bars.  The
reference marks in the middle and right-hand pair of bars
provide shorter lengths to compare, so we are able
to more easily perceive the small differences.  For example, for the
pair of bars on the right, we can compare the lengths of the short white
gaps rather than comparing the longer black bars.

```{r}
#| echo: false
#| label: fig-relative-ref
#| fig-cap: A demonstration that reference marks can make it easier to perceive small differences.  Each pair of black bars is the same and each pair differs by a small amount.  In the middle pair, grid lines make it easier to compare the distances from the ends of the bars to the nearest grid line and, in the right-hand pair, a border added to each bar, where the two borders are the same length, makes it easier to compare the white gaps rather than the bars themselves.
pushViewport(viewport(height=.8, layout=grid.layout(1, 3)))
pushViewport(viewport(layout.pos.col=1), 
             viewport(width=.8, height=.8))
grid.rect(c(.35, .65), offset, width=colWidth, height=c(.8, .85), 
          just="bottom", gp=gpar(fill="black"))
popViewport(2)
pushViewport(viewport(layout.pos.col=2), 
             viewport(width=.8, height=.8))
grid.segments(.2, seq(0, 1, .1), .8, seq(0, 1, .1), gp=gpar(col="black"))
grid.rect(c(.35, .65), offset, width=colWidth, height=c(.8, .85), 
          just="bottom", gp=gpar(fill="black"))
popViewport(2)
pushViewport(viewport(layout.pos.col=3), 
             viewport(width=.8, height=.8))
grid.rect(c(.35, .65), offset, width=colWidth, height=.95, 
          just="bottom")
grid.rect(c(.35, .65), offset, width=colWidth, height=c(.8, .85), 
          just="bottom", gp=gpar(fill="black"))
popViewport(2)
```

@fig-bar-district-grid shows a bar plot of the number of offenders in each 
police district (@fig-bar-district) with some grid lines added.
Although it is difficult to compare the bars for Northland and Tasman
because they are quite far apart and there are distractors in between
(@sec-distance-effect), the addition of the grid lines makes it easier
to perform the comparison because we are able to compare the distances
from the ends of the bars to the nearest grid line.  Those distances are
much shorter than the bars so it is easier to perceive that those two
bars are almost identical.

One reason why grid lines are effective is because they allow
comparisons of smaller lengths or distances, which means that
it is easier to perceive smaller changes.

```{r}
#| echo: false
#| label: fig-bar-district-grid
#| fig-cap: A bar plot of the number of offenders for different police districts, with the districts ordered from North to South.  This is very similar to @fig-bar-district, just with some grid lines added.
crimeTemp <- crimeDistrict2021
crimeTemp$district <- factor(crimeTemp$district,
		             levels=rev(c("Northland", "Waitematā",
			              "Auckland City",
			              "Counties/Manukau", "Waikato",
				      "Bay of Plenty",
				      "Eastern", "Central",
				      "Wellington",
				      "Tasman", "Canterbury",
				      "Southern")))
ggplot(crimeTemp) +
    geom_col(aes(x=count, y=district), width=.9) +
    scale_x_continuous(expand=expansion(c(0, .05))) +
    theme(panel.grid.major.x=element_line(colour="black", linewidth=.1),
          aspect.ratio=1)
```

## Summary

::: {.content-visible when-format=html}

::: {.oldsummary .callout-note title='Précis *(click to expand/contract)*' icon=false collapse=true}
  
{{< include intro-summary.md >}}

{{< include visual-model-summary.md >}}

{{< include mapping-visual-features-summary.md >}}

{{< include position-summary.md >}}

:::

:::

{{< include quantitative-summary.md >}}


---

```{=latex}
\theendnotes
```
