
```{r}
#| echo: false
#| message: false
source("common.R")
```

# Review {#sec-review}

Add review at the end that reinforces the threads like visual feature, 
visual shape, visual object and encodings

If we encode X using Y encoding what can we decode?

Possible encoding end points:

* visual feature
* visual summary
* emergent feature
* visual shape 
* visual object
* text

Possible decodings from encoding end points:

* qual/quant data value
* data summary
* higher level information/knowledge/wisdom

Do two versions (at least):

* Set of questions for when you already have a plot and want to understand why
  it works or not

* Another set for if you want to create a data vis

Basic questions:

* What are you asking the visual system to do?
* Is the visual system good at that task? 
* Are you asking the viewer to produce visual summaries
  when it would be easier to produce data summaries?

Present process by which a data vis can be broken down.
What data symbols are being used?  (simple or complex)
What encodings are being used? 
What do we want to decode? (data or summaries)
If encoding back to raw data, are the encodings accurate?
What can be encoded back to?
Are encodings congruent with data?

...

List of questions to evaluate a data vis (maybe nest these a bit):

* What do you want the viewer to be able to decode?
* Do the encodings support that goal?

* What are the data variables
* What visual features are they encoded to
* What type of data
* Can the visual features represent the data
* Is there the capacity to represent all categories
* How accurate are the decodings
* Is there congruence or dissonance in the encoding
* Encoding data or data summary
* Do visual features interact
* Are there emergent features
* Is the data symbol a visual shape
* Is that appropriate
* A visual object
* Is that appropriate
* Learned encodings somehow (distinct from implicit encodings)

* Do all visual differences correspond to differences in the data?
* Are all differences in the data represented by visual differences? 

* Does each data value encode to its own data symbol
  If not, individual data values may be difficult to decode 
* Do multiple data values encode to a single data symbol
  If not, data summaries may be difficult to decode 


And so on ...

## Examples

* Use heatmap as case study?

* do the side-by-side pie chart idea as an example of reasoning your
  way to a new visualisation.

* Move the discussion/revisit of tables from text section to here ?

* Another possible application of data vis knowledge:
  Analyse one of the weird boxplot variations that look a bit like a ziggurat?
  
  This is letter-value plot:

  + "Letter-Value Plots: Boxplots for Large Data"
    Hofmann et al (2015)

  + Nice explanation of calculations ...

    24  Beyond the Boxplot: Tukey’s Letter-Value Summaries – Exploratory Data Analysis in R https://share.google/M2qXCsIxsSsqM7jFQ 

  + Labelling is an example of a text encoding that has no learned decoding
    For most people
    So the text is just a symbol like a pch
    You get difference but that is all

  + {lvplot}

    This line suggests that width is calculated to be proportional
    Or at least can be ...

    https://github.com/hadley/lvplot/blob/51ad401be5e017d37883ec5d149bc1818246b6f5/R/draw.r#L43

## Case study: Lines of best fit

@fig-lm-smooth shows a scatter plot of the winning margin
versus the winning score for Tier One nations at Rugby World Cup matches.
The hemisphere of the losing team is encoded as the colour of the data
points and a trend line has been added for each hemisphere.

The trend lines are encodings of data summaries, least-squares
regressions, to line data symbols.  The data summaries consist
of intercept and slope values, which are encoded as the vertical positions
and the angles of the lines, respectively (@fig-stat-vis-decode).

The data symbol is very simple and decoding the data summaries is
also simple because it just involves decoding angle and vertical position.
For example, we can see that, in general, the winning margin is larger
when the winning score is larger, but 
the trend lines suggest that the increase in the winning margin
is slower for southern-hemisphere teams (they fight harder in defeat).

```{r}
#| echo: false
#| label: fig-lm-smooth
#| fig-cap: A scatter plot of the winning margin versus the winning score for matches between Tier One nations at Rugby World Cups.  There is a light shading of the lower-right diagonal to indicate that the winning margin cannot exceed the winning score.  In addition, points on the diagonal represent matches where the losing team failed to score any points.  The straight lines represent a least-squares regression model fit.
#| warning: false
rwcWinner <- do.call(rbind, lapply(split(rwcAll, rwcAll$match), 
                                   function(x) {
                                       if (all(x$scored == x$conceded)) {
                                           winner <- 1
                                           loser <- 2
                                       } else {
                                           winner <- x$scored > x$conceded
                                           loser <- !winner
                                       }
                                       x$oppSphere <- x$hemisphere[loser]
                                       x[winner, ]
                                   }))
ggplot(rwcWinner, aes(scored, scored - conceded, colour=oppSphere)) +
    annotate("polygon", x=c(0, Inf, Inf), y=c(0, Inf, 0), fill="grey97") +
    geom_smooth(method="lm", se=FALSE) +
    geom_point(alpha=.2) +
    scale_x_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning score") +
    scale_y_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning margin") +
    scale_colour_discrete(guide=guide_legend(title="opposition\nhemisphere")) +
    coord_cartesian(clip="off") +
    theme(aspect.ratio=1)
```

@fig-loess-smooth shows the same data as @fig-lm-smooth, but this time
there are curved trend lines added rather than straight trend lines.[^loess]
The curves are encodings of data summaries again, but this time 
loess model fits.  This means that the encoding involves a series
of predicted values encoded as horizontal and vertical positions for
the curve.  In other words, the data summaries are encoded as visual features
that produce a visual shape

[^loess]:
    These are *loess* model fits (locally-estimated scatterplot smoothing).

```{r}
#| echo: false
#| label: fig-loess-smooth
#| fig-cap: Scatter plot
#| warning: false
ggplot(rwcWinner, aes(scored, scored - conceded, colour=oppSphere)) +
    annotate("polygon", x=c(0, Inf, Inf), y=c(0, Inf, 0), fill="white") +
    geom_smooth(method="loess", se=FALSE, span=1) +
    geom_point(alpha=.2) +
    scale_x_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning score") +
    scale_y_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning margin") +
    scale_colour_discrete(guide=guide_legend(title="opposition\nhemisphere")) +
    coord_cartesian(clip="off") +
    theme(aspect.ratio=1)
```

Nice example of a complex calculation, the results of which are what gets
visualised.  Also example of performing the calculation numerically rather
than asking the viewer to perform it visually.  This was in summaries.qmd,
but it cannot be there because it requires scatter plots, which come in
the next section.
Add this after mention of visualising models?

Add footnote to Hadley's more complex model visualisation paper
https://vita.had.co.nz/papers/model-vis.pdf
https://dl.acm.org/doi/10.1002/sam.11271

model fits like lm straight line and ref Hadley's data-space vs model-space
"data models" to mirror "data statistics"
(revisit box plots as VERY simple models? also talk about failing
assumptions of these models, like unimodal for boxplot)

footnote cite to Tukey for visualising phenomena rather than raw data?
"Data-Based Graphics: Visual Display in the Decades to Come"
Tukey (1990)

+ "Beware of Validation by Eye:
   Visual Validation of Linear Trends in Scatterplots"
  Braun et al (2024)
  People are biased towards orthogonal distance to line rather than
  vertical distance.

loess curves
Another example of data -> summary -> features -> shapes -> summaries

## Case study: 2D density plots

Stat summary visualised 

Compare with visual summary of density of points

Different pros for each approach

refer back to @sec-pos-scarce (overplotting)
and mention that shapes are one solution,
with all of the drawbacks of not being able to decode individual
data values.
Footnote surjective/injective?

## Case study: kiviat plots or radar plots or spider plots

  (David Peebles (2008))

## Case study:  Letter-value plots

@fig-lv-default
shows a letter-value plot of 
taxi time for over 450,000 flights within the United States,
broken down by airline.[^lvplot]

[^lvplot]:
    This plot reproduces Figure 2 from @letter-value. This is the default
    letter-value plot produced by the R package {lvplot} [@lvplot].

We will use the information in this book to analyse the data visualisation
in @fig-lv-default.  We will break @fig-lv-default down into 
the encodings that are being used and the effectiveness of those encodings.

```{r}
#| echo: false
#| label: fig-lv-default
#| fig-height: 6
#| fig-cap: A letter-value plot
ggplot(ontime, aes(UniqueCarrier, TaxiOut + TaxiIn)) +
  geom_lv(aes(fill = after_stat(LV))) +
  scale_fill_lv() +
  scale_y_sqrt(breaks = c(10, 50, 100, 150, 200, 250)) +
  xlab("Carrier") +
  ylab("Taxi Time (in minutes)") +
  theme_bw()
```

The purpose of a letter-value plot is to visualise the distribution
of a set of data values.
The information that we want to decode is about the distribution
of the data values:  the average value, the spread of the values,
the skewness, and so on.

The main motivation for the letter-value plot design is that a normal
boxplot produces *way* too many outliers for such a large amount of data.
This is shown in @fig-box-plot.
Furthermore, alternatives to the boxplot, like violin plots,
that fix the problem of too many outliers, are reliant on tuning parameters,
like kernels and bandwidths (@sec-density).

```{r}
#| echo: false
#| warning: false
#| label: fig-box-plot
#| fig-height: 6
#| fig-cap: A box plot
ggplot(ontime, aes(UniqueCarrier, TaxiOut + TaxiIn)) +
  geom_boxplot() +
  scale_y_sqrt(breaks = c(10, 50, 100, 150, 200, 250)) +
  xlab("Carrier") +
  ylab("Taxi Time (in minutes)") +
  theme_bw()
```

The letter-value plot aims to fix the problem of outliers, while at the 
same time, like
the original boxplot, remain free
of tuning parameters.
This is achieved by displaying as many letter values as are supported
by the amount of data, with only individual data values beyond
those limits shown as outliers.

The letter-value plot is an example of a novel data visualisation, so
this provides a good opportunity to analyse whether a letter-value plot
is an effective encoding of information.[^strawman]

[^strawman]:
    In order to incorporate as many points as possible,
    this analysis just focuses on the default {lvplot} settings that
    are used in @fig-lv-default
    and even sinks to the level of critiquing the specific data set
    (and non-linear scales) that are used in this specific plot. 
    It is a very unfair attack on letter-value plots in
    general and the {lvplot} package in particular.

### Letter values

A boxplot is based on the median and the upper and lower quartiles of
a data set.  These statistics are just the first two levels of 
a more general concept of *letter values* introduced in 
"Exploratory Data Analysis" by @tukey1977eda.

Letter values are order statistics, where the $i$th letter value
approximates the quantile corresponding to a tail area of $2^{−i}$.
The first letter value is the median, which splits the data into 
halves ($2^{-1}$), the second letter values are the upper and lower
quartiles, which leave a quarter of the data in each tail ($2^{-2}$),
and so on.  The amount of data between letter value $i$ and
letter value $(i+1)$ is $2^{-(i+1)}$.  For example,
the amount of data between the upper quartile (letter value 2) 
and letter value 3 is an eighth ($2^{-3}$).

The letter values are named M for median, F for fourths, E for eighths,
then, following the alphabet, D for sixteenths, C for thirty-seconds, down to
A for one-hundred and twenth-eighths, then starting again at the end of
the alphabet, in reverse alphabetic
order, Z for two-hundred and fifty-sixths, Y for five-hundred and twelfths,
and so on.

### Data symbols

There are two types of data symbol in a letter-value plot:
complex data symbols consisting of multiple rectangles, which 
we will call *letter-value* symbols, to
represent the distribution of the data values; and
very simple data points, which represent the outliers (if any).

### Encoding visual features

In @fig-lv-default, the airline (Carrier), qualitative data values,
are encoded as the horizontal **position** of the letter-value symbols.
The encoding of airline as the position of data symbols
is very effective;  we can easily differentiate between the
airlines.

The other data values, the taxi times, are encoded to the
letter-value data symbols within the plot.
However, the letter-value plot encodes **data summaries**---letter 
values---rather 
than raw data values.[^letter-value]
Consequently, all that we can expect to decode from the plot is information
about the letter values, such as the location of specific letter values,
plus possibly data summaries of letter values, such as the
spread of letter values.

[^letter-value]:
    Letter values are an interesting data summary in the sense that
    each letter value corresponds to a raw data value.
    In other words, letter values are a subset of the original raw data
    values, chosen to represent boundaries beyond which a certain
    proportion of the raw data values lie.

    This is not entirely true because a letter value may fall between
    two raw data values.

Each letter value is encoded as the vertical **position** of the top
(or bottom) of a rectangle within the letter-value symbol.
The median (letter value 1) is marked as a single white line rather than
a rectangle.  This is effective for decoding individual letter values
for an airline.  For example, we can see that one of the letter values
for the AA airline is just over 50.

**TODO**
Explain that a variable number of letter values are used for
each airline - only letter values that can be estimated from
the data.

The *depth* of each letter value is encoded as the width (**length**)
of a rectangle within the letter-value symbol.
The depth is how many data values lie beyond that letter value;
for example, the depth of the median is half the total number of 
data values.
This is also an encoding of a data summary.
However, the actual encoding is only based on the *order* of the
depth of the letter values,
with regular decrements 
from widest at the median (letter value 1) to narrowest at the final
letter value (e.g., letter value 13 for airline AA).
In effect, we are encoding a summary of a data summary.
Using **length** for the encoding is effective, but we can only decode
ordinal information about the depth of the letter values
(because that is all that we encoded).

The *data density*---the number of data values between adjacent letter 
values---is encoded as the **colour** of a rectangle.
The **luminance** of the rectangle varies from darkest for the highest density
(approximately one quarter of the data values lie between the median and
the upper quartile) to lightest for lowest density
(approximately one ten-thousandth of the data values between the most
extreme letter values).
In addition, the **hue**/**chroma** of a rectangle encodes a categorisation 
of the letter values (yet another data summary);
every fourth letter value is orange, while all the other letter values
are grey.

The luminance encoding of density only allows us to decode ordinal information
about the data density between letter values.  
Like the widths of the rectangles,
the shades of the rectangles steadily
decrease upwards from the median and downwards from the median.
We can also identify rectangles that represent the same letter value for
different airlines by finding rectangles with the same shade in the
letter-value symbols for two airlines.
This is less effective due to the relatively limited capacity of
luminance encoding (@sec-ordinal).
The different shades of grey are much more difficult to identify than,
for example, the different horizontal locations for different airlines.

The chroma encoding of every fourth letter value is very effective for
decoding between those letter 
values and all of the others, which is essentually an encoding of two groups
(orange versus grey).  
It is also very effective for identifying the same
letter value across different airlines.
The difference in chroma between the oranges
and the greys makes the oranges easy to identify
and there are only three different
luminances for the orange rectangles, so they are more
easily differentiated.

The outliers in a letter-value plot are encoded differently.
The individual taxi times for outliers are encoded as the
vertical position of data points, with the airline encoded as the
horizontal position, as for the letter-value symbols.
These allow for very simple and effective decodings.
It is even possible to rapidly decode numbers of outliers 
thanks to the small number of data symbols involved
(@sec-visual-summaries).

@tbl-lv-encodings summarises the encodings that are involved in a 
letter-value plot.
There are several effective encodings for both quantitative and qualitative
data values, but the encoding of data density and the encoding of
letter-value depth both lose some information.
One overall point is that this is quite a large
number of encodings for a single data visualisation.  This means that
the encodings in @fig-lv-default produce a relatively complex data
visualisation.  There are many visual features changing, which
makes each individual encoding in @fig-lv-default more
difficult to decode (@sec-differences).  Compare this with
@fig-box-plot which, for all its faults, is a much
simpler data visualisation because it involves fewer changes in visual features.

::: {#tbl-lv-encodings}

| Value | Raw/Summary | Type | Encoding |
| :----- | :----------- | :---- | :-------- |
| airline | raw data value | qualitative | position (horiz) |
| taxi outlier | raw data value | quantitative | position (vert) |
| taxi letter value | summary of taxi | quantitative | position (vert) |
| letter value depth | summary of letter value | qualitative | length (width) |
| taxi density | summary of taxi | quantitative | colour (luminance) |
| fourth letter value | summary of letter value | qualitative | colour (chroma) |

The encodings involved in a letter-value plot.
:::

### Combining encodings

There is one obvious interaction between encodings in @fig-lv-default.
The width of the rectangles (letter-value depth) and the height of the
rectangles (difference between adjacent letter values) together produce
the emergent feature of **area** for the rectangles.

Unfortunately, this area does not correspond to any data value or summary of 
the data.  Letter value depth multiplied by difference between letter values
does not give a meaningful value.
This makes @fig-lv-default confusing because the 
rectangles areas do not decode to anything useful 
(@sec-danger-combine).[^unfair]

[^unfair]:
    This criticism is only relevant to the default letter-value plot.
    A variation of the letter-value plot design involves a different encoding
    of width so that the area of the rectangle reflects the 
    proportion of the data between the letter values.

Even worse, the **implicit decoding** of the areas of the 
rectangles in the letter-value data symbols
suggests that, for example, the (darkest) rectangle just above the mean
is about five times the size of the 
highest (lightest) 
rectangle for the "AA" airline.
However, the proportion of the data that lies between the median 
and the next letter value is actually over sixteen thousand times
more than the proportion of the data that lies between the top two letter
values.
In other words, the area of the rectangles is actually *misleading*.

There is also a **redundant encoding** in @fig-lv-default (@sec-redundant).
Because the depth of a letter value is related to the 
order of the letter value,
both the luminance of each rectangle in the letter-value symbol
and the width of the rectangle encode essentially the same information.
This has the benefit of providing more than one way to decode 
this information.
However, given that we have already identified that there are
a large number of encodings in @fig-lv-default, this may be a luxury that
we can ill afford.

### Non-linear scales

An extra complication with the encoding of vertical position
in @fig-lv-default is that the vertical scale is non-linear;
it is actually the *square root* of the letter values that are
encoded as the vertical position of the tops (or bottoms) of the
rectangles in the letter-value data symbols.
This makes the decoding of letter values and the comparison 
of letter values much more difficult because the same physical difference
does not decode to the same difference between letter values.
For example, for the AA airline, although most of the letter values
appear to be similar distances apart, in terms of taxi data values, 
the letter values are further apart for more extreme letter values
(@sec-non-linear-pos).

We have already established that the areas of the rectangles 
in the letter-value data symbols provide no useful information and
are potentially misleading.
The non-linear scale only makes that potentially 
misleading information even more
misleading.

### Visual shapes

The letter-value data symbols encode multiple letter values within
a single data symbol, which produces a **visual shape**
(@sec-visual-shapes).
We have a data symbol that is widest at the median of the data values
and narrows towards either extreme.

This allows us to decode useful data summaries about the distribution of 
the taxi times.
For example, the fact that all of the letter-value data symbols
narrow more rapidly below the mean than they do above the mean
suggests a skewed distribution of taxi times for all airlines.

It is also possible to identify airlines that have similar 
distributions based on similar visual shapes.
These are examples of effective decodings of data summaries from the
visual shapes produced by the letter-value data symbols.

One caveat is that the widths of the visual shapes are potentially
misleading because they do not narrow anywhere near as fast as
the distribution of taxi times.
In other words, the shape of the letter-value data symbols
does not provide accurate decoding of quantitative
features of the distribution of taxi times, only broad qualitative 
features.
For example, we can decode *whether* the distribution is symmetric or skewed,
but we cannot decode *how* skewed the distribution is.

### Encoding encodings

We have already discussed that the y-axis scale is non-linear.
The square root of taxi times, rather then the raw taxi times, are encoded 
as the vertical position of the tops (or bottoms) of rectangles.
This creates the usual problems for encoding a non-linear scale
as tick marks and text labels.
For example, the square-root taxi time of 
10 is encoded as a position about two-thirds
of the way up the y-axis, but this is encoded as the text "100".

On one hand, this is a valuable encoding of the value 10 because it
saves us the mental effort of having to reverse the square root transformation.
If we had encoded this position as the text "10", then we would be required
to perform the mental calculation to decode the 
taxi time 100.  

On the other hand, encoding the square root taxi times as text that 
corresponds to the taxi times causes problems because we cannot 
linearly interpolate between the text values.
Yes, we can correctly decode the precise taxi time 100 from the
text "100", and 50 from "50", but it is very difficult to decode
any taxi times in between.

The x-axis is more straightforward because it reflects a simpler encoding.
Each airline is encoded as the horizontal position of
a letter-value data symbol.
On the x-axis, we encode each airline as the horizontal position of
a tick mark and a text label.
A vertical grid line connecting the letter-value data symbols and 
the tick marks helps to form a strong visual grouping of data symbols
with their respective labels.

However, the encoding of airlines to text is missing a learned encoding
(@sec-learned-text),
at least for infrequent travellers internal to the United States of America.
For example, not all viewers will immediately decode the text "MQ" to
the name of an airline.
In effect, the x-axis tick labels only involve an encoding of airline 
as the **pattern** of the text.  We can only tell that
there are different airlines because they use different combinations of letters.

The legend provides information about the colour encoding of letter values.
The data density is encoded as the luminance of a stack of squares,
with every fourth square having a different hue/chroma (orange).
In addition, the letter values are encoded as individual text characters
alongside the squares.

One problem is that there is no learned decoding for the letter values.
For example, a letter value of "Z" is not immediately decoded as 
a meaningful data value.
As with the x-axis, the encoding 
only provides us with a different
text pattern for each letter value.
We can identify that the letter values are different from
each other, but nothing more.

In particular, there is no way to decode the absolute data density 
from the legend.  The luminance of the squares gives an indication of
relative data density (darker squares are higher density)

Another minor problem is that the letter values are placed alongside 
each rectangle, but they really correspond to the boundaries between
the rectangles in the letter-value data symbols.
For example, the true position of letter value "F" is at the top (and bottom)
of the black rectangle in each letter-value data symbol.

Finally, there is no explanation of the width of the rectangles in the
letter-value data symbols.  There is no guide that explains how
the depth of the letter values is encoded as the width of the 
letter-value data symbols.

### Encoding structure

Another problem with the legend in @fig-lv-default is that
the squares within the legend are not consistent with the
rectangles in the letter-value data symbols
(@sec-repetition).
Specifically, the rectangles in the letter-value data symbols
have a grey border, but the squares in the legend have no border.
This also makes it harder to match the colours in the legend
with the colours in letter-value data symbols because
the surrounding colours are not the same (@sec-colour-relative).

A final small inconsistency in the legend is that the 
luminance of the squares in the legend decrease from top to bottom,
while the luminance of the rectangles in the letter-value data symbols
decreases from the middle both up and down, with the clearest decrease
upwards, from middle to top.

There are also several horizontal grid lines in @fig-lv-default.
While these can help with decoding individual letter values that
lie exactly on a grid line, thanks to the non-linear y-scale,
decoding letter values relative to the grid lines is restricted
to ordinal comparisons.
The large number of grid lines also adds to the overall visual complexity
of @fig-lv-default that we discussed earlier.

### Building something better?

Having used the information in this book to
identify a number of problems with the encodings in @fig-lv-default,
we now attempt to use the information 
to make improvements to @fig-lv-default.

@fig-lv-zoom shows a data visualisation that is based on the same data
taxi data as @fig-lv-default, but uses a different set of encodings.
One solution to the complexity of @fig-lv-default
would be to avoid multiple encodings by 
producing several, simpler data visualisations for different purposes.
For example, one data visualisation that just shows the distribution
shapes and another that shows the location of specific letter-value taxi times.
However, @fig-lv-zoom still attempts to fit all of the information
from @fig-lv-default into a single data visualisation.

This is unlikely to be the optimal solution, but we can at least 
provide justifications for
the design choices that are being made.

```{r}
#| echo: false
source("flights.R")
source("lv-key.R")
```

```{r}
#| echo: false
#| label: fig-lv-zoom
#| warning: false
#| fig-height: 8
#| fig-cap: An improved letter-value plot?
ggwhole <- ggplot(flights) +
    grid_panel(lvkey) +
    coord_cartesian(clip="off") +
    geom_rect(aes(xmin=left, ymin=as.numeric(airline) - 1.5*height, 
                  xmax=right, ymax=as.numeric(airline) + 1.5*height,
                  fill=level), colour="white", alpha=.15,
              data=blockdf) +
    geom_rect(aes(xmin=left, ymin=as.numeric(airline) - 1.5*height, 
                  xmax=right, ymax=as.numeric(airline) + 1.5*height),
                  fill=palette.colors(9, "okabe-ito")[9], colour="white", 
                  alpha=.15,
              data=outBlockdf) +
    geom_line(aes(x, as.numeric(airline) + 1.5*y, group=airline), 
                  data=linedf, colour="black", linewidth=1.5/96*25.4) +
    geom_line(aes(x, as.numeric(airline) + -1.5*y, group=airline), 
                  data=linedf, colour="black", linewidth=1.5/96*25.4) +
    geom_segment(aes(x=median, y=as.numeric(airline) + 1.5*medspike, 
                     xend=median, yend=as.numeric(airline) + -1.5*medspike,),
                 data=unique(spikedf[c("median", "medspike", "airline")]), 
                 lineend="butt") +
    geom_point(aes(taxi, airline), data=outdf, size=2, pch=21, 
               colour="white", fill="black") +
    scale_y_discrete(expand=expansion(c(.05, .05))) +
    scale_colour_okabe_ito(order=c(9, 1:8)) +
    scale_fill_okabe_ito(order=c(9, 1:8)) +
    ylab(NULL) +
    xlab("Taxi Time (in minutes)") +
    theme_bw() +
    theme(plot.margin=margin(60, 5, 5, 5),
          panel.grid.major=element_blank(),
          panel.grid.minor.x=element_blank(),
          axis.ticks.y=element_blank(),
          axis.text.y=element_blank(),
          legend.position="none")
gglower <- ggplot(flights) +
    coord_cartesian(clip="off") +
    geom_rect(aes(xmin=left, ymin=as.numeric(airline) - 1.5*height, 
                  xmax=right, ymax=as.numeric(airline) + 1.5*height,
                  fill=level), colour="white", alpha=.2,
              data=subset(blockdf, side == "lower")) +
    geom_rect(aes(xmin=left, ymin=as.numeric(airline) - 1.5*height, 
                  xmax=right, ymax=as.numeric(airline) + 1.5*height),
                  fill=palette.colors(9, "okabe-ito")[9], colour="white", 
                  alpha=.2,
              data=outBlockdf[c(TRUE, FALSE),]) +
    geom_line(aes(x, as.numeric(airline) + 1.5*y, group=airline), 
                  data=subset(linedf, side == "lower"), colour="black", 
                  linewidth=1.5/96*25.4) +
    geom_line(aes(x, as.numeric(airline) + -1.5*y, group=airline), 
                  data=subset(linedf, side == "lower"), colour="black", 
                  linewidth=1.5/96*25.4) +
    geom_segment(aes(x=median, y=as.numeric(airline) + 1.5*medspike, 
                     xend=median, yend=as.numeric(airline) + -1.5*medspike,),
                 data=unique(spikedf[c("median", "medspike", "airline")]), 
                 lineend="butt") +
    geom_point(aes(taxi, airline), size=2, pch=21, 
               data=subset(outdf, side == "lower"),
               colour="white", fill="black") +
    scale_y_discrete(expand=expansion(c(.05, .05))) +
    scale_colour_okabe_ito(order=c(9, 1:8)) +
    scale_fill_okabe_ito(order=c(9, 1:8)) +
    ylab(NULL) +
    xlab("below median") +
    theme_bw() +
    theme(plot.margin=margin(60, 5, 5, 5),
          panel.grid.major=element_blank(),
          panel.grid.minor.x=element_blank(),
          legend.position="none")
grid.newpage()
pushViewport(viewport(layout=grid.layout(1, 2, widths=c(1, 2))))
pushViewport(viewport(layout.pos.col=1))
print(gglower, newpage=FALSE)
popViewport()
pushViewport(viewport(layout.pos.col=2))
print(ggwhole, newpage=FALSE)
popViewport()
```

* The airlines are encoded as the vertical position of letter-value symbols
  (just a swapping of axes from @fig-lv-default).
  This is to allow for the longer y-axis labels.

* The airline has been encoded as the full airline names in the
  text labels on the y-axis.
  It is now possible to decode the airline for each 
  letter-value data symbol.

* The airlines have been ordered by median.  This simplifies
  the data visualisation a little because there is a simple 
  monotonic trend in
  the median lines, rather than a series of increases and decreases.
  It also makes comparisons easier between more similar airlines
  because they are now closer together (@sec-distance-effect).

* Instead of a non-linear scale for encoding the taxi values,
  the encoding is linear. 
  This makes it possible to decode individual letter value taxi times
  and compare them between airlines.

* Because the linear scale makes it even more difficult to see
  the distribution below the median, an extra panel has been added 
  that shows the distribution below the median with a higher-resolution
  scale.

* Taxi times are encoded to a letter-value data symbol for each airline
  based on the letter values and the 
  data density.  The total area of each 
  letter-value data symbol is the same and the narrowness of the data
  symbol properly encodes the amount of data between different letter values;
  we can more effectively decode data summaries of the distribution
  of taxi times for different airlines.

* Letter values are encoded as the position of coloured bars 
  behind the letter-value data symbols.
  This allows the location of the letter values to be decoded
  regardless of the narrowness of the letter-value data symbol.
  There is also a bar that encodes the extent of the outliers.

* Letter values are also encoded as the hue of the coloured bars.
  This qualitative colour palette is used to make it easier 
  to differentiate letter values and compare them between airlines.
  Each hue is repeated twice to reduce the amount of visible change,
  while still allowing easy decoding.
  The colours are from the Okabe-Ito colour palette (@sec-palettes),
  with light grey reserved for the outliers bar.
  
* The legend is more consistent with the letter-value data symbols.
  The letter values are placed at the borders of the bars.
  The data density is encoded as vulgear-fraction text labels.
    

---

```{=latex}
\theendnotes
```
