
```{r}
#| echo: false
#| message: false
source("common.R")
```

# Review {#sec-review}

Add review at the end that reinforces the threads like visual feature, 
visual shape, visual object and encodings

If we encode X using Y encoding what can we decode?

Possible encoding end points:

* visual feature
* visual summary
* emergent feature
* visual shape 
* visual object
* text

Possible decodings from encoding end points:

* qual/quant data value
* data summary
* higher level information/knowledge/wisdom

Do two versions (at least):

* Set of questions for when you already have a plot and want to understand why
  it works or not

* Another set for if you want to create a data vis

Basic questions:

* What are you asking the visual system to do?
* Is the visual system good at that task? 
* Are you asking the viewer to produce visual summaries
  when it would be easier to produce data summaries?

Present process by which a data vis can be broken down.
What data symbols are being used?  (simple or complex)
What encodings are being used? 
What do we want to decode? (data or summaries)
If encoding back to raw data, are the encodings accurate?
What can be encoded back to?
Are encodings congruent with data?

...

List of questions to evaluate a data vis (maybe nest these a bit):

* What do you want the viewer to be able to decode?
* Do the encodings support that goal?

* What are the data variables
* What visual features are they encoded to
* How many encodings are there? (complexity)
* How many encodings are there *per data symbol* ?
* What type of data
* Can the visual features represent the data
* Is there the capacity to represent all categories
* How accurate are the decodings
* Is there congruence or dissonance in the encoding
* Encoding data or data summary
* Do visual features interact
* Are there emergent features
* Is the data symbol a visual shape
* Is that appropriate
* A visual object
* Is that appropriate
* Learned encodings somehow (distinct from implicit encodings)

* Do all visual differences correspond to differences in the data?
* Are all differences in the data represented by visual differences? 

* Does each data value encode to its own data symbol
  If not, individual data values may be difficult to decode 
* Do multiple data values encode to a single data symbol
  If not, data summaries may be difficult to decode 


And so on ...

## Examples

* Use heatmap as case study?

* do the side-by-side pie chart idea as an example of reasoning your
  way to a new visualisation.

* Move the discussion/revisit of tables from text section to here ?

## Case study: Lines of best fit

@fig-lm-smooth shows a scatter plot of the winning margin
versus the winning score for Tier One nations at Rugby World Cup matches.
The hemisphere of the losing team is encoded as the colour of the data
points and a trend line has been added for each hemisphere.

The trend lines are encodings of data summaries, least-squares
regressions, to line data symbols.  The data summaries consist
of intercept and slope values, which are encoded as the vertical positions
and the angles of the lines, respectively (@fig-stat-vis-decode).

The data symbol is very simple and decoding the data summaries is
also simple because it just involves decoding angle and vertical position.
For example, we can see that, in general, the winning margin is larger
when the winning score is larger, but 
the trend lines suggest that the increase in the winning margin
is slower for southern-hemisphere teams (they fight harder in defeat).

```{r}
#| echo: false
#| label: fig-lm-smooth
#| fig-cap: A scatter plot of the winning margin versus the winning score for matches between Tier One nations at Rugby World Cups.  There is a light shading of the lower-right diagonal to indicate that the winning margin cannot exceed the winning score.  In addition, points on the diagonal represent matches where the losing team failed to score any points.  The straight lines represent a least-squares regression model fit.
#| warning: false
rwcWinner <- do.call(rbind, lapply(split(rwcAll, rwcAll$match), 
                                   function(x) {
                                       if (all(x$scored == x$conceded)) {
                                           winner <- 1
                                           loser <- 2
                                       } else {
                                           winner <- x$scored > x$conceded
                                           loser <- !winner
                                       }
                                       x$oppSphere <- x$hemisphere[loser]
                                       x[winner, ]
                                   }))
ggplot(rwcWinner, aes(scored, scored - conceded, colour=oppSphere)) +
    annotate("polygon", x=c(0, Inf, Inf), y=c(0, Inf, 0), fill="grey97") +
    geom_smooth(method="lm", se=FALSE) +
    geom_point(alpha=.2) +
    scale_x_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning score") +
    scale_y_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning margin") +
    scale_colour_discrete(guide=guide_legend(title="opposition\nhemisphere")) +
    coord_cartesian(clip="off") +
    theme(aspect.ratio=1)
```

@fig-loess-smooth shows the same data as @fig-lm-smooth, but this time
there are curved trend lines added rather than straight trend lines.[^loess]
The curves are encodings of data summaries again, but this time 
loess model fits.  This means that the encoding involves a series
of predicted values encoded as horizontal and vertical positions for
the curve.  In other words, the data summaries are encoded as visual features
that produce a visual shape

[^loess]:
    These are *loess* model fits (locally-estimated scatterplot smoothing).

```{r}
#| echo: false
#| label: fig-loess-smooth
#| fig-cap: Scatter plot
#| warning: false
ggplot(rwcWinner, aes(scored, scored - conceded, colour=oppSphere)) +
    annotate("polygon", x=c(0, Inf, Inf), y=c(0, Inf, 0), fill="white") +
    geom_smooth(method="loess", se=FALSE, span=1) +
    geom_point(alpha=.2) +
    scale_x_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning score") +
    scale_y_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning margin") +
    scale_colour_discrete(guide=guide_legend(title="opposition\nhemisphere")) +
    coord_cartesian(clip="off") +
    theme(aspect.ratio=1)
```

Nice example of a complex calculation, the results of which are what gets
visualised.  Also example of performing the calculation numerically rather
than asking the viewer to perform it visually.  This was in summaries.qmd,
but it cannot be there because it requires scatter plots, which come in
the next section.
Add this after mention of visualising models?

Add footnote to Hadley's more complex model visualisation paper
https://vita.had.co.nz/papers/model-vis.pdf
https://dl.acm.org/doi/10.1002/sam.11271

model fits like lm straight line and ref Hadley's data-space vs model-space
"data models" to mirror "data statistics"
(revisit box plots as VERY simple models? also talk about failing
assumptions of these models, like unimodal for boxplot)

footnote cite to Tukey for visualising phenomena rather than raw data?
"Data-Based Graphics: Visual Display in the Decades to Come"
Tukey (1990)

+ "Beware of Validation by Eye:
   Visual Validation of Linear Trends in Scatterplots"
  Braun et al (2024)
  People are biased towards orthogonal distance to line rather than
  vertical distance.

loess curves
Another example of data -> summary -> features -> shapes -> summaries

## Case study: 2D density plots

Stat summary visualised 

Compare with visual summary of density of points

Different pros for each approach

refer back to @sec-pos-scarce (overplotting)
and mention that shapes are one solution,
with all of the drawbacks of not being able to decode individual
data values.
Footnote surjective/injective?

## Case study: kiviat plots or radar plots or spider plots

  (David Peebles (2008))

## Case study:  A letter-value plot

@fig-lv-default
shows a letter-value plot of 
taxi time for over 450,000 flights within the United States,
broken down by airline.[^lvplot]
The letter-value plot is an example of a novel data visualisation, so
this provides a good opportunity to analyse whether a letter-value plot
is an effective encoding of information.[^strawman]

[^lvplot]:
    This plot reproduces Figure 2 from @letter-value, which was
    produced with the R package {lvplot} [@lvplot].

[^strawman]:
    In order to incorporate as many points as possible,
    this analysis just focuses on the default {lvplot} settings that
    are used in @fig-lv-default
    and even sinks to the level of critiquing the specific data set
    (and non-linear scales) that are used in this specific plot. 
    This is neither an attack on letter-value plots in
    general nor an attack on the {lvplot} package in particular.

```{r}
#| echo: false
#| label: fig-lv-default
#| fig-height: 6
#| fig-cap: A letter-value plot of taxi times for over 450,000 flights within the United States.
ggplot(ontime, aes(UniqueCarrier, TaxiOut + TaxiIn)) +
  geom_lv(aes(fill = after_stat(LV))) +
  scale_fill_lv() +
  scale_y_sqrt(breaks = c(10, 50, 100, 150, 200, 250)) +
  xlab("Carrier") +
  ylab("Taxi Time (in minutes)") +
  theme_bw()
```

The purpose of a letter-value plot is to visualise the distribution
of a set of data values.
The information that we want to decode is about the distribution
of the data values:  the average value, the spread of the values,
the skewness, and so on.

The main motivation for the letter-value plot design is that a normal
boxplot produces *way* too many outliers for such a large amount of data.
This is shown in @fig-box-plot.
Furthermore, alternatives to the boxplot, like violin plots,
that fix the problem of too many outliers, are reliant on tuning parameters,
like kernels and bandwidths (@sec-density).

```{r}
#| echo: false
#| warning: false
#| label: fig-box-plot
#| fig-height: 6
#| fig-cap: A box plot
ggplot(ontime, aes(UniqueCarrier, TaxiOut + TaxiIn)) +
  geom_boxplot() +
  scale_y_sqrt(breaks = c(10, 50, 100, 150, 200, 250)) +
  xlab("Carrier") +
  ylab("Taxi Time (in minutes)") +
  theme_bw()
```

The letter-value plot aims to fix the problem of outliers, while at the 
same time, like
the original boxplot, remain free
of tuning parameters.
This is achieved by displaying as many letter values as are supported
by the amount of data, with only individual data values beyond
those limits shown as outliers.

### Letter values

The well-known
boxplot is based on the median and the upper and lower quartiles of
a data set, but these data summaries are just the first two levels of 
a more general concept of *letter values* introduced in 
"Exploratory Data Analysis" by @tukey1977eda.

Letter values are order statistics, where the $i$th letter value
approximates the quantile corresponding to a tail area that contains $2^{−i}$
of the observations in the data set.
The first letter value is the median, which splits the data into 
halves ($2^{-1}$), the second letter values are the upper and lower
quartiles, which leave a quarter of the data in each tail ($2^{-2}$),
and so on.  The amount of data between letter value $i$ and
letter value $(i+1)$ is $2^{-(i+1)}$.  For example,
the amount of data between the upper quartile (letter value 2) 
and letter value 3 is an eighth ($2^{-3}$).

The letter values are named M for median, F for fourths, E for eighths,
then, following the alphabet in reverse, 
D for sixteenths, C for thirty-seconds, down to
A for one-hundred and twenth-eighths, then starting again at the end of
the alphabet, in reverse alphabetic
order, Z for two-hundred and fifty-sixths, Y for five-hundred and twelfths,
and so on.

### Breaking down

We will use the framework that we have developed
in this book to analyse the data visualisation
in @fig-lv-default.  We will break @fig-lv-default down into 
the encodings that are being used and evaluate
the effectiveness of those encodings.

Data symbols
: There are two types of data symbol in a letter-value plot:
  complex data symbols consisting of multiple rectangles, which 
  we will call *letter-value* data symbols, to
  represent the distribution of the data values; and
  very simple data points, which represent the outliers (if any).

Encoding visual features
: In @fig-lv-default, the airlines (Carrier), which are qualitative data values,
  are encoded as the horizontal **position** of the letter-value symbols.
  The encoding of airline as the position of data symbols
  is very effective;  we can easily differentiate between the
  airlines.

  The other data values, the taxi times, are encoded to the
  letter-value data symbols within the plot.
  However, the letter-value plot encodes **data summaries**---letter 
  values---rather 
  than raw data values (@sec-summaries).[^letter-value]
  Consequently, all that we can expect to decode from the plot is information
  about the letter values, such as the location of specific letter values,
  plus possibly data summaries of letter values, such as the
  spread of letter values.
  
  [^letter-value]:
      Letter values are an interesting data summary in the sense that
      each letter value can correspond to a raw data value.
      In other words, letter values are a subset of the original raw data
      values, chosen to represent boundaries beyond which a certain
      proportion of the raw data values lie.
  
      This is not true for all letter values because a letter value
      may fall between two raw data values.
  
  The number of letter values that are displayed (before resorting to 
  showing individual points for outliers) is based on the amount of data values,
  so there are fewer rectangles for airlines with fewer tax times.
  For example, airline HA has fewer taxi times than airline AA, so ten
  letter values are displayed for airline HA, but thirteen letter values 
  are displayed for airline AA.[^cutoff]
  
  [^cutoff]:
      The calculation of the number of letter values to display
      is based on the uncertainty in the estimate of the letter values
      (assuming Gaussian data).
      We stop once a 95% confidence interval for a letter value
      overlaps with the neighbouring letter value.
  
      This calculation is relevant only for the default letter-value plot.
      Variations of the letter-value plot design allow for 
      different calculations of the number of letter values to display.
  
  Each letter value is encoded as the vertical **position** of the top
  (or bottom) of a rectangle within the letter-value symbol.
  The median (letter value 1) is marked as a single white line rather than
  a rectangle.  This is effective for decoding individual letter values
  for an airline.  For example, we can see that one of the letter values
  for the AA airline is just over 50.
  
  The *depth* of each letter value is encoded as the width (**length**)
  of a rectangle within the letter-value symbol.
  The depth is how many data values lie beyond that letter value;
  for example, the depth of the median is half the total number of 
  data values.
  This is also an encoding of a data summary.
  However, the actual encoding is only based on the *order* of the
  depth of the letter values,
  with regular decrements 
  from widest at the median (letter value 1) to narrowest at the final
  letter value (e.g., letter value 13 for airline AA).
  In effect, we are encoding a summary of a data summary.
  Using **length** for the encoding is effective, but we can only decode
  ordinal information about the depth of the letter values
  because that is all that we have encoded.
  For example, the width of the rectangle that encodes the depth
  of the fifth letter value (the bright orange rectangle) for
  the AA airline is just the fourth-widest rectangle.
  In particular, that rectangle is *not* proportional to the actual depth
  of the fifth letter value.  There is approximately one thirty-second
  of the data values beyond the fifth letter value, so this rectangle would
  have to be sixteen times narrower than the median for the widths
  to be proportional to the depth of the letter values.
  
  The *data density*---the number of data values between adjacent letter 
  values---is encoded as the **colour** of a rectangle.
  The **luminance** of the rectangle varies from darkest for the highest density
  (approximately one quarter of the data values lie between the median and
  the upper quartile) to lightest for lowest density
  (approximately one ten-thousandth of the data values between the most
  extreme letter values).
  In addition, the **hue**/**chroma** of a rectangle encodes a categorisation 
  of the letter values (yet another data summary);
  every fourth letter value is orange, while all the other letter values
  are grey.
  
  The luminance encoding of density only allows us to decode ordinal information
  about the data density between letter values.
  Like the widths of the rectangles,
  the shades of the rectangles steadily
  decrease upwards from the median and downwards from the median.
  We can also identify rectangles that represent the same letter value for
  different airlines by finding rectangles with the same shade in the
  letter-value symbols for two airlines.
  This is less effective due to the relatively limited capacity of
  luminance encoding (@sec-ordinal).
  For example,
  the different shades of grey are much more difficult to identify and
  compare between different airlines than,
  for example, the different horizontal locations for different airlines.
  
  The chroma encoding of every fourth letter value is very effective for
  decoding between those letter 
  values and all of the others, which is essentually an encoding of two groups
  (orange versus grey).
  It is also very effective for identifying the same
  letter value across different airlines.
  The difference in chroma between the oranges
  and the greys makes the oranges easy to identify
  and there are only three different
  luminances for the orange rectangles, so they are more
  easily differentiated.
  For example, it is easy to identify the rectangle for the fifth letter value 
  for all airlines;  these are the brightest orange rectangles.
  
  The outliers in a letter-value plot are encoded differently.
  The individual taxi times for outliers are encoded as the
  vertical position of data points, with the airline encoded as the
  horizontal position, as for the letter-value symbols.
  These allow for very simple and effective decodings.
  It is even possible to rapidly decode the number of outliers for
  each airline
  thanks to the small number of data symbols involved
  (@sec-visual-summaries).
  
  @tbl-lv-encodings summarises the encodings that are involved in a 
  letter-value plot and what we can decode from them.
  There are several effective encodings for both quantitative and qualitative
  data values, but the encoding of data density and the encoding of
  letter-value depth both lose some information.
  Those are quantitative data summaries that have only ordinal encodings.
  One overall point is that this is quite a large
  number of encodings for a single data visualisation.  This means that
  the encodings in @fig-lv-default produce a relatively complex data
  visualisation.  There are many visual features changing, which
  makes each individual encoding in @fig-lv-default more
  difficult to decode (@sec-differences).  Compare this with
  @fig-box-plot which, for all its faults, is a much
  simpler data visualisation because it involves fewer changes in visual features.
  
  In defence of @fig-lv-default, these decodings of individual data
  values and data summaries, while potentially useful, are not the primary 
  aim of the data visualisation, so they are less important
  than how well we can decode the distribution of the data values,
  which we will come to later.

::: {#tbl-lv-encodings}

| Value | Raw/Summary | Type | Encoding | Decoding |
| :----- | :----------- | :---- | :-------- |
| airline | raw data value | qualitative | position (horiz) | airline |
| taxi outlier | raw data value | quantitative | position (vert) | taxi time |
| taxi letter value | summary of taxi | quantitative | position (vert) | letter value |
| letter value depth | summary of letter value | qualitative | length (width) | depth order |
| taxi density | summary of taxi | quantitative | colour (luminance) | density order |
| fourth letter value | summary of letter value | qualitative | colour (chroma) | letter value order |

The encodings involved in a letter-value plot.
:::

Combining encodings
: There is one obvious interaction between the encodings in @fig-lv-default.
  The width of the rectangles (letter-value depth order) and the height of the
  rectangles (difference between adjacent letter values) together produce
  the emergent feature of **area** for the rectangles.
  
  Unfortunately, this area does not correspond to any data value or summary of 
  the data.  Letter value depth order 
  multiplied by difference between letter values
  does not give a meaningful value.
  This makes @fig-lv-default confusing because the 
  rectangles areas do not decode to anything useful 
  (@sec-danger-combine).[^unfair]
  
  [^unfair]:
      This point is only relevant to the default letter-value plot.
      A variation of the letter-value plot design involves a different encoding
      of width so that the area of the rectangle reflects the 
      proportion of the data between the letter values.
  
  Furthermore, the **implicit decoding** of the areas of the 
  rectangles in the letter-value data symbols
  suggests that, for example, the (darkest) rectangle just above the mean
  is about five times the size of the 
  highest (lightest) 
  rectangle for the "AA" airline.
  However, the proportion of the data that lies between the median 
  and the next letter value is actually over sixteen thousand times
  more than the proportion of the data that lies between the top two letter
  values.
  In other words, the area of the rectangles is actually *misleading*.
  
  There is also a **redundant encoding** in @fig-lv-default (@sec-redundant).
  Because the order of a letter value is related to the 
  data density order,
  both the luminance of each rectangle in the letter-value symbol
  and the width of the rectangle decode to essentially the same information.
  For example, the bright orange rectangles, which correspond to the data 
  density of the fifth
  letter value, all have exactly the same width.
  This has the benefit of providing more than one way to decode information.
  However, given that we have already identified that there are
  a large number of encodings in @fig-lv-default, this may be a luxury that
  we can ill afford.
  
Non-linear scales
: An extra complication with the encoding of vertical position
  in @fig-lv-default is that the vertical scale is non-linear 
  (@sec-non-linear-pos);
  it is actually the *square root* of the letter values that are
  encoded as the vertical position of the tops (or bottoms) of the
  rectangles in the letter-value data symbols.
  This makes the decoding of letter values and the comparison 
  of letter values much more difficult because the same physical difference
  does not decode to the same difference between letter values.
  For example, for the AA airline, although most of the letter values
  appear to be similar distances apart, in terms of taxi data values, 
  the letter values are in reality further apart for more extreme letter values.
  
  We have already established that the areas of the rectangles 
  in the letter-value data symbols provide no useful information and
  are potentially misleading.
  The non-linear scale only makes that potentially 
  misleading information even more
  misleading.

Visual shapes
: The letter-value data symbols encode multiple letter values within
  a single data symbol, which produces a **visual shape**
  (@sec-visual-shapes).
  We have a data symbol that is widest at the median of the data values
  and narrows towards either extreme.
  
  This allows us to decode useful data summaries about the distribution of 
  the taxi times.
  For example, the fact that all of the letter-value data symbols
  narrow more rapidly below the mean than they do above the mean
  suggests a skewed distribution of taxi times for all airlines
  (or at least a skewed distribution of the square root of taxi times).
  
  It is also possible to identify airlines that have either similar or different
  distributions based on similar or different visual shapes.
  For example, the letter-value 
  data symbol for the HA airline is clearly shorter than
  the other data symbols, which tells us that the range of taxi times
  is smaller for the HA airline.
  
  These are examples of effective decodings of data summaries from the
  visual shapes that are produced by the letter-value data symbols.
  This effectiveness is important because decoding information about
  the distribution of taxi times is the main purpose of a letter-value
  plot.
  
  One caveat is that the widths of the visual shapes are potentially
  misleading because they do not narrow anywhere near as fast as
  the distribution of taxi times.
  In other words, the shape of the letter-value data symbols
  does not provide an accurate decoding of quantitative
  features of the distribution of taxi times, only broad qualitative 
  features.
  For example, we can decode *whether* the distribution is symmetric or skewed,
  but we cannot decode *how* skewed the distribution is.

Encoding encodings
: We have already discussed that the y-axis scale is non-linear.
  The square root of taxi times, rather then the raw taxi times, are encoded 
  as the vertical position of the tops (or bottoms) of rectangles.
  This creates the usual problems to encode a non-linear scale
  as tick marks and text labels.
  For example, the square-root taxi time of 
  10 is encoded as a position about two-thirds
  of the way up the y-axis,
  which accurately reflects the fact that 10 is approximately
  two-thirds of the range of square-root taxi times.
  However, this value of 10 is also encoded as the text "100".
  
  On one hand, this is a valuable encoding of the value 10 because it
  saves us the mental effort of having to reverse the square root transformation.
  If we had encoded this position as the text "10", then we would be required
  to perform the mental calculation to decode the 
  taxi time 100.  
  
  On the other hand, encoding the square root taxi times as text that 
  corresponds to the taxi times causes problems because we cannot 
  linearly interpolate between the text values.
  We can correctly decode the precise taxi time 100 from the
  text "100", and 50 from "50", but it is very difficult to decode
  any taxi times in between.
  
  The x-axis is more straightforward because it reflects a simpler encoding.
  Each airline is encoded as the horizontal position of
  a letter-value data symbol.
  On the x-axis, we encode each airline as the horizontal position of
  a tick mark and a text label.
  A vertical grid line connecting the letter-value data symbols and 
  the tick marks helps to form a strong visual grouping of data symbols
  with their respective labels.
  
  However, the encoding of airlines to text is missing a learned encoding
  (@sec-learned-text),
  at least for infrequent travellers internal to the United States of America.
  For example, not all viewers will immediately decode the text "MQ" to
  the name "Envoy Air".
  In effect, the x-axis tick labels only involve an encoding of airline 
  as the **pattern** of the text.  We can only tell that
  there are different airlines because they use different combinations of letters.
  
  The legend provides information about the colour encoding of the data density
  between letter values.
  The data density is encoded as the luminance of a stack of squares,
  with every fourth square having a different hue/chroma (orange).
  In addition, the letter values are encoded as individual text characters
  alongside the squares.
  
  One problem is that there is no learned decoding for the letter values.
  For example, a letter value of "Z" is not immediately decoded as 
  a meaningful data value.
  Not all viewers will immediately decode "Z" to a tail area of $2^{-8}$.
  As with the x-axis, the encoding 
  only provides us with a different
  text pattern for each letter value.
  We can identify that the letter values are different from
  each other, but little more.
  
  In particular, there is no way to decode the absolute data density 
  from the legend.  The luminance of the squares only gives an indication of
  the data density order thanks to the implicit 
  decoding of darker to more (darker squares are higher density;
  @sec-ordinal)
  
  Another minor problem is that the letter values are placed alongside 
  each rectangle, but they really correspond to the boundaries between
  the rectangles in the letter-value data symbols.
  For example, the true position of letter value "F" is at the top (or bottom)
  of the black rectangle in each letter-value data symbol.
  
  Finally, there is no explanation of the width of the rectangles in the
  letter-value data symbols.  There is no legend that explains how
  the depth order of the letter values is encoded as the width of the 
  letter-value data symbols.

Encoding structure
: Another problem with the legend in @fig-lv-default is that
  the squares within the legend are not consistent with the
  rectangles in the letter-value data symbols
  (@sec-repetition).
  Specifically, the rectangles in the letter-value data symbols
  have a grey border, but the squares in the legend have no border.
  This also makes it harder to match the colours in the legend
  with the colours in letter-value data symbols because
  the surrounding colours are not the same (@sec-colour-relative).
  
  A final small inconsistency in the legend is that the 
  luminance of the squares in the legend decrease from top to bottom,
  while the luminance of the rectangles in the letter-value data symbols
  decreases from the middle both up and down, with the clearest decrease
  upwards, from middle to top.
  
  There are also several horizontal grid lines in @fig-lv-default.
  While these can help with decoding individual letter values that
  lie exactly on a grid line, thanks to the non-linear y-scale,
  decoding letter values relative to the grid lines is restricted
  to ordinal comparisons.
  The large number of grid lines also adds to the overall visual complexity
  of @fig-lv-default that we discussed earlier.

### Building back

Having used the framework in this book to
identify a number of problems with the encodings in @fig-lv-default,
we now attempt to use the information 
to make improvements to @fig-lv-default.

@fig-lv-zoom shows a data visualisation that is based on the same 
taxi time data as @fig-lv-default, but uses a different set of encodings.
One solution to the complexity of @fig-lv-default
would be to avoid multiple encodings by 
producing several, simpler data visualisations for different purposes.
For example, one data visualisation that just shows the distribution
shapes and another that shows the location of specific letter-value taxi times.
However, @fig-lv-zoom still attempts to fit all of the information
from @fig-lv-default into a single data visualisation.

@fig-lv-zoom does not aim to be an optimal solution, but rather
a demonstration that we can use this framework to suggest changes
based on deliberate reasoning and that we can provide justifications for
those changes.

```{r}
#| echo: false
source("flights.R")
source("lv-key.R")
```

```{r}
#| echo: false
#| label: fig-lv-zoom
#| warning: false
#| fig-height: 8
#| fig-cap: An improved letter-value plot?
alpha <- .15
ggwhole <- ggplot(flights) +
    grid_panel(lvkey) +
    coord_cartesian(clip="off") +
    geom_rect(aes(xmin=left, ymin=as.numeric(airline) - 1.5*height, 
                  xmax=right, ymax=as.numeric(airline) + 1.5*height,
                  fill=level), colour=rgb(1,1,1,.5), alpha=alpha,
              data=blockdf) +
    geom_rect(aes(xmin=left, ymin=as.numeric(airline) - 1.5*height, 
                  xmax=right, ymax=as.numeric(airline) + 1.5*height),
                  fill=palette.colors(9, "okabe-ito")[9], 
                  colour=rgb(1,1,1,.5), alpha=alpha,
              data=outBlockdf) +
    geom_line(aes(x, as.numeric(airline) + 1.5*y, group=airline), 
                  data=linedf, colour="black", linewidth=1.5/96*25.4) +
    geom_line(aes(x, as.numeric(airline) + -1.5*y, group=airline), 
                  data=linedf, colour="black", linewidth=1.5/96*25.4) +
    geom_segment(aes(x=median, y=as.numeric(airline) + 1.5*medspike, 
                     xend=median, yend=as.numeric(airline) + -1.5*medspike,),
                 data=unique(spikedf[c("median", "medspike", "airline")]), 
                 lineend="butt") +
    geom_point(aes(taxi, airline), data=outdf, size=2, pch=21, 
               colour="white", fill="black") +
    scale_y_discrete(expand=expansion(c(.05, .05))) +
    scale_colour_okabe_ito(order=c(9, 1:8)) +
    scale_fill_okabe_ito(order=c(9, 1:8)) +
    ylab(NULL) +
    xlab("Taxi Time (in minutes)") +
    theme_bw() +
    theme(plot.margin=margin(60, 5, 5, 5),
          panel.grid.major.y=element_blank(),
          panel.grid.minor.x=element_blank(),
          axis.ticks.y=element_blank(),
          axis.text.y=element_blank(),
          legend.position="none")
gglower <- ggplot(flights) +
    coord_cartesian(clip="off") +
    geom_rect(aes(xmin=left, ymin=as.numeric(airline) - 1.5*height, 
                  xmax=right, ymax=as.numeric(airline) + 1.5*height,
                  fill=level), colour=rgb(1,1,1,.5), alpha=alpha,
              data=subset(blockdf, side == "lower")) +
    geom_rect(aes(xmin=left, ymin=as.numeric(airline) - 1.5*height, 
                  xmax=right, ymax=as.numeric(airline) + 1.5*height),
                  fill=palette.colors(9, "okabe-ito")[9], 
                  colour=rgb(1,1,1,.5), alpha=alpha,
              data=outBlockdf[c(TRUE, FALSE),]) +
    geom_line(aes(x, as.numeric(airline) + 1.5*y, group=airline), 
                  data=subset(linedf, side == "lower"), colour="black", 
                  linewidth=1.5/96*25.4) +
    geom_line(aes(x, as.numeric(airline) + -1.5*y, group=airline), 
                  data=subset(linedf, side == "lower"), colour="black", 
                  linewidth=1.5/96*25.4) +
    geom_segment(aes(x=median, y=as.numeric(airline) + 1.5*medspike, 
                     xend=median, yend=as.numeric(airline) + -1.5*medspike,),
                 data=unique(spikedf[c("median", "medspike", "airline")]), 
                 lineend="butt") +
    geom_point(aes(taxi, airline), size=2, pch=21, 
               data=subset(outdf, side == "lower"),
               colour="white", fill="black") +
    scale_y_discrete(expand=expansion(c(.05, .05))) +
    scale_colour_okabe_ito(order=c(9, 1:8)) +
    scale_fill_okabe_ito(order=c(9, 1:8)) +
    ylab(NULL) +
    xlab("below median") +
    theme_bw() +
    theme(plot.margin=margin(60, 5, 5, 5),
          panel.grid.major.y=element_blank(),
          panel.grid.minor.x=element_blank(),
          legend.position="none")
grid.newpage()
pushViewport(viewport(layout=grid.layout(1, 2, widths=c(1, 2))))
pushViewport(viewport(layout.pos.col=1))
print(gglower, newpage=FALSE)
popViewport()
pushViewport(viewport(layout.pos.col=2))
print(ggwhole, newpage=FALSE)
popViewport()
```

Data symbols
: One issue that we identified with @fig-lv-default
  was the high complexity of the letter-value data symbols,
  which arises from having multiple encodings within a single
  data symbol.
  One option is to have more data symbols, each with fewer
  encodings.

  There are *three* types of data symbols in @fig-lv-zoom.
  As in @fig-lv-default, there are data points to represent outliers
  and there is a letter-value data symbol to represent the 
  distribution of taxi times.
  In addition, there are rectangles that represent the letter values.

Encoding visual features
: A number of encodings are the same as before. 
  For example, the airline and taxi times for outliers are
  encoded as the position of data points, though
  the axes have been swapped, so that taxi times are now
  horizontal positions and airlines are vertical positions.

  Also as before, instead of raw taxi times, letter-value data summaries are
  encoded (other than for outliers).
  However, the encoding of letter values is different.
  Letter values are encoded as the position of coloured rectangles so that
  the edges of the rectangles decode to the individual letter values
  and the width of the
  rectangles decodes to differences between letter values.
  There are also rectangles that span the range of the outliers.
  These encodings are effective for decoding quantitative letter values
  from the edges and widths of the rectangles.

  In addition, letter values (or the spaces between) are encoded as
  the colour of the rectangles.
  Pairs of consecutive letter values are encoded to different hues
  from the Okabe-Ito colour palette (@sec-palettes),
  with light grey reserved for the outlier rectangles.
  Each hue is repeated twice to reduce the amount of change,
  while still allowing easy decoding.
  Although this encoding loses the quantitative information
  of the letter values,
  the qualitative encoding 
  is effective for identifying the same letter values across
  different airlines.
  
  The combined encodings of letter values 
  mean that it is easy to decode and compare individual letter values.
  For example, we can see that letter value E for Southwest Airlines
  is approximately 25 minutes
  and we can see that Envoy Air has the highest letter value B.

Visual shapes
: The letter values are also encoded as the horizontal positions
  of a black polygon for each airline.
  The data density of the taxi times is encoded
  as the vertical positions of the polygon so that
  the total area of each 
  letter-value data symbol is the same 
  (i.e., the polygon reflects the density distribution of the taxi times).
  This means that the narrowness of the polygon
  properly encodes the amount of data between different letter values.
  In other words, the implicit decoding of the area of the 
  polygons is not misleading.

  The polygon for each airline has little value for decoding individual
  letter values, but it produces a visual shape that
  allows us to decode data summaries of the distribution
  of taxi times for different airlines.
  For example, we can see that all airlines are right-skewed,
  but Hawaiian Airlines and Southwest Airlines are more skewed than
  the other airlines.

  Detail about the spread of values within the long tails (longer
  taxi times) can be decoded from the coloured rectangles.
  For example, Envoy Air has a fatter tail (more longer taxi times)
  than other airlines because, for each letter value, the Envoy Air rectangle
  is further to the right than for other airlines (light blues are
  further right, greens are further right, yellows are further right, etc).

Non-linear scales
: Another feature of @fig-lv-default that created some problems
  was the use of a non-linear scale.
  In @fig-lv-zoom, the encoding of taxi values
  and letter values to horizontal position is linear. 
  This makes it possible to decode individual letter value 
  and outlier taxi times
  and compare them between airlines.
  For example, we can see that the longest taxi time for
  Southwest Airlines is approximately 50 minutes longer than
  the longest taxi time for Hawaiian Airlines. 

  Because the linear scale makes it even more difficult to see
  the distribution below the median, an extra panel has been added 
  that shows the distribution below the median with a higher-resolution
  scale.
  For example, we can see that Hawaiian Airlines and Southwest Airlines 
  have a much more truncated distribution below the median than all other
  airlines.

Encoding encodings
: We identified several issues with the labelling of axes and the
  legend in @fig-lv-default and these have been improved in 
  @fig-lv-zoom.

  The airline has been encoded as the full airline names in the
  text labels on the y-axis.
  These text labels are full words with learned decodings.
  This makes it possible not only to differentiate between airlines,
  but also to decode the actual airline name for each 
  letter-value data symbol.
  For example, we can see that the bottom airline is Southwest Airlines 
  rather than WN.
  
  The airlines are also encoded as the vertical position of letter-value 
  symbols, which now run horizontally rather than vertically
  (just a swapping of axes from @fig-lv-default).
  This is to allow space for the airline labels.

  The legend in @fig-lv-zoom has changed because there are different
  data symbols and encodings involved, but it is also
  enhanced in several ways:
  the legend has a polygon similar to the letter-value data symbol
  that is used for each airline (but based on a standard normal distribution);
  the legend has coloured rectangles corresponding to the rectangles in the
  main plot (but with widths based on a standard normal distribution);
  the letter values are placed at the coorect positions on the 
  borders of the bars; and
  examples of the data density between letter values 
  is encoded as vulgar-fraction text labels.
  These changes make the decoding of letter-value names, letter-value positions,
  and data density all more accurate.
    
As an additional improvement,
the airlines have been ordered by median.  This simplifies
the data visualisation a little because there is a simple 
monotonic trend in
the median lines, rather than a series of increases and decreases.
It also makes comparisons easier between more similar airlines
because they are now closer together (@sec-distance-effect).

Overall, the most important decoding of the distribution of taxi times
is improved because the visual shape of the polygon provides this
decoding while not being misleading, supplemented by the coloured rectangles,
which show more detail about the tails of the distribution.
In addition, thanks to linear scales and improved axes and legends,
the decoding and comparison of individual taxi times is also more
effective.
@fig-lv-zoom is not perfect, but it demonstrates that
the framework that we have developed allows us not only to deconstruct 
and identify issues with an existing data visualisation, but also 
to suggest ways to make improvements.

---

```{=latex}
\theendnotes
```
