
```{r}
#| echo: false
#| message: false
source("common.R")
```

# Review {#sec-review .unnumbered}

Add review at the end that reinforces the threads like visual feature, 
visual shape, visual object and mappings

If we encode X using Y encoding what can we decode?

Possible encoding end points:

* visual feature
* visual summary
* emergent feature
* visual shape 
* visual object
* text

Possible decodings from encoding end points:

* qual/quant data value
* data summary
* higher level information/knowledge/wisdom

Do two versions (at least):

* Set of questions for when you already have a plot and want to understand why
  it works or not

* Another set for if you want to create a data vis

Basic questions:

* What are you asking the visual system to do?
* Is the visual system good at that task? 
* Are you asking the viewer to produce visual summaries
  when it would be easier to produce data summaries?

Present process by which a data vis can be broken down.
What data symbols are being used?  (simple or complex)
What mappings are being used? 
What do we want to decode? (data or summaries)
If mapping back to raw data, are the mappings accurate?
What can be mapped back to?
Are mappings congruent with data?

...

List of questions to evaluate a data vis (maybe nest these a bit):

* What do you want the viewer to be able to decode?
* Do the mappings support that goal?

* What are the data variables
* What visual features are they mapped to
* What type of data
* Can the visual features represent the data
* Is there the capacity to represent all categories
* How accurate are the decodings
* Is there congruence or dissonance in the mapping
* Mapping data or data summary
* Do visual features interact
* Are there emergent features
* Is the data symbol a visual shape
* Is that appropriate
* A visual object
* Is that appropriate
* Learned mappings somehow (distinct from implicit mappings)

* Do all visual differences correspond to differences in the data?
* Are all differences in the data represented by visual differences? 

* Does each data value map to its own data symbol
  If not, individual data values may be difficult to decode 
* Do multiple data values map to a single data symbol
  If not, data summaries may be difficult to decode 


And so on ...

## Examples

* Use heatmap as case study?

* do the side-by-side pie chart idea as an example of reasoning your
  way to a new visualisation.

* Move the discussion/revisit of tables from text section to here ?

* Another possible application of data vis knowledge:
  Analyse one of the weird boxplot variations that look a bit like a ziggurat?
  
  This is letter-value plot:

  + "Letter-Value Plots: Boxplots for Large Data"
    Hofmann et al (2015)

  + Nice explanation of calculations ...

    24  Beyond the Boxplot: Tukey’s Letter-Value Summaries – Exploratory Data Analysis in R https://share.google/M2qXCsIxsSsqM7jFQ 

  + Labelling is an example of a text encoding that has no learned decoding
    For most people
    So the text is just a symbol like a pch
    You get difference but that is all

  + {lvplot}

    This line suggests that width is calculated to be proportional
    Or at least can be ...

    https://github.com/hadley/lvplot/blob/51ad401be5e017d37883ec5d149bc1818246b6f5/R/draw.r#L43

## Case study: Lines of best fit

@fig-lm-smooth shows a scatter plot of the winning margin
versus the winning score for Tier One nations at Rugby World Cup matches.
The hemisphere of the losing team is encoded as the colour of the data
points and a trend line has been added for each hemisphere.

The trend lines are mappings of data summaries, least-squares
regressions, to line data symbols.  The data summaries consist
of intercept and slope values, which are encoded as the vertical positions
and the angles of the lines, respectively (@fig-stat-vis-decode).

The data symbol is very simple and decoding the data summaries is
also simple because it just involves decoding angle and vertical position.
For example, we can see that, in general, the winning margin is larger
when the winning score is larger, but 
the trend lines suggest that the increase in the winning margin
is slower for southern-hemisphere teams (they fight harder in defeat).

```{r}
#| echo: false
#| label: fig-lm-smooth
#| fig-cap: A scatter plot of the winning margin versus the winning score for matches between Tier One nations at Rugby World Cups.  There is a light shading of the lower-right diagonal to indicate that the winning margin cannot exceed the winning score.  In addition, points on the diagonal represent matches where the losing team failed to score any points.  The straight lines represent a least-squares regression model fit.
#| warning: false
rwcWinner <- do.call(rbind, lapply(split(rwcAll, rwcAll$match), 
                                   function(x) {
                                       if (all(x$scored == x$conceded)) {
                                           winner <- 1
                                           loser <- 2
                                       } else {
                                           winner <- x$scored > x$conceded
                                           loser <- !winner
                                       }
                                       x$oppSphere <- x$hemisphere[loser]
                                       x[winner, ]
                                   }))
ggplot(rwcWinner, aes(scored, scored - conceded, colour=oppSphere)) +
    annotate("polygon", x=c(0, Inf, Inf), y=c(0, Inf, 0), fill="grey97") +
    geom_smooth(method="lm", se=FALSE) +
    geom_point(alpha=.2) +
    scale_x_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning score") +
    scale_y_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning margin") +
    scale_colour_discrete(guide=guide_legend(title="opposition\nhemisphere")) +
    coord_cartesian(clip="off") +
    theme(aspect.ratio=1)
```

@fig-loess-smooth shows the same data as @fig-lm-smooth, but this time
there are curved trend lines added rather than straight trend lines.[^loess]
The curves are mappings of data summaries again, but this time 
loess model fits.  This means that the mapping involves a series
of predicted values encoded as horizontal and vertical positions for
the curve.  In other words, the data summaries are encoded as visual features
that produce a visual shape

[^loess]:
    These are *loess* model fits (locally-estimated scatterplot smoothing).

```{r}
#| echo: false
#| label: fig-loess-smooth
#| fig-cap: Scatter plot
#| warning: false
ggplot(rwcWinner, aes(scored, scored - conceded, colour=oppSphere)) +
    annotate("polygon", x=c(0, Inf, Inf), y=c(0, Inf, 0), fill="white") +
    geom_smooth(method="loess", se=FALSE, span=1) +
    geom_point(alpha=.2) +
    scale_x_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning score") +
    scale_y_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning margin") +
    scale_colour_discrete(guide=guide_legend(title="opposition\nhemisphere")) +
    coord_cartesian(clip="off") +
    theme(aspect.ratio=1)
```

Nice example of a complex calculation, the results of which are what gets
visualised.  Also example of performing the calculation numerically rather
than asking the viewer to perform it visually.  This was in summaries.qmd,
but it cannot be there because it requires scatter plots, which come in
the next section.
Add this after mention of visualising models?

Add footnote to Hadley's more complex model visualisation paper
https://vita.had.co.nz/papers/model-vis.pdf
https://dl.acm.org/doi/10.1002/sam.11271

model fits like lm straight line and ref Hadley's data-space vs model-space
"data models" to mirror "data statistics"
(revisit box plots as VERY simple models? also talk about failing
assumptions of these models, like unimodal for boxplot)

footnote cite to Tukey for visualising phenomena rather than raw data?
"Data-Based Graphics: Visual Display in the Decades to Come"
Tukey (1990)

+ "Beware of Validation by Eye:
   Visual Validation of Linear Trends in Scatterplots"
  Braun et al (2024)
  People are biased towards orthogonal distance to line rather than
  vertical distance.

loess curves
Another example of data -> summary -> features -> shapes -> summaries

## Case study: 2D density plots

Stat summary visualised 

Compare with visual summary of density of points

Different pros for each approach

refer back to @sec-pos-scarce (overplotting)
and mention that shapes are one solution,
with all of the drawbacks of not being able to decode individual
data values.
Footnote surjective/injective?

## Case study: kiviat plots or radar plots or spider plots

  (David Peebles (2008))

## Case study: letter-value boxplots
