
# TODO {.unnumbered}

+ Put more of Review in Exec Summary 

+ Add the questions from review piecemeal to the chapter summaries

  Eg for intro
  One question we can immediately ask of any data vis is 
  Do the data contain the information that we want to decode?
  We have no hope of decoding information that is not encoded in the first place

  The review can just contain the main question
  Do the data contain the information that we want to decode?

  Continue that pattern for each chapter summary

  Eg for visual model
  Is the data vis trying to encode many simple values with many simple features
  or few complex values with familiar visual objects
  or both
  (if neither, then we might have a problem)


  These questions also lead nicely on to possible exercises:
  e.g., for intro
  give data vis
  give questions (decodings)
  order the questions in terms of easiest to hardest
  ties allowed
  impossible an option
  ask at least one question that involves data that has NOT been encoded
  ask question that involves raw data when only data summary available
  ask question that requires lots of mental arithmetic from raw data values

* add exercises at end of reach chapter

  Build them up from 787 labs 

+ Add ref to Coren and Girgus "Seeing is Deceiving" book for
  difference between horizontal and vertical decoding of line length?

+ shrink lots of plots into .9x.9 viewport ?

  Also look at adding titles and tweaking axis titles everywhere
  (OR rely on figure captions instead of titles?)

+ check for "encode to" instead of "encode as"

+ add review at end of each section and/or preview to start of 
  each section to reinforce the building concepts.

+ Use the  technique of foreshadowing summaries (at the start of each chapter)
  By the end of this chapter we will see how encoding using visual features 
  is an effective way to visualize data
  Examples of things to foreshadow are questions that Julia has raised
  Like what are we wanting to decode?

+ Also acknowledge things that are being ignored
  You might notice that we are ignoring the fact that there are two encodings
  We will get to their interaction in chapter 5
  Might need to fill these in as students ask questions

+ Keep all important points and chapter summaries 
  in terms of encoding and decoding

+ Rewrite visual model with much more emphasis on what you are able to decode
  Eg lots of simple info from visual features
  Worry less about giving the game away for later sections 

+ Add more back refs to @sec-visual-focus and @sec-memory 
  wherever the complexity/simplicity
  of a data visualisation is discussed (likely to already reference
  visual differences and/or visual similarities).

+ Check that every visual system topic is connected to a visual decoding topic.
  Possibly rename sections, e.g., "visual differences" -> "decoding differences"

+ Check that I have used "pattern" to refer to data point "shape" everywhere
  (both to avoid conflict with "visual shape" and to allow connection to
  early detection of "pattern" or "texture" or "spatial frequency" in 
  visual processing).

+ Is there a single data vis example that has a realistically
  large amount of data in it ?

  There are a couple in Visual Shapes

  Could add in an example or two in Multiple Mappings

+ Check Few (2004) quote!

+ Try pal_npg()(2)[1] as highlight colour ?

+ Flesh out all footnotes

+ Add alt text to ALL figures (fig-alt)

+ Check that colour palettes are ok for CVD

+ Add to accessibility
  Font contrast
  Size
  Color contrast
  W3c recommendations

+ Add contrast guidelines (such as WCAG recommendations)
  for text size and colour contrasts (at least)

+ Look for "Total" data frames and replace with subset(year == 2021) data frames
  (in example data visualisations)

+ Add example in design of ordering legend items same as
  data symbols OR congruently (e.g., North above South)

  "Looking at bar graphs in particular, Feeny, Hola, Liversedge, Findlay, and
  Metcalf(2000; see also Fischer, 2000) recently documented the importance of
  spatial congruencybetween textual and graphical information. Participants
  compared two bars labelled A andB, with a statement about the relation between
  A and B underneath (e.g. A > B).Veriﬁcation times were faster when the spatial
  order of the statement matched that ofthe labels in the graph"

  Designing bar graphs: orientation matters
  Martin H. Fischer, Nele Dewulf, Robin L. Hill
  https://onlinelibrary.wiley.com/doi/epdf/10.1002/acp.1105

+ Add ...

  "Misleading Beyond Visual Tricks: How People Actually Lie with Charts"
  Lisnic et al (2023)

  ... to footnote about ethical issues in data vis ?

+ Add ...

  "Ten Simple Rules for Better Figures"
  Rougier et al (2014)

  ... to examples of laundry list of data vis guidelines  ?

  Some of the rules are examples of very vague and unhelpful guidelines
  (while others are more specific, like start bars at zero)
  
+ Review case studies ...

  * Use heatmap as case study?

  * do the side-by-side pie chart idea as an example of reasoning your
    way to a new visualisation.

  * Move the discussion/revisit of tables from text section to here ?

  Case study: 2D density plots

  Stat summary visualised 

  Compare with visual summary of density of points

  Different pros for each approach

  refer back to @sec-pos-scarce (overplotting)
  and mention that shapes are one solution,
  with all of the drawbacks of not being able to decode individual
  data values.
  Footnote surjective/injective?

  Case study: kiviat plots or radar plots or spider plots

  (David Peebles (2008))

+ review https://friendly.github.io/6135/lectures/GraphPerception.pdf
  from https://friendly.github.io/6135/index.html
  (especially the stuff on basic models of visual perception
  and cognition!)
  to see that (i) I am covering everything and (ii) I am not
  telling too many lies.

+ Note that "correlations" do not have to be linear, e.g.,
  some of the pairwise scores in R Journal submission 2024-88 [p 3]
  are designed to pick up non-linear correlations (?)

+ Need to work in the idea that what is most is important is whether    
  the viewer can easily see what you want them to see.
  So, for example, it is ok to use less effective mappings like in
  a heatmap if you do not need to viewer to extract raw data values.
  Maybe this point needs to be made several times throughout the book.

* Early on when talk about how visual system is good at stuff and that is why
  data vis works

  Data vis is trading advantage of the visual system to immediately and
  effortlessly convey a message

  Flip the script and point out that if the viewer cannot immediately see what
  you want them to see then you are either doing it wrong or data vis is the
  wrong way to communicate the message

* When talk about basic data types and what you can do with them,
  maybe introduce the explicit idea of "comparison" and all tasks
  basically being based on comparison (with some examples of breaking
  down tasks).
  There is a nice simple list of tasks (rather than some overblown
  task taxonomy) in 
  "Low-Level Components of Analytic Activity in Information Visualization"
  Amar et al (2005)
  which describes comparison as a "fundamental cognitive action" that
  underlies all data vis tasks.

* Check whether HDVW covers all of the types of tasks in 
  "Low-Level Components of Analytic Activity in Information Visualization"
  Amar et al (2005)

* Add spine plot to congruence ?
  Proportions via bar plot

* In visual-model, make a BRIEF statement at the end of each section 
  (maybe in black box) about the implications for data vis. 
  BUT keep it in terms of encode/decode

* Add the purpose of each type of plot as they are introduced so that 
  explanation of effectiveness is relative to it's purpose
  Eg a bar plot is effective at... Because... 

* add plot of differences between lines in crimegroup line plot ?
  (do NOT perceive vertical difference)
  citation?  see franconeri 2021

* In Multiple Mappings, add an example of a numerical data summary
  that is encoded as basic visual features (e.g., MDS, PCA, ...).
  Compare encoding data summary as visual feature with
  encoding data as visual shape.
  Decoding more straightforward, but issues with interpreting
  data summaries?  

* check have not missed any major pros or cons for choropleth maps
  https://www.practicalreporting.com/blog/2024/8/21/ive-stopped-using-colored-region-choropleth-maps-should-you

  Should have section on maps (cartography) as well ?
  This is the source of a lot of the background information

  Also make sure that the issue of larger regions with less population 
  is a major source of misrepresentation.
  If you have something like count of crimes, then you first need to
  adjust by population (crime rate), but then you ALSO need to adjust
  by region size (population density).
  G. Beconyte, A. Balciunas, A. Sturaite, and R. Viliuviene, “Where maps lie:
  Visualization of perceptual fallacy in choropleth maps at different levels of
  aggregation,” ISPRS International Journal of Geo-Information, vol. 11, no. 1,
  p. 64, 2022. [Online].  Available: https://doi.org/10.3390/ijgi11010064

* Change from qualitative and quantitative to categorical and continuous ?
  (to be consistent about avoiding technical terminology)

  Shift nominal, ordinal, etc to footnote

  Just add ordered data in colour

  This would be a BIG edit.

* attributions for data sources 
  AND for images (like robber image)

* citations to stuff on pictographs and glyphs and icon-based plots
  (in clutter section?)

* add citations for R packages used in plots and diagrams.
  General one for R and ggplot2, then specific ones for specific plots
  Add citation to {ggsci} for default (discrete) palette

  Maybe an appendix ?

  R package {wordcloud} [@pkg-wordcloud].
  ggsci for colour palettes?

  Include {gMOIP} and {geometry} for HCL diagrams.

  Package for wind figure (e.g., rNOMADS)

* Get visualisation of uncertainty in somewhere ?
  Variability is a data summary, which we map to a visual feature.
  Box plots already do this.
  Only opportunity to demonstrate gradients as a visual feature?

* Add section on "Mapping Unintentionally" or something that looks at
  the information that is conveyed by style (e.g., conservative/liberal,
  young/old) and the interaction between the viewer's pre-existing
  attitudes.

  Goes in text section?

* Add footnote in @sec-quant-features to acknowledge that there are
  many other types of data (time series, cyclic, data with a 
  central neutral value, ...).  We see some of these in later chapters.

* Use the Data Rocks advice to generate a colour palette for this document ?
  
  https://buttondown.com/datarocksnz/archive/design-matters-6-the-ultimate-dashboard-colour/

* An index that links to all of the places that different sorts of plots
  are discussed ?

  Use Quarto book format (each .qmd becomes its own chapter) and
  then get search functionality for free (which is essentially 
  an automated index).

* For 3d rgl plot, add contingency code so that, if webgl is not available
  it falls back to a static image (online).

  Consider generating animations of both 3d rgl with rglplus::rgl.makemovie()
  and rayshader plots with render_movie()

* Match plot font to text font ?
  Would need to set more explicit text font on web 
  (currently system-ui is top of the list)

* Add sub-figs for all two-column figures (and change
  links in text to point to a) and b) rather than "left" and "right"
  
* Check all references, especially the ones generated by chatgpt !


* case study: tree maps ?

  Perceptual Guidelines for Creating Rectangular Treemaps | IEEE Journals & Magazine | IEEE Xplore https://share.google/9CuP4oxXZKq7K1Ijs 

## Case study: Lines of best fit

@fig-lm-smooth shows a scatter plot of the winning margin
versus the winning score for Tier One nations at Rugby World Cup matches.
The hemisphere of the losing team is encoded as the colour of the data
points and a trend line has been added for each hemisphere.

The trend lines are encodings of data summaries, least-squares
regressions, to line data symbols.  The data summaries consist
of intercept and slope values, which are encoded as the vertical positions
and the angles of the lines, respectively (@fig-stat-vis-decode).

The data symbol is very simple and decoding the data summaries is
also simple because it just involves decoding angle and vertical position.
For example, we can see that, in general, the winning margin is larger
when the winning score is larger, but 
the trend lines suggest that the increase in the winning margin
is slower for southern-hemisphere teams (they fight harder in defeat).

```{r}
#| echo: false
#| label: fig-lm-smooth
#| fig-cap: A scatter plot of the winning margin versus the winning score for matches between Tier One nations at Rugby World Cups.  There is a light shading of the lower-right diagonal to indicate that the winning margin cannot exceed the winning score.  In addition, points on the diagonal represent matches where the losing team failed to score any points.  The straight lines represent a least-squares regression model fit.
#| warning: false
rwcWinner <- do.call(rbind, lapply(split(rwcAll, rwcAll$match), 
                                   function(x) {
                                       if (all(x$scored == x$conceded)) {
                                           winner <- 1
                                           loser <- 2
                                       } else {
                                           winner <- x$scored > x$conceded
                                           loser <- !winner
                                       }
                                       x$oppSphere <- x$hemisphere[loser]
                                       x[winner, ]
                                   }))
ggplot(rwcWinner, aes(scored, scored - conceded, colour=oppSphere)) +
    annotate("polygon", x=c(0, Inf, Inf), y=c(0, Inf, 0), fill="grey97") +
    geom_smooth(method="lm", se=FALSE) +
    geom_point(alpha=.2) +
    scale_x_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning score") +
    scale_y_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning margin") +
    scale_colour_discrete(guide=guide_legend(title="opposition\nhemisphere")) +
    coord_cartesian(clip="off") +
    theme(aspect.ratio=1)
```

@fig-loess-smooth shows the same data as @fig-lm-smooth, but this time
there are curved trend lines added rather than straight trend lines.[^loess]
The curves are encodings of data summaries again, but this time 
loess model fits.  This means that the encoding involves a series
of predicted values encoded as horizontal and vertical positions for
the curve.  In other words, the data summaries are encoded as visual features
that produce a visual shape

[^loess]:
    These are *loess* model fits (locally-estimated scatterplot smoothing).

```{r}
#| echo: false
#| label: fig-loess-smooth
#| fig-cap: Scatter plot
#| warning: false
ggplot(rwcWinner, aes(scored, scored - conceded, colour=oppSphere)) +
    annotate("polygon", x=c(0, Inf, Inf), y=c(0, Inf, 0), fill="white") +
    geom_smooth(method="loess", se=FALSE, span=1) +
    geom_point(alpha=.2) +
    scale_x_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning score") +
    scale_y_continuous(limits=c(0, 100), expand=expansion(c(0, .05)),
                       name="winning margin") +
    scale_colour_discrete(guide=guide_legend(title="opposition\nhemisphere")) +
    coord_cartesian(clip="off") +
    theme(aspect.ratio=1)
```

Nice example of a complex calculation, the results of which are what gets
visualised.  Also example of performing the calculation numerically rather
than asking the viewer to perform it visually.  This was in summaries.qmd,
but it cannot be there because it requires scatter plots, which come in
the next section.
Add this after mention of visualising models?

Add footnote to Hadley's more complex model visualisation paper
https://vita.had.co.nz/papers/model-vis.pdf
https://dl.acm.org/doi/10.1002/sam.11271

model fits like lm straight line and ref Hadley's data-space vs model-space
"data models" to mirror "data statistics"
(revisit box plots as VERY simple models? also talk about failing
assumptions of these models, like unimodal for boxplot)

footnote cite to Tukey for visualising phenomena rather than raw data?
"Data-Based Graphics: Visual Display in the Decades to Come"
Tukey (1990)

+ "Beware of Validation by Eye:
   Visual Validation of Linear Trends in Scatterplots"
  Braun et al (2024)
  People are biased towards orthogonal distance to line rather than
  vertical distance.

loess curves
Another example of data -> summary -> features -> shapes -> summaries


