
# Visual Objects

@fig-flags shows a scatter plot of the number of clean breaks and the
number of tries for countries
at the 2023 Rugby World Cup (@tbl-rwc-2023).
This shows the same positive relationship between clean breaks and tries
that we saw in @fig-scatter.
However, this scatter plot is using a much more complicated data symbol:
the flags of the different countries rather than a simple circle for
each country.

```{r}
#| echo: false
#| label: fig-flags
#| fig-cap: A scatter plot with flags as data symbols
size <- 10
flag <- function(data, coords) {
    width <- unit(size, "mm")
    height <- data$ar * width
    flags <- lapply(1:nrow(data),
                    function(i) {
                        grobTree(rectGrob(coords$x[i], coords$y[i], 
                                          width, height[i], 
                                          gp=gpar(fill="white")),
                                 rasterGrob(readPNG(data$path[i]), 
                                            coords$x[i], coords$y[i], width))
                    })
    do.call(grobTree, flags)
}
ggplot(rwcFlags) +
    grid_panel(flag, aes(breaks, tries, path=path, ar=ar)) +
    theme(aspect.ratio=2/3)
```

The flag data symbols in @fig-flags are examples of **visual objects**.
The flags are not simple shapes like bars or circles.
On the other hand, the relatively complex flags are, at least for
some viewers and for some flags, extremely familiar.

There are simple mappings from number of clean breaks to the
horizontal **position** of each flag and from number of tries to the
vertical **position** of each flag, but each country value is not
being mapped to a basic feature of a simple shape.

This section looks at the effectiveness of using visual objects 
as data symbols.

## Icons

Do dog and cat from master class ?

## Maps

Link to MacEachren and Bertin

## Case study: Chernoff faces

@fig-chernoff visualises the 2023 Rugby World Cup data (@tbl-rwc-2023) using
Chernoff faces.
The data symbol in this visualisation is a little unusual; each
country is represented by a combination of circles, curves, and lines
that together produce a **visual object** that resembles a face.
The mappings in this visualisation are also a little unusual.
The number of points scored is mapped to the **curvature** of the
smile, the number of tries is mapped to the **angle** of the eye brows,
and the number of clean breaks is mapped to the horizontal **position** of 
the eyes.

```{r}
#| echo: false
#| label: fig-chernoff
#| fig-cap: A Chernoff faces plot of the ...
rwcChernoff <- RWCperGame
rwcChernoff$country <- reorder(rwcChernoff$country, rwcChernoff$points)
ggplot(rwcChernoff) +
    geom_chernoff(aes(x="", y="", smile=points, brow=tries, 
                      eyes=breaks),
                  size=20, fill="white") +
    facet_wrap(vars(country), strip.position="bottom") +
    coord_cartesian(clip="off") +
    theme(panel.border=element_blank(),
          strip.background=element_blank(),
          axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank(),
          legend.position="none",
          aspect.ratio=1)
```

The mappings in a Chernoff face are *not* very accurate (see @sec-accuracy;
**curvature** is ranked *below* colour in terms of accuracy), so
Cheroff faces are *not* very effectuve for 
mapping *back* to raw data values.

However, the familiarity of Chernoff faces means that we are easily
able to perceive common combinations of the facial features.
For example, we can see several sad faces at the top-left of @fig-chernoff
and several happy faces at the bottom-right.
We can perceive countries that share similar combinations
of performance measures.
In other words, we can perceive multi-dimensional correlations
and multi-dimensional groupings in the data.

In this case, the happier faces have been arranged to
correspond to the more successful teams and the 
teams have been ordered from least points scored to most points scored.
The resulting proximity helps with perceiving groups of similar faces.

> A Chernoff face is effective at conveying multi-dimensional relationships
  because the data symbol is a familiar visual object from which we can
  easily perceive simple human emotions.

