
# Visual Objects

@fig-scatter-country shows a scatter plot of the number of clean breaks and the
number of tries for countries
at the 2023 Rugby World Cup (@tbl-rwc-2023).
This data visualisation is similar to @fig-scatter, but it adds an
extra mapping of country names to the **shape** of the data symbols.
This is a reasonable mapping, despite the large number of different
countries, because each different shape only appears once.
We are able to distinguish all countries from each other and,
thanks to the legend, we can identify individual countries.

```{r}
#| echo: false
#| label: fig-scatter-country
#| fig-cap: A scatter plot with a different shape per country
ggplot(RWCperGame) +
    geom_point(aes(breaks, tries, shape=country), fill="grey", size=3) +
    scale_shape_manual(values=(1:25)[-c(15:18, 20)], 
                       guide=guide_legend(ncol=2), name=NULL) +
    theme(aspect.ratio=2/3)
```

@fig-flags shows another scatter plot of the number of clean breaks and the
number of tries for countries
at the 2023 Rugby World Cup, but this time with flags as the data symbols.
In one sense, this is just a variation on @fig-flags because the country
names are mapped to the **shape** of a data symbol. 
Each data symbol is a shape that allows us to distinguish between the
different countries.

The difference between the data symbols in @fig-scatter-country and 
the data symbols on @fig-flags is that the flags are not only more
complicated shapes, but they are familiar shapes.
The flags convey differences, but they
also convey the more sophisticated concept of the country
names.
In the terms of @sec-visual-processing, 
the flags are **visual objects** that might take longer to
process, but convey more information.

```{r}
#| echo: false
#| label: fig-flags
#| fig-cap: A scatter plot with flags as data symbols
size <- 10
flagGroup <- function(data, coords) {
    width <- unit(size, "mm")
    height <- data$ar * width
    grobTree(rectGrob(coords$x, coords$y, 
                      width, height, 
                      gp=gpar(fill="white")),
             rasterGrob(readPNG(data$path), 
                        coords$x, coords$y, width))
}
flag_key <- function(data, ...) {
    i <- data$shape
    data <- rwcFlags[i, ]
    width <- unit(size, "mm")
    height <- data$ar * width
    grobTree(rectGrob(.5, .5,
                      width, height, 
                      gp=gpar(fill="white")),
             rasterGrob(readPNG(data$path), 
                        .5, .5, width))
}
ggplot(rwcFlags) +
    grid_group(flagGroup, 
               aes(breaks, tries, path=path, ar=ar, 
                   shape=country),
               inherit.aes=FALSE,
               show.legend=TRUE, key_glyph=flag_key) +
    scale_shape_manual(values=1:nrow(rwcFlags), guide=guide_legend(ncol=2)) +
    theme(legend.key.width=unit(size, "mm"),
          legend.key.height=unit(max(rwcFlags$ar)*size + 1, "mm"),
          aspect.ratio=2/3)
```

@fig-flags-direct demonstrates this point by repeating the scatter plot,
but removing the legend.  We are still able to identify not only 
that the data symbols represent different things, but we can identify
the country names as well.

```{r}
#| echo: false
#| label: fig-flags-direct
#| fig-cap: A scatter plot with flags as data symbols
flagPanel <- function(data, coords) {
    width <- unit(size, "mm")
    height <- data$ar * width
    flags <- lapply(1:nrow(data),
                    function(i) {
                        grobTree(rectGrob(coords$x[i], coords$y[i], 
                                          width, height[i], 
                                          gp=gpar(fill="white")),
                                 rasterGrob(readPNG(data$path[i]), 
                                            coords$x[i], coords$y[i], width))
                    })
    do.call(grobTree, flags)
}
ggplot(rwcFlags) +
    grid_panel(flagPanel, 
               aes(breaks, tries, path=path, ar=ar)) +
    theme(aspect.ratio=2/3)
```

Just to make the point completely clear, @fig-scatter-country-no-legend
shows @fig-scatter-country without a legend.
There is nothing about the shape of the data symbols in 
@fig-scatter-country-no-legend that conveys the country names.
We can clearly see that @fig-flags-direct contains more information
than @fig-scatter-country-no-legend.

```{r}
#| echo: false
#| label: fig-scatter-country-no-legend
#| fig-cap: A scatter plot with a different shape per country
ggplot(RWCperGame) +
    geom_point(aes(breaks, tries, shape=country), fill="grey", size=3) +
    scale_shape_manual(values=(1:25)[-c(15:18, 20)], 
                       guide="none") +
    theme(aspect.ratio=2/3)
```

> A plot that uses visual objects as data symbols can be effective
  because the visual objects can convey more information
  and more complex information.

@fig-flags-direct also demonstrates a weakness of **visual objects**.
The flags work well for identifying the country names *only if*
we are familiar with all of the different flags.
For example, it may only be Australians and New Zealanders who can 
reliably identify the difference between their two flags.

> A plot that uses visual objects as data symbols may *not* be effective
  if the visual objects are only familiar to some people.

## Icons

Do dog and cat from master class ?

## Case study: Choropleth maps {#sec-choropleth}

@fig-map shows a choropleth map of the
average crime rate, over the period from 2011 to 2021, 
in each police district of New Zealand.
The data values for this map are shown in @tbl-map-data.

```{r}
#| echo: false
#| label: tbl-map-data
#| tbl-cap: The average crime rate over the period 2011 to 2021 for each police district of New Zealand.
kable(crimeDistrictTotal[c("district", "rate")], digits=2)
```

The data symbol in @fig-map is a polygon representing the geographic
boundary of each police district.  This implicitly entails a mapping
from police district to the **position** of each polygon.
Crime rate is mapped to the fill colour of each region.

The use of polygons as data symbols is effective because
the polygons form familiar shapes with familiar spatial arrangements
(familiar to a New Zealand audience at least).
The polygons are **visual objects** that convey the concepts
of districts within New Zealand.
The spatial arrangements are also a form of visual **congruence**;
the arrangements mirror the real-world spatial relationships 
between the regions
(see @sec-congruence).

```{r}
#| echo: false
#| label: fig-map
#| fig-cap: A choropleth map of the average crime rate over the period 2011 to 2021 for each police district.
districts <- 
    st_read("Data/YouthCrime/SHP/nz-police-district-boundaries-29-april-2021.shp",
            quiet=TRUE)
centroids <- st_coordinates(st_centroid(st_geometry(districts)))
districts$X <- centroids[,1]
districts$Y <- centroids[,2]
districts <- inner_join(districts, crimeDistrictTotal,
                        by=join_by(D_MACRON == district))
ggplot(districts) +
    geom_sf(aes(fill=rate)) +
    scale_fill_continuous(name="crime rate") +
    theme(axis.ticks=element_blank(),
          axis.text=element_blank())
```

> A choropleth map is effective because it maps geographic regions to
  familiar visual objects that convey 

A weakness of @fig-map is that it maps the **quantitative** crime rate
to colour (chroma and/or luminance).  
This means that we can only map *back* to ordinal comparisons
between regions (see @sec-ordinal).

> A choropleth map is *only* effective for making ordinal comparisons
  between geographic regions because it maps quantitative values to
  colour.

@fig-district-bar shows a bar plot of the same data,
which demonstrates the advantages and disadvantages of
the choropleth map in @fig-map.
In @fig-district-bar, it is much easier to compare the crime rate
values because crime rates are mapped to bar **length**, but
it is much harder to associate the different regions with their
spatial arrangements because all we have are the region names.

```{r}
#| echo: false
#| label: fig-district-bar
crimeDistrictBar <- crimeDistrictTotal
crimeDistrictBar$district <- 
    reorder(crimeDistrictBar$district, 
            crimeDistrictBar$total/crimeDistrictBar$avgPop)
ggplot(crimeDistrictBar) +
    geom_col(aes(total/avgPop, district)) +
    scale_x_continuous(name="crime rate", expand=expansion(c(0, .05))) +
    scale_y_discrete(name=NULL) + 
    theme(aspect.ratio=1)
```

Another potential weakness with choropleth maps is the fact that
regions differ in terms of **area**.
We know from @sec-more-is-more that larger data symbols are
associated with larger data values.
The larger polygons accurately
reflect the fact that some 
police districts are geographically larger than other, 
but there is potential for larger polygons
to be erroneously mapped *back* to larger crime rates.

Add links to alternatives like cartograms (though I really don't like
those because you lose the familiar shapes so what's the point)
and dots in regions and ...

Link to MacEachren and Bertin

```{r}
#| echo: false
#| label: fig-map-dot
#| fig-cap: A circle map of the average crime rate over the period 2011 to 2021 for each police district.
ggplot(districts) +
    geom_sf() +
    geom_point(aes(X, Y, size=rate)) +
    scale_fill_continuous(name="crime rate") +
    theme(axis.ticks=element_blank(),
          axis.text=element_blank())
```

## Case study: Chernoff faces

@fig-chernoff visualises the 2023 Rugby World Cup data (@tbl-rwc-2023) using
Chernoff faces.
The data symbol in this visualisation is a little unusual; each
country is represented by a combination of circles, curves, and lines
that together produce a **visual object** that resembles a face.
The mappings in this visualisation are also a little unusual.
The number of points scored is mapped to the **curvature** of the
smile, the number of tries is mapped to the **angle** of the eye brows,
and the number of clean breaks is mapped to the horizontal **position** of 
the eyes.

```{r}
#| echo: false
#| label: fig-chernoff
#| fig-cap: A Chernoff faces plot of the ...
rwcChernoff <- RWCperGame
rwcChernoff$country <- reorder(rwcChernoff$country, rwcChernoff$points)
ggplot(rwcChernoff) +
    geom_chernoff(aes(x="", y="", smile=points, brow=tries, 
                      eyes=breaks),
                  size=20, fill="white") +
    facet_wrap(vars(country), strip.position="bottom") +
    coord_cartesian(clip="off") +
    theme(panel.border=element_blank(),
          strip.background=element_blank(),
          axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank(),
          legend.position="none",
          aspect.ratio=1)
```

The mappings in a Chernoff face are *not* very accurate (see @sec-accuracy;
**curvature** is ranked *below* colour in terms of accuracy), so
Cheroff faces are *not* very effectuve for 
mapping *back* to raw data values.

However, the familiarity of Chernoff faces means that we are easily
able to perceive common combinations of the facial features.
For example, we can see several sad faces at the top-left of @fig-chernoff
and several happy faces at the bottom-right.
We can perceive countries that share similar combinations
of performance measures.
In other words, we can perceive multi-dimensional correlations
and multi-dimensional groupings in the data.

In this case, the happier faces have been arranged to
correspond to the more successful teams and the 
teams have been ordered from least points scored to most points scored.
The resulting proximity helps with perceiving groups of similar faces.

> A Chernoff face is effective at conveying multi-dimensional relationships
  because the data symbol is a familiar visual object from which we can
  easily perceive simple human emotions.

