
# Visual Objects

@fig-scatter-country shows a scatter plot of the number of clean breaks and the
number of tries for countries
at the 2023 Rugby World Cup (@tbl-rwc-2023).
This data visualisation is similar to @fig-scatter, but it adds an
extra mapping of country names to the **shape** of the data symbols.
This is a reasonable mapping, despite the large number of different
countries, because each different shape only appears once.
We are able to distinguish all countries from each other and,
thanks to the legend, we can identify individual countries.

```{r}
#| echo: false
#| label: fig-scatter-country
#| fig-cap: A scatter plot with a different shape per country
ggplot(RWCperGame) +
    geom_point(aes(breaks, tries, shape=country), fill="grey", size=3) +
    scale_shape_manual(values=(1:25)[-c(15:18, 20)], 
                       guide=guide_legend(ncol=2), name=NULL) +
    theme(aspect.ratio=2/3)
```


@fig-flags shows a data visualisation of the number of clean breaks and the
number of tries for countries
at the 2023 Rugby World Cup (@tbl-rwc-2023).

@fig-flags is similar to the scatter plot of the same data in 
@fig-scatter.
In both cases, there are simple mappings from number of clean breaks to the
horizontal **position** of a data symbol and from number of tries to the
vertical **position** of a data symbol.
However, the data symbol in @fig-scatter was a simple dot and the
data symbol in @fig-flags is a flag image.
Furthermore, @fig-flags contains
an additional mapping:  each country is mapped to a different flag image.

```{r}
#| echo: false
#| label: fig-flags
#| fig-cap: A scatter plot with flags as data symbols
size <- 10
flagGroup <- function(data, coords) {
    width <- unit(size, "mm")
    height <- data$ar * width
    grobTree(rectGrob(coords$x, coords$y, 
                      width, height, 
                      gp=gpar(fill="white")),
             rasterGrob(readPNG(data$path), 
                        coords$x, coords$y, width))
}
flag_key <- function(data, ...) {
    i <- data$shape
    data <- rwcFlags[i, ]
    width <- unit(size, "mm")
    height <- data$ar * width
    grobTree(rectGrob(.5, .5,
                      width, height, 
                      gp=gpar(fill="white")),
             rasterGrob(readPNG(data$path), 
                        .5, .5, width))
}
ggplot(rwcFlags) +
    grid_group(flagGroup, 
               aes(breaks, tries, path=path, ar=ar, 
                   shape=country),
               inherit.aes=FALSE,
               show.legend=TRUE, key_glyph=flag_key) +
    scale_shape_manual(values=1:nrow(rwcFlags), guide=guide_legend(ncol=2)) +
    theme(legend.key.width=unit(size, "mm"),
          legend.key.height=unit(max(rwcFlags$ar)*size + 1, "mm"),
          aspect.ratio=2/3)
```

```{r}
#| echo: false
#| label: fig-flags-direct
#| fig-cap: A scatter plot with flags as data symbols
flagPanel <- function(data, coords) {
    width <- unit(size, "mm")
    height <- data$ar * width
    flags <- lapply(1:nrow(data),
                    function(i) {
                        grobTree(rectGrob(coords$x[i], coords$y[i], 
                                          width, height[i], 
                                          gp=gpar(fill="white")),
                                 rasterGrob(readPNG(data$path[i]), 
                                            coords$x[i], coords$y[i], width))
                    })
    do.call(grobTree, flags)
}
ggplot(rwcFlags) +
    grid_panel(flagPanel, 
               aes(breaks, tries, path=path, ar=ar)) +
    theme(aspect.ratio=2/3)
```

POINT OUT that the flags demonstrate visual objects - familiarity
BUT ALSO that they demonstrate the pitfalls, like if you do not 
recognise some of the flags!

The flag data symbols in @fig-flags are examples of **visual objects**.
The country data values are not being mapped to a basic visual feature
of a data symbol.  Instead, each country is mapped to a flag data symbol that
is very different in many details from the other flag data symbols.

We could say that we are mapping the country to the **shape** of the flag,
which is a reasonable mapping because we can perceive differences
between the **shape** of the different flags, which allows us to
map *back* to the different countries.
However, the flags are more than just noticeably different from each other.
At least for
some viewers, each flag is extremely familiar, so the data symbol
conveys the actual country data value, not just that each country is different.

the country names more effectively than, for example,
mapping country names to the **colour** of data points.

This section looks at the effectiveness of using visual objects 
as data symbols.

## Icons

Do dog and cat from master class ?

## Case study: Choropleth maps {#sec-choropleth}

```{r}

```

The major advantage is the mapping *back* from map regions to 
the corresponding conceptual regions.

One weakness is that we use colour for quantitative values.

One weakness is that areas change without corresponding changes in the data.

Link to MacEachren and Bertin

## Case study: Chernoff faces

@fig-chernoff visualises the 2023 Rugby World Cup data (@tbl-rwc-2023) using
Chernoff faces.
The data symbol in this visualisation is a little unusual; each
country is represented by a combination of circles, curves, and lines
that together produce a **visual object** that resembles a face.
The mappings in this visualisation are also a little unusual.
The number of points scored is mapped to the **curvature** of the
smile, the number of tries is mapped to the **angle** of the eye brows,
and the number of clean breaks is mapped to the horizontal **position** of 
the eyes.

```{r}
#| echo: false
#| label: fig-chernoff
#| fig-cap: A Chernoff faces plot of the ...
rwcChernoff <- RWCperGame
rwcChernoff$country <- reorder(rwcChernoff$country, rwcChernoff$points)
ggplot(rwcChernoff) +
    geom_chernoff(aes(x="", y="", smile=points, brow=tries, 
                      eyes=breaks),
                  size=20, fill="white") +
    facet_wrap(vars(country), strip.position="bottom") +
    coord_cartesian(clip="off") +
    theme(panel.border=element_blank(),
          strip.background=element_blank(),
          axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank(),
          legend.position="none",
          aspect.ratio=1)
```

The mappings in a Chernoff face are *not* very accurate (see @sec-accuracy;
**curvature** is ranked *below* colour in terms of accuracy), so
Cheroff faces are *not* very effectuve for 
mapping *back* to raw data values.

However, the familiarity of Chernoff faces means that we are easily
able to perceive common combinations of the facial features.
For example, we can see several sad faces at the top-left of @fig-chernoff
and several happy faces at the bottom-right.
We can perceive countries that share similar combinations
of performance measures.
In other words, we can perceive multi-dimensional correlations
and multi-dimensional groupings in the data.

In this case, the happier faces have been arranged to
correspond to the more successful teams and the 
teams have been ordered from least points scored to most points scored.
The resulting proximity helps with perceiving groups of similar faces.

> A Chernoff face is effective at conveying multi-dimensional relationships
  because the data symbol is a familiar visual object from which we can
  easily perceive simple human emotions.

